
---
title: "`r gsub('.Rmd', ' ', gsub('_', ' ', knitr::current_input()))`"
author: "`r Sys.getenv('USER')`"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: yes
---
  



mount drive locally as follows (not run on knit)

```{r eval=FALSE, include=TRUE}

server=hbeale@mustard:/; mount_name=mustard
umount $server
mnt_point=/Users/hbeale/servers/${mount_name}
mkdir -p $mnt_point

```

```{r setup}

library(tidyverse)
library(jsonlite)
library(janitor)

```


```{r}

data_base_dir <- "data"
# mustard_base <- "/Users/hbeale/servers/mustard/private/"
# mustard_prefix <-  "/Users/hbeale/servers/mustard/"
#mustard_base <- "/private/"
#data_dir <- file.path(mustard_base, "groups/treehouse/working-projects/hbeale/ckcc2_2023_07")
data_dir <- "."
paths_file <- "original_tertiary_locations_as_of_2023_10_25-10_27_50_2023.tsv"
samples_file <- "ckcc2_samples_and_metadata_from_redcap_2023.10.24_13.26.34.tsv"

paths_raw <- read_tsv(paths_file,
                      col_types = cols(.default = "c")) %>%
  mutate(tertiary_path = file.path(
    data_base_dir,
    Sample,
    "tertiary",
    recorded_protocol_version,
    compendium_version))

nrow(paths_raw)

project_samples <- read_tsv(samples_file,
                      col_types = cols(.default = "c"))
nrow(project_samples)

# reminder; there are more lines in the paths_raw tibble than
# in the project_samples tibble because some samples have
# multiple tert outputs


# check whether all samples are present in both data sources
# if both groups are identical,  result will be "character(0)"
setdiff(paths_raw$Sample, project_samples$THid)
setdiff(project_samples$THid, paths_raw$Sample)

tertiary_output_locations <- paths_raw %>%
  select(sample_id = Sample, tertiary_path) 

```

# Compendium and protocol version survey
```{r}


paths_raw %>%
  select(Sample, compendium_version,
         recorded_protocol_version)
#install.packages("janitor")

tabyl(paths_raw , compendium_version)
tabyl(paths_raw , recorded_protocol_version)
```

# Load druggable genes
```{r}

# druggable_genes <- read_tsv(file.path(data_dir, "treehouseDruggableGenes_2020-03_25.txt"))

```

# Load "automatedLeadsIdentified" outlier results 
From this, we'll determine whether the gene was considered druggable at the time
```{r}
automatedLeadsIdentified <- tertiary_output_locations %>%
 group_by(sample_id) %>% 
  mutate(automatedLeadsIdentified_raw = map(file.path(tertiary_path, "automatedLeadsIdentified.tsv"), 
                         read_tsv, col_types = cols(.default = "c"))) %>%
unnest(automatedLeadsIdentified_raw)
```

# Make list of genes that were identified as druggable at the time
```{r}
druggable_genes <- automatedLeadsIdentified %>%
  filter(assay == "druggableUpOutlier",
         results != "No druggableUpOutliers") %>%
  select(sample_id, gene = results, details) %>%
  mutate(pc_up = str_detect(details, "pc"),
         pd_up = str_detect(details, "pd"),
         druggable = TRUE) %>%
#  filter(sample_id == "TH34_1381_S01",
 #        gene == "VEGFA") %>%
  select(-details) %>%
  replace_na(list(pc_up = FALSE, pd_up = FALSE)) 
  
```


# pathway support
```{r}
warty_automatedLeadsIdentified <- automatedLeadsIdentified %>%
  ungroup %>%
  select(TH_id = THid, assay, results, details) %>%
  filter(results != "No druggableUpOutliers",
         results != "no enriched druggable geneSets",
         str_detect(assay, "geneSets")) %>%
  select(TH_id, results, details) %>%
  mutate(gene_and_outlier_type = str_remove(details, "^druggable: ") %>%
           str_split(pattern = ", ")) %>%
  unnest(gene_and_outlier_type) %>%
  mutate(gene = str_remove(gene_and_outlier_type, "\\(.*$") %>%
           str_trim(),
           outlier_type = str_remove(gene_and_outlier_type, "^.*\\(") %>%
           str_remove("\\)"),
         outlier_type = str_split(outlier_type, pattern = " and ")) %>%
  unnest(outlier_type)
         
# sources of pathways        
pathway_sources <- warty_automatedLeadsIdentified %>%
  select(TH_id, gene, outlier_type, results) %>%
  #mutate(results = str_sub(results, start = 1, end = 10)) %>%
  mutate(source = str_remove(results, "_.*$")) 

pathway_sources %>%
  tabyl(source)

pathway_sources %>%
  group_by(source) %>%
  slice_head(n=1)

outliers_with_pathway_support <- warty_automatedLeadsIdentified %>%
  select(TH_id, gene, results) %>%
  distinct() %>%
  group_by(TH_id, gene) %>%
  summarize(n_pathways = n()) %>%
  mutate(has_pathway_support = TRUE)
   
```

# Save pathway details
```{r}
output_name <- paste0("contemporarily_druggable_up_outliers_with_pathway_support_gathered_", format(Sys.time(), "%Y_%m_%d-%H_%M_%S_%Y"), "_based_on_", paths_file)

output_name


write_tsv(warty_automatedLeadsIdentified %>%
  select(TH_id, gene, outlier_type, results),
  output_name)         


```

# Load pan cancer outlier results
details on file structure: https://github.com/UCSC-Treehouse/reference-file-info/blob/master/Explainer%20for%20outlier_results_files.txt
```{r}

outlier_results_file_data <- tertiary_output_locations %>%
 group_by(sample_id) %>% 
  mutate(leads_raw = map(file.path(tertiary_path, paste0("outlier_results_", sample_id)), 
                         read_tsv, col_types = cols(.default = "c"))) %>%
unnest(leads_raw)

```

# Select pan-cancer up outliers 
```{r}

pc_outliers <- outlier_results_file_data %>%
  filter(pc_outlier=="pc_up") %>%
  select(sample_id, gene = Gene) %>%
  mutate(pc_up = TRUE) 

```

# Import pan-disease up outliers
### notes on using personalized_consensus_counts up_outliers
j4$personalized_consensus_counts$up_outliers lists all genes that are up outliers relative to more than 1 pan disease cohort. I  confirmed with code that everything in the personalized_consensus_counts up_outliers in top 5

takes about 2 seconds per samples
```{r}
# code for samples with two cohorts, currently TH34_1349_S01, TH34_1349_S02


get_pan_disease_outliers <- function(this_tert_path){
  j4 <- fromJSON(file.path(this_tert_path, "4.0.json"))
  tibble(gene = names(j4$personalized_consensus_counts$up_outliers))
}

pan_disease_outliers_raw <- tertiary_output_locations %>%
  mutate(pan_disease_outliers_raw = map(tertiary_path, get_pan_disease_outliers)) %>%
  unnest(pan_disease_outliers_raw)

pan_disease_outliers <- pan_disease_outliers_raw %>%
  select(sample_id, gene) %>%
  mutate(pd_up = TRUE)

```

# Combine
```{r}
all_outliers <- pc_outliers  %>%
  full_join(pan_disease_outliers, by = c("sample_id", "gene")) %>%
  left_join(druggable_genes %>%
              select(-pd_up),
            by = c("sample_id", "gene", "pc_up")) %>%
  left_join(outliers_with_pathway_support %>%
              select(-n_pathways),
            by=c("sample_id"="TH_id", "gene")) %>%
  arrange(has_pathway_support)  %>%
  replace_na(list(druggable = FALSE, 
                  pd_up = FALSE,
                  pc_up = FALSE,
                  has_pathway_support = FALSE)) 
```


```{r}


all_outliers %>%
  filter(pc_up, !pd_up) %>%
  head()


all_outliers %>%
  filter(pd_up, !pc_up) %>%
  head()


all_outliers %>%
  filter(pc_up, pd_up) %>%
  head()

all_outliers %>%
  filter(druggable) %>%
  head()


all_outliers %>%
  filter(has_pathway_support) %>%
  head()

```


```{r}

output_name <- paste0("outliers_gathered_", format(Sys.time(), "%Y_%m_%d-%H_%M_%S_%Y"), "_based_on_", paths_file)

output_name

write_tsv(all_outliers, output_name)

```

