---
title: "`r gsub('.Rmd', ' ', gsub('_', ' ', knitr::current_input()))`"
author: "`r Sys.getenv('USER')`"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
  toc: yes
---
  
  
updated the report 2023.10.23_09.35.56 to include lead/target/rationale #4
updated 2023.10.23_16.46.48 to include date of presentation

```{r setup}

library(tidyverse)
library(redcapAPI)
library(janitor)
library(haven)
library(kableExtra)

```
# Load samples
```{r}

samples <- read_tsv("ckcc2_samples_and_metadata_from_redcap_2023.10.23_14.44.26.tsv") %>%
  mutate(donor_key = str_remove(donorID, "^TH.*_"))
  
```



### Retrieve data from redcap

```{r}

source("~/Documents/Dropbox/ucsc/projects/redcapAPI/rcConfig.R")

hb_findings_report_ID=538


rcon <- redcapConnection(url= uri, token= token)

findings_data_raw <- as_tibble (exportReports(rcon, hb_findings_report_ID)) 

findings_data <- findings_data_raw %>% 
  mutate_all(as.character) %>%
  type_convert() %>%
  filter(donor_key %in% samples$donor_key)
```

# Review donors with multiple findings records
```{r}

tabyl(findings_data,
      current)


multiple_entries_to_review <- findings_data %>%
  filter(duplicated(donor_key)) %>%
  pull(donor_key)

dupe_findings <- findings_data %>%
  arrange(donor_key, redcap_repeat_instance) %>%
  filter(donor_key %in% multiple_entries_to_review) 

dupe_findings %>%
  kableExtra::kbl() %>%
  kableExtra::kable_styling() %>% 
  group_rows(index = setNames(rle(as.vector(dupe_findings$donor_key))[[1]],
                              rle(as.vector(dupe_findings$donor_key))[[2]]))


```


## observations
1149 has two findings for one sample, and the target, rationale and lead are the same; select redcap_repeat_instance == 2

1162 has two findings for one sample, and the target, rationale and lead are the same; select redcap_repeat_instance == 2

1349 has two findings for two samples, and they differ; use both

1351 has two findings for two sets of samples. The target, rationale and lead are the same; select redcap_repeat_instance == 2

1380 has two findings for two samples, and they differ; use both (one has no results)

1445 has two findings for two samples, and they differ; use both

1447 has two findings for two samples, and they differ; use both

2351 has two findings for one sample. The target, rationale and lead for #2 has additional mutation info; select redcap_repeat_instance == 2

2410 has two hits; redcap_repeat_instance == 2 is marked current, while #1 is not; use #2

## record observations

```{r}

which_of_multiple_findings_to_keep <- tribble(
  ~donor_key, ~redcap_repeat_instance,
  1149, 2,
  1162, 2,
  1349, 1,
  1349, 2,
  1351, 2,
  1380, 1,
  1380, 2,
  1445, 1,
  1445, 2,
  1447, 1,
  1447, 2,
  2351, 2,
  2410, 2
)

findings_to_keep <- bind_rows(
  findings_data %>%
    filter(! donor_key %in% multiple_entries_to_review) %>%
    select(donor_key, redcap_repeat_instance), #%>%
    #type_convert(),
  which_of_multiple_findings_to_keep)


```



```{r}

findings_data_long_raw <- left_join(findings_to_keep,
                                    findings_data,
                                    by = c("donor_key", "redcap_repeat_instance")) %>%
  select(donor_key, samples_used, redcap_repeat_instance, tb_date, starts_with(c("target_", "rationale_", "lead_"))) %>%
  pivot_longer(cols = starts_with(c("target_", "rationale_", "lead_"))) %>%
  mutate(
    name_raw = name,
    name = str_remove(name, "_.*$"),
    iteration = str_remove(name_raw, "^.*_"))


findings_data_long <- findings_data_long_raw %>%
  select(-name_raw) %>%
  pivot_wider() %>%
  unnest_longer(starts_with(c("target", "rationale", "lead"))) %>%
  filter(! is.na(target), 
         ! is.na(rationale), 
         ! is.na(lead)) %>%
  arrange(donor_key,
          redcap_repeat_instance)

```


```{r}

underscore2space <- function(x) str_replace_all(x, "_", " ")

findings_data_long %>%
  rename_with(underscore2space) %>%
  kableExtra::kbl() %>%
  kableExtra::kable_styling()

```


```{r}

output_file_name <- 
  paste0("findings_", format(Sys.time(), "%Y%m%d_%H%M%S"), ".tsv")

output_file_name

write_tsv(findings_data_long, output_file_name)

```
