---
title: "CKCC2_curation_vs_automated_findings"
author: "Yvonne Vasquez"
date: "`r Sys.Date()`"
output: html_document
---


### Load libraries
```{r}

library(tidyverse)
library("readxl")

```

### Load data
# This file describes for all findings presented to Stanford, if they were found via our CARE pipeline using canonical cohorts or curated cohorts. It also describes if a finding was a pan-cancer outlier, pan-disease ouliter, expressed mutation, highly expressed gene (95th percentile). For pd outliers it further describes if it was a consensus outlier found in at least two cohorts or if it was a single-cohort outlier.

```{r}

ckcc2_findings_table <- read_tsv("CKCC2_evidence_for_findings_2023.11.15_13.16.11.tsv")
nrow(ckcc2_findings_table)

```
### Quantify how many outliers fell into the following categories
# 1) pan-cancer outlier only (pc)
# 2) pan-disease ouliter only (pd)
# 3) Bothpan-cancer and pan-disease outlier (pc,pd)
# 4) expressed mutation (mutation)
# 5) expressed fusion (fusion)
# 6) highly expressed gene (95th percentile)

```{r}

ckcc2_finding_type <- plyr::count(ckcc2_findings_table$summary_of_finding)
ckcc2_finding_type

```
### Quantify how many findings were identified by our canonical CARE pan-cancer pipeline

```{r}

pc_findings <- ckcc2_findings_table %>%
  filter(pc_outlier == TRUE)
nrow(pc_findings)

```


### Quantify how many pan-disease findings were identified and how many used curated vs canonical cohorts

```{r}

# Filter for pan-disease findings and quantify how many there were
pd_findings <- ckcc2_findings_table %>%
  filter(pd_outlier == TRUE)
nrow(pd_findings)

# Quantify how many findings were identified by our canonical CARE pan-disease pipeline (canonical consensus pd outliers)
pd_canonical_consensus <- pd_findings %>%
  filter(`canonical vs curated pd` == "canonical", `pd type` == "consensus")
nrow(pd_canonical_consensus)

# Quantify how many findings were identified by our canonical pd pipeline once we added a curated pd cohort (curated consensus)
pd_curated_consensus <- pd_findings %>%
  filter(`canonical vs curated pd` == "curated", `pd type` == "consensus")
nrow(pd_curated_consensus)


pd_findings_quantified <- plyr::count(pd_findings$`canonical vs curated pd`)
pd_findings_quantified

pd_findings_cohort_type <- plyr::count(pd_findings$`pd type`)
pd_findings_cohort_type

```

### Quantify how many findings were identified by curation
```{r}

curated_findings <- ckcc2_findings_table %>%
  filter(`canonical vs curated pd`== "curated")
nrow(curated_findings)

curated_findings_type <- plyr::count(curated_findings$summary_of_finding)
curated_findings_type
```

