---
title: "`r gsub('.Rmd', ' ', gsub('_', ' ', knitr::current_input()))`"
author: "`r Sys.getenv('USER')`"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: yes
---
  


```{r setup, include = FALSE}

library(tidyverse)
library(janitor)
library(khroma)
library(kableExtra)
```


```{r}
underscore_to_space <- function(x) str_replace_all(x, "_", " ")

```

```{r}

outliers <- read_tsv("../input_data/druggable_outliers_from_treehouse_and_other_cohorts_2023_11_09-13_46_32_2023.tsv") %>%
  mutate(high_level_cohort = ifelse(str_detect(comparison_cohort, "Treehouse"),
                                    "Treehouse",
                                    comparison_cohort))

         
         
```

# Define cohort codes
```{r}

cohort_codes <- tibble(
  cohort_name = 
    c("PEDAYA", "TCGA", "TH03_TH34", "Treehouse_pc", "Treehouse_pd"),
  cohort_code = 
    c("P", "T", "S", "C", "D"))

```

# Tile plot of all outliers
```{r fig.height=10, fig.width=10}

ggplot(outliers) +
  geom_tile(aes(x=comparison_cohort,
                y=gene, 
                fill = comparison_cohort)) +
  facet_wrap(~Sample_ID,
             nrow = 1) +
  theme(#axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    axis.text.x = element_blank(),
    strip.text.x = element_text(angle = 90),
        ) +
  xlab("")  +
  scale_fill_bright()

  
```

# Heatmap shows number of cohorts in which outlier were  detected
I can make this look better if we decide to use it, but it's non-trivial
```{r fig.height=10, fig.width=6}
outliers_heatmap_data <- outliers %>%
  group_by(Sample_ID, gene) %>%
  summarize(n_outliers = n()) 

ggplot(outliers_heatmap_data) +
  geom_tile(aes(x=Sample_ID,
                y=gene,
                fill = n_outliers), 
            color = "black")  +
  #theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) 


```
```{r}

library(ggVennDiagram)
raw_outliers_for_venn <- outliers %>%
  mutate(sample_gene = paste(Sample_ID, gene, sep = "_")) %>%
  arrange(comparison_cohort) %>%
  select(sample_gene, comparison_cohort) %>%
  group_split(comparison_cohort)


list_of_outliers_for_venn <-  lapply(raw_outliers_for_venn, function(x) x %>% pull(sample_gene))
names(list_of_outliers_for_venn) <- unique(outliers$comparison_cohort) %>% sort

ggVennDiagram(list_of_outliers_for_venn,
              show_intersect = TRUE)

 
ggVennDiagram(list_of_outliers_for_venn) + 
  scale_fill_distiller(palette = "Reds", direction = 1)


```





# Annotate with combined full low level cohort names
```{r}
collapse_fun <- function(x){ paste(x,collapse = ", ") }

all_outliers_combined_wide <- outliers %>%
  select(-pathway_support, -donor_ID) %>%
  pivot_wider(names_from = Sample_ID,
              values_from = comparison_cohort,
              values_fn = collapse_fun)

n_distinct(outliers$Sample_ID)
n_distinct(outliers$gene)
all_outliers_combined_long <- all_outliers_combined_wide %>%
  pivot_longer(-gene,
               names_to = "Sample_ID",
               values_to = "comparison_cohorts") %>%
  na.omit()
```


# How many outliers are present in each combination of cohorts?
```{r}
tabyl(all_outliers_combined_long,
      comparison_cohorts) %>%
  arrange(desc(n)) %>%
  adorn_pct_formatting() %>%
  adorn_totals() %>%
  kbl() %>%
  kable_styling(full_width = F)
```


# Tile plot of combination of outliers
```{r fig.height=10, fig.width=10}

ggplot(all_outliers_combined_long) +
  geom_tile(aes(x=Sample_ID,
                y=gene,
                fill = comparison_cohorts))  +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
                
n_distinct(all_outliers_combined_long$Sample_ID)                

```




# Annotate with combined full high level cohort names
```{r}

high_level_all_outliers_combined_wide <- outliers %>%
  select(-pathway_support, -donor_ID, -comparison_cohort) %>%
  distinct() %>%
  pivot_wider(names_from = Sample_ID,
              values_from = high_level_cohort,
              values_fn = collapse_fun)

n_distinct(outliers$Sample_ID)
n_distinct(outliers$gene)
high_level_all_outliers_combined_long <- high_level_all_outliers_combined_wide %>%
  pivot_longer(-gene,
               names_to = "Sample_ID",
               values_to = "comparison_cohorts") %>%
  na.omit()
```


# How many outliers are present in each high level combination of cohorts?
```{r}
tabyl(high_level_all_outliers_combined_long,
      comparison_cohorts) %>%
  arrange(desc(n)) %>%
  adorn_pct_formatting() %>%
  adorn_totals() %>%
  kbl() %>%
  kable_styling(full_width = F)
```


# Annotate with minimal combined cohort abbreviations
```{r}
collapse_fun_no_coma <- function(x){ paste(x,collapse = "") }

# backslashes prevent asterisks from being interpreted as italics in the kbl table

all_outliers_min_abbrev_combined_wide <- outliers %>%
  left_join(cohort_codes,
            by=c("comparison_cohort"="cohort_name")) %>%
  mutate(cohort_code_pathway = ifelse(pathway_support,
                                      paste0(cohort_code, "\\*"),
                                      cohort_code)) %>%
  select(-pathway_support, -donor_ID,
         -comparison_cohort,
         -cohort_code) %>%
  pivot_wider(names_from = Sample_ID,
              values_from = cohort_code_pathway,
              values_fn = collapse_fun_no_coma,
              values_fill = "")


all_outliers_min_abbrev_combined_wide %>%
  arrange(gene) %>%
  rename_all(underscore_to_space) %>%
  kbl() %>%
  kable_styling(full_width = F,
                bootstrap_options = "bordered")
```




# Annotate with combined cohort abbreviations
```{r}

all_outliers_abbrev_combined_wide <- outliers %>%
  left_join(cohort_codes,
            by=c("comparison_cohort"="cohort_name")) %>%
  select(-pathway_support, -donor_ID,
         -comparison_cohort) %>%
  pivot_wider(names_from = Sample_ID,
              values_from = cohort_code,
              values_fn = collapse_fun,
              values_fill = "")

```

#Out
```{r}


all_outliers_abbrev_combined_wide %>%
  arrange(gene) %>%
  rename_all(underscore_to_space) %>%
  kbl() %>%
  kable_styling(full_width = F,
                bootstrap_options = "bordered")
```

# Summary table for all outliers and low level cohorts
```{r}
n_outliers_detected_by_any_method <- outliers %>%
  select(Sample_ID, gene) %>%
  distinct %>%
  nrow()

n_outliers_with_pathway_support_detected_by_any_method <- outliers %>%
  filter(pathway_support) %>%
  select(Sample_ID, gene) %>%
  distinct %>%
  nrow()
# these have pathway support in at least one cohort


outlier_summary <- outliers %>% 
  group_by(comparison_cohort) %>%
  summarize(n_outliers_detected = n(),
         n_outliers_with_pathway_support = sum(pathway_support),
         pct_outliers_with_pathway_support = 100*n_outliers_with_pathway_support/n_outliers_detected,
         pct_outliers_detected = 100*n_outliers_detected/n_outliers_detected_by_any_method)

outlier_summary_with_totals <- 
bind_rows(outlier_summary,
          tibble(comparison_cohort= " Total",
                 n_outliers_detected = n_outliers_detected_by_any_method,
                 n_outliers_with_pathway_support = n_outliers_with_pathway_support_detected_by_any_method,
                 pct_outliers_with_pathway_support = 100*n_outliers_with_pathway_support_detected_by_any_method/n_outliers_detected_by_any_method))
                 
                 
  
outlier_summary_with_totals %>% 
  rename_all(underscore_to_space) %>%
  kbl(format.args = list(big.mark = ","), digits = c(NA, 0, 0, 0, 0)) %>%
  kable_styling(full_width = F)

```

# Summary table for all outliers and high level cohorts
```{r}
n_outliers_detected_by_any_method <- outliers %>%
  select(Sample_ID, gene) %>%
  distinct %>%
  nrow()

n_outliers_with_pathway_support_detected_by_any_method <- outliers %>%
  filter(pathway_support) %>%
  select(Sample_ID, gene) %>%
  distinct %>%
  nrow()
# these have pathway support in at least one cohort


high_level_outlier_summary <- outliers %>% 
  group_by(high_level_cohort, Sample_ID, gene) %>%
  summarize(pathway_support = any(pathway_support)) %>%
  group_by(high_level_cohort) %>%
  summarize(n_outliers_detected = n(),
         n_outliers_with_pathway_support = sum(pathway_support),
         pct_outliers_with_pathway_support = 100*n_outliers_with_pathway_support/n_outliers_detected,
         pct_outliers_detected = 100*n_outliers_detected/n_outliers_detected_by_any_method)

high_level_outlier_summary_with_totals <- 
bind_rows(high_level_outlier_summary %>%
            arrange(desc(high_level_cohort)),
          tibble(high_level_cohort= " Total",
                 n_outliers_detected = n_outliers_detected_by_any_method,
                 n_outliers_with_pathway_support = n_outliers_with_pathway_support_detected_by_any_method,
                 pct_outliers_with_pathway_support = 100*n_outliers_with_pathway_support_detected_by_any_method/n_outliers_detected_by_any_method))
                 
                 
  
high_level_outlier_summary_with_totals %>% 
  rename_all(underscore_to_space) %>%
  kbl(format.args = list(big.mark = ","), digits = c(NA, 0, 0, 0, 0)) %>%
  kable_styling(full_width = F)

```


## Combined high and low level tables
```{r}

high_low <- bind_rows(
  high_level_outlier_summary_with_totals %>% 
    rename(comparison_cohort=high_level_cohort) %>%
    mutate(index = c(1, 4:7)),
  outlier_summary_with_totals %>%
    filter(str_detect(comparison_cohort, "Treehouse")) %>%
    mutate(index = 2:3)
) %>% 
  arrange(index) %>%
  select(-index)

  

high_low %>%
  rename_all(underscore_to_space) %>%
  kbl(format.args = list(big.mark = ","), digits = c(NA, 0, 0, 0, 0)) %>%
  kable_styling(full_width = F)



```


# REPEAT ANALYSIS USING ONLY OUTLIERS WITH PATHWAY SUPPORT

# Tile plot of outliers with pathway support
```{r fig.height=10, fig.width=10}

ggplot(outliers %>%
         filter(pathway_support)) +
  geom_tile(aes(x=comparison_cohort,
                y=gene, 
                fill = comparison_cohort)) +
  facet_wrap(~Sample_ID,
             nrow = 1) +
  theme(#axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    axis.text.x = element_blank(),
    strip.text.x = element_text(angle = 90),
        ) +
  xlab("")  +
  scale_fill_bright()

  
```


# Heatmap shows number of cohorts in which outlier were  detected
I can make this look better if we decide to use it, but it's non-trivial
```{r fig.height=10, fig.width=6}

pathway_outliers_heatmap_data <- outliers %>%
  filter(pathway_support) %>%
  group_by(Sample_ID, gene) %>%
  summarize(n_outliers = n()) 

ggplot(pathway_outliers_heatmap_data) +
  geom_tile(aes(x=Sample_ID,
                y=gene,
                fill = n_outliers), 
            color = "black")  +
  #theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) 


```


```{r}

raw_pathway_support_outliers_for_venn <- outliers %>%
  filter(pathway_support)  %>%
  mutate(sample_gene = paste(Sample_ID, gene, sep = "_")) %>%
  arrange(comparison_cohort) %>%
  select(sample_gene, comparison_cohort) %>%
  group_split(comparison_cohort)


list_of_pathway_support_outliers_for_venn <-  lapply(raw_pathway_support_outliers_for_venn, function(x) x %>% pull(sample_gene))
names(list_of_pathway_support_outliers_for_venn) <- outliers %>%
  filter(pathway_support) %>%
  arrange(comparison_cohort) %>%
  select(comparison_cohort) %>%
  distinct() %>%
  pull(comparison_cohort)

ggVennDiagram(list_of_pathway_support_outliers_for_venn,
              show_intersect = TRUE)

 
ggVennDiagram(list_of_pathway_support_outliers_for_venn) + 
  scale_fill_distiller(palette = "Reds", direction = 1)


```


# Annotate with combined full cohort names
```{r}


outliers_with_pathway_support_combined_wide <- outliers %>%
  filter(pathway_support) %>%
  select(-pathway_support, -donor_ID) %>%
  pivot_wider(names_from = Sample_ID,
              values_from = comparison_cohort,
              values_fn = collapse_fun)

outliers_with_pathway_support_combined_long <- outliers_with_pathway_support_combined_wide %>%
  pivot_longer(-gene,
               names_to = "Sample_ID",
               values_to = "comparison_cohorts") %>%
  na.omit()
```


# How many outliers with pathway support are present in each combination of cohorts?
```{r}
tabyl(outliers_with_pathway_support_combined_long,
      comparison_cohorts) %>%
  arrange(desc(n)) %>%
  adorn_pct_formatting() %>%
  adorn_totals() %>%
  kbl() %>%
  kable_styling(full_width = F)
```


```{r fig.height=10, fig.width=10}

ggplot(outliers_with_pathway_support_combined_long) +
  geom_tile(aes(x=Sample_ID,
                y=gene,
                fill = comparison_cohorts))  +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
                
n_distinct(outliers_with_pathway_support_combined_long$Sample_ID)                

```



# Annotate with combined cohort abbreviations
```{r}

outliers_with_pathway_support_abbrev_combined_wide <- outliers %>%
  filter(pathway_support) %>%
  left_join(cohort_codes,
            by=c("comparison_cohort"="cohort_name")) %>%
  select(-pathway_support, -donor_ID,
         -comparison_cohort) %>%
  pivot_wider(names_from = Sample_ID,
              values_from = cohort_code,
              values_fn = collapse_fun,
              values_fill = "")

```

# Big table of outliers with pathway support
```{r}


outliers_with_pathway_support_abbrev_combined_wide %>%
  arrange(gene) %>%
  rename_all(underscore_to_space) %>%
  kbl() %>%
  kable_styling(full_width = F,
                bootstrap_options = "bordered")
```

