---
title: "`r gsub('.Rmd', ' ', gsub('_', ' ', knitr::current_input()))`"
author: "`r Sys.getenv('USER')`"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
 pdf_document:
    toc: yes
---
  
 Version 2023.11.21_11.24.15 - focuses on genes found only relative to TCGA, not relative to any other cohort. Previous versions focused on outliers detected relative to TCGA and not treehouse, irrespective of whatever other cohorts they were outliers in


```{r setup, include = FALSE}

library(tidyverse)
library(janitor) 
library(gridExtra)
library(cowplot)
library(ggforce)
library(khroma)
library(ggrepel)
library(kableExtra)

```



```{r}

outliers <- read_tsv("../input_data/druggable_outliers_from_treehouse_and_other_cohorts_2023_11_09-13_46_32_2023.tsv") %>%
  mutate(high_level_cohort = ifelse(str_detect(comparison_cohort, "Treehouse"),
                                    "Treehouse",
                                    comparison_cohort))

         
         
```



# COMPARE DISTRIBUTIONS FOR FOR OUTLIERS ACROSS COHORTS

```{r}

outlier_genes_detected <- unique(outliers$gene)

expr <- read_tsv("../input_data/druggable_TumorCompendium_v11_PolyA_hugo_log2tpm_58581genes_2020-04-09.tsv.gz") %>%
  rename(Sample_ID = TH_id) %>%
  filter(Gene %in% outlier_genes_detected)

stanford_samples  <- read_tsv("../gather_input_data/comparison_to_non_CARE_cohorts/data/TH03_TH34_rollup.sample_list.txt",
                              col_names = "Sample_ID") %>%
  mutate(cohort = "TH03_TH34")

TCGA_samples  <- read_tsv("../gather_input_data/comparison_to_non_CARE_cohorts/data/TCGA_rollup.sample_list.txt",
                              col_names = "Sample_ID") %>%
  mutate(cohort = "TCGA")

PEDAYA_samples  <- read_tsv("../gather_input_data/comparison_to_non_CARE_cohorts/data/PEDAYA_rollup.sample_list.txt",
                              col_names = "Sample_ID") %>%
  mutate(cohort = "PEDAYA")

pan_cancer_samples <- expr %>%
  select(Sample_ID) %>%
  distinct() %>%
  mutate(cohort = "Treehouse_pc")


samples_in_cohorts <- bind_rows(
  stanford_samples,
  TCGA_samples,
  PEDAYA_samples,
  pan_cancer_samples)


tabyl(samples_in_cohorts,
      cohort)
```


# expression in samples not in the compendium
```{r}

rsem_path <- "../input_data/non_compendium_expression"

gene_name_conversion <- read_tsv(file.path(rsem_path,
                                           "EnsGeneID_Hugo_Observed_Conversions.txt"))
relevant_gene_name_conversion <- gene_name_conversion %>%
  filter(HugoID %in% outlier_genes_detected)

rsem_kitchen_sink_data <- tibble(file_name = list.files(
  path = rsem_path,
  pattern = "_rsem_genes.results")) %>%
  rowwise() %>%
  mutate(rsem_raw = list(read_tsv(file.path(rsem_path, file_name),
                                     show_col_types = FALSE
                                     ))) %>%
  unnest(rsem_raw) %>%
  filter(gene_id %in% relevant_gene_name_conversion$EnsGeneID) %>%
  mutate(Sample_ID = str_extract(file_name, "TH[R]?[0-9]{2}_[0-9]{4}_S[0-9]{2}")) %>%
  left_join(relevant_gene_name_conversion, 
            by=c("gene_id"="EnsGeneID")) %>%
    group_by(Sample_ID, HugoID) %>%
    summarize(sum_TPM = sum(TPM),
              n=n()) %>%
    mutate(log2TPM1 = log2(sum_TPM +1))

table(rsem_kitchen_sink_data$n)

patient_expression_from_rsem_files <- rsem_kitchen_sink_data %>%
  select(gene = HugoID,
         log2TPM1,
         Sample_ID)

patient_expression_from_compendia <- outliers %>%
  select(Sample_ID, gene) %>%
  distinct() %>%
  left_join(expr, 
            by=c("Sample_ID", "gene"="Gene")) %>%
  na.omit() # excludes samples not in compendium

patient_expression <- bind_rows(
  patient_expression_from_rsem_files,
  patient_expression_from_compendia)
  
length(outlier_genes_detected)

```



# Calculate statistics for each cohort
```{r}

cohort_thresholds_raw <- left_join(samples_in_cohorts,
                                   expr,
                                   by=c("Sample_ID")) %>%
  group_by(Gene, cohort) %>%
  summarize(q25 = quantile(log2TPM1, 0.25),
            median = median(log2TPM1),
            q75 = quantile(log2TPM1, 0.75),
            IQR = q75-q25,
            up_outlier_threshold = q75 + (1.5*IQR))
  
```

# assess changes for all genes
```{r}
cohort_thresholds <- cohort_thresholds_raw %>%
  pivot_longer(c(-Gene, -cohort)) %>%
  pivot_wider(names_from = cohort, values_from = value) %>%
  mutate(change_in_ped_relative_to_TCGA = 
              (PEDAYA - TCGA) / TCGA,
         change_in_treehouse_relative_to_TCGA = 
            (Treehouse_pc - TCGA) / Treehouse_pc)
```


# changes plotted
```{r}
ggplot(cohort_thresholds) +
  geom_point(aes(x=change_in_ped_relative_to_TCGA,
                 y=change_in_treehouse_relative_to_TCGA,
                 color = name)) +
  scale_color_brewer(palette = "Set1")
```


```{r}
cohort_thresholds %>%
  filter(name %in% c("median", "IQR")) %>%
  select(Gene, name, change_in_ped_relative_to_TCGA) %>%
  pivot_wider(names_from = name,
              values_from = change_in_ped_relative_to_TCGA) %>%
  ggplot + 
  geom_point(aes(x=median, y=IQR)) + 
  ggtitle("The IQR usually increased irrespective of the direction of change of the median", "fraction change_in_ped_relative_to_TCGA")


```

```{r fig.height=8, fig.width=8}

cohort_thresholds_for_plot <- cohort_thresholds %>%
         rename(stat = name) %>%
  mutate(stat = factor(stat, levels = c("q25", "median", "q75", "IQR", "up_outlier_threshold"))) %>%
  group_by(Gene) %>%
         mutate(order_by_me = change_in_ped_relative_to_TCGA[stat == "q25"]) 

# %>%
#   ungroup %>%
#   mutate(Gene = factor(Gene) %>% fct_reorder(Gene, order_by_me, .fun = min))
# levels(cohort_thresholds_for_plot$Gene)


ggplot(cohort_thresholds_for_plot) +
  #geom_tile(aes(x=stat, y= Gene, fill = change_in_ped_relative_to_TCGA)) +
  geom_tile(aes(x=stat, y= fct_reorder(Gene, order_by_me), fill = change_in_ped_relative_to_TCGA)) +
  scale_fill_gradient2("pct change in PEDAYA\nrelative to TCGA")
  
#geom_tile(aes(x=stat, y= fct_reorder(Gene, change_in_ped_relative_to_TCGA), fill = change_in_ped_relative_to_TCGA))
```
# plot boxplots for TCGA and PEDAYA 
```{r}

TP_cohort_expr <- left_join(samples_in_cohorts %>%
                                     filter(cohort %in% c("PEDAYA", "TCGA")),
                                   expr,
                                   by=c("Sample_ID")) 
  
```


```{r fig.height=12, fig.width=6}

TP_cohort_expr_subset <- TP_cohort_expr %>%
  slice_sample(n = 10000)

ggplot(TP_cohort_expr_subset) +
  geom_boxplot(aes(y=cohort, x=log2TPM1,
                   fill = cohort),
               outlier.shape = NA) +
  facet_wrap(~Gene, ncol = 1,
             strip.position = "left") +
 theme(strip.text.y.left = element_text(angle = 0),
       axis.text.y = element_blank(),
        panel.spacing = unit(0.2, "lines"))  +
  scale_fill_bright() 
  
```

```{r}

table(outliers$comparison_cohort)

TCGA_only_outliers <- outliers %>%
  group_by(gene, Sample_ID) %>%
  mutate(TCGA_only = "TCGA" %in% comparison_cohort & 
           n() == 1) %>%
  filter(TCGA_only) %>%
  arrange(Sample_ID, gene)

n_distinct(TCGA_only_outliers$gene)

```



```{r}

TP_cohort_expr_of_TCGA_only_outliers <- TP_cohort_expr %>%
         filter(Gene %in% TCGA_only_outliers$gene)

ggplot(TP_cohort_expr_of_TCGA_only_outliers) +
  geom_boxplot(aes(y=cohort, x=log2TPM1,
                   fill = cohort),
               outlier.shape = NA) +
  facet_wrap(~Gene, ncol = 1,
             strip.position = "left") +
 theme(strip.text.y.left = element_text(angle = 0),
       axis.text.y = element_blank(),
        panel.spacing = unit(0.2, "lines"))  +
  scale_fill_bright() 

manual_annotation_of_pedaya_relative_to_TCGA <- tribble(
  ~Gene, ~IQR, ~shift,
  "BCL6", "wider", "none",
  "CCND1", "wider", "lower",
  "CCND3", "wider", "higher",
  "CDK9", "wider", "higher",
  "ETV1", "wider", "lower",
  "FGFR1", "wider", "higher",
  "FGFR2", "wider", "lower",
  "FGFR3", "wider", "lower",
  "GATA2", "wider", "none",
  "HMOX1", "wider", "lower",
  "IGF2", "wider", "none",
  "KDR", "wider", "lower",
  "MTOR", "similar", "higher",
  "PARP2", "similar", "higher",
  "PDGFRA", "wider", "lower",
  "PIK3CD", "wider", "higher",
  "PTCH1", "wider", "higher",
  "RPTOR", "similar", "higher",
  "TSC2", "wider", "higher",
  "VEGFA", "wider", "lower") %>%
  filter(Gene %in% TCGA_only_outliers$gene)

# cat(paste0("\"", unique(sort(TCGA_only_outliers$gene)), "\""), sep = ",\n")
  
```


```{r}

anno_cohort_thresholds_of_TCGA_only_outliers <-
  left_join(cohort_thresholds_for_plot %>%
              filter(Gene %in% TCGA_only_outliers$gene),
            manual_annotation_of_pedaya_relative_to_TCGA)


  ggplot(anno_cohort_thresholds_of_TCGA_only_outliers  %>%
              filter(stat == "median"),
         aes(x=PEDAYA, y=TCGA, color = shift)) +
  #geom_histogram(aes(x=change_in_ped_relative_to_TCGA))
  geom_point() +
  geom_label_repel(aes(label = Gene)) +
  geom_abline() +
  coord_equal() +
  #scale_y_continuous(breaks = c(0:3)) +
  expand_limits(y=0,x=0) +
  ggtitle("medians for genes that are outliers in TCGA and not Treehouse") +
  scale_color_highcontrast() 

# cohort_thresholds_for_plot %>%
#   filter(Gene %in% TCGA_only_outliers$gene) %>%
#   filter(stat == "median") %>%
#   arrange(change_in_ped_relative_to_TCGA)

```


# Compare IQRs
```{r}

ggplot(anno_cohort_thresholds_of_TCGA_only_outliers  %>%
         filter(stat == "IQR"),
       aes(x=PEDAYA, y=TCGA, color = IQR)) +
  #geom_histogram(aes(x=change_in_ped_relative_to_TCGA))
  geom_point() +
  geom_text_repel(aes(label = Gene)) +
  geom_abline() +
  coord_equal() +
  scale_y_continuous(breaks = c(0:3)) +
  expand_limits(y=0,x=0) +
  ggtitle("IQRs for genes that are outliers in TCGA and not Treehouse") +
  scale_color_highcontrast() 

```

# Review data to  identify a good quantitative IQR cutoff
```{r}

anno_cohort_thresholds_of_TCGA_only_outliers %>%
  mutate(abs_change = PEDAYA-TCGA) %>%
  select(Gene, stat, PEDAYA, TCGA,
         pct_change = change_in_ped_relative_to_TCGA,
         abs_change,
         IQR, shift) %>%
  arrange(abs_change) %>% # change to pct_change to view alternative consideration
  adorn_pct_formatting(,,,pct_change) %>%
  filter(stat == "IQR") %>%
  kbl(digits = c(NA, NA, 1, 1, NA, 2, NA, NA)) %>%
  kable_styling(full_width = F)

# IQR change greater than 0.5 log2TPM1

# percent cutoff doesn't really work, because some relatively small pcts are large in log2tpm1 space,
# e.g. FGFR3, BCL6, wider
  
```




# Review data to see identify a good quantitative shift cutoff
```{r}

anno_cohort_thresholds_of_TCGA_only_outliers %>%
  mutate(abs_change = PEDAYA-TCGA) %>%
  select(Gene, stat, PEDAYA, TCGA,
         pct_change = change_in_ped_relative_to_TCGA,
         abs_change,
         IQR, shift) %>%
  arrange(abs_change) %>%
  adorn_pct_formatting(,,,pct_change) %>%
  filter(stat == "median") %>%
  kbl(digits = c(NA, NA, 1, 1, NA, 2, NA, NA)) %>%
  kable_styling(full_width = F)

# if PEDAYA median is 0.25 higher or lower than TCGA median, the shift is higher or lower, respectively

# percent cutoff doesn't really work, because some relatively small pcts are large in log2tpm1 space,

  
```


# Summary of differences

```{r}
#manual_annotation_of_pedaya_relative_to_TCGA

tabyl(manual_annotation_of_pedaya_relative_to_TCGA, IQR)

tabyl(manual_annotation_of_pedaya_relative_to_TCGA, shift)
```

# Text summary

Each sample in the study was compared to TCGA, PEDAYA, Stanford and the full Treehouse compendium.
Often genes would be identified as outliers with respect to multiple cohorts. 
We investigated the exceptions, genes found as outliers with respect to only one cohort. 
The cohort with the largest number of uniquely detected outliers was TCGA, 
We analyzed the distribution of expression for the outlier genes that were 
identified in any samples by comparison with TCGA and no other cohorts. Because the addition of pediatric samples is the primary difference between TCGA and the other cohorts, we focussed on comparing the PEDAYA cohort and TCGA

Of those 16, 15 had wider distributions (>0.5log2(TPM+1) bigger) in PEDAYA than TCGA, while one was similar. 
For 7, the median was higher (by more than 0.25 log2TPM1) in PEDAYA than TCGA.
For 6, the median was lower (by more than 0.25 log2TPM1) in PEDAYA than TCGA. 
For 3, the medians were similar in PEDAYA and TCGA. 


The Treehouse compendium is 77% TCGA (which is 96% adult). The remaining 23% of the Treehouse compendium is 97% PEDAYA (<=30). In the genes we looked at, the changes in distribution between TCGA and Treehouse compendium were consistent with the effects of adding the distribution of PEDAYA samples. 
The differences in the treehouse compendium from TCGA is mostly due to the addition of PEDAYA samples. 
