---
title: "`r gsub('.Rmd', ' ', gsub('_', ' ', knitr::current_input()))`"
author: "`r Sys.getenv('USER')`"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: yes
---
  


```{r setup, include = FALSE}

library(tidyverse)
library(janitor) 
library(gridExtra)
library(cowplot)
library(ggforce)
library(khroma)
library(ggrepel)
library(kableExtra)

```



```{r}

outliers <- read_tsv("../input_data/druggable_outliers_from_treehouse_and_other_cohorts_2023_11_09-13_46_32_2023.tsv") %>%
  mutate(high_level_cohort = ifelse(str_detect(comparison_cohort, "Treehouse"),
                                    "Treehouse",
                                    comparison_cohort))

         
         
```



# COMPARE DISTRIBUTIONS FOR FOR OUTLIERS ACROSS COHORTS

```{r}

outlier_genes_detected <- unique(outliers$gene)

expr <- read_tsv("../input_data/druggable_TumorCompendium_v11_PolyA_hugo_log2tpm_58581genes_2020-04-09.tsv.gz") %>%
  rename(Sample_ID = TH_id) %>%
  filter(Gene %in% outlier_genes_detected)

stanford_samples  <- read_tsv("../gather_input_data/comparison_to_non_CARE_cohorts/data/TH03_TH34_rollup.sample_list.txt",
                              col_names = "Sample_ID") %>%
  mutate(cohort = "TH03_TH34")

TCGA_samples  <- read_tsv("../gather_input_data/comparison_to_non_CARE_cohorts/data/TCGA_rollup.sample_list.txt",
                              col_names = "Sample_ID") %>%
  mutate(cohort = "TCGA")

PEDAYA_samples  <- read_tsv("../gather_input_data/comparison_to_non_CARE_cohorts/data/PEDAYA_rollup.sample_list.txt",
                              col_names = "Sample_ID") %>%
  mutate(cohort = "PEDAYA")

pan_cancer_samples <- expr %>%
  select(Sample_ID) %>%
  distinct() %>%
  mutate(cohort = "Treehouse_pc")


samples_in_cohorts <- bind_rows(
  stanford_samples,
  TCGA_samples,
  PEDAYA_samples,
  pan_cancer_samples)


tabyl(samples_in_cohorts,
      cohort)
```


# expression in samples not in the compendium
```{r}

rsem_path <- "../input_data/non_compendium_expression"

gene_name_conversion <- read_tsv(file.path(rsem_path,
                                           "EnsGeneID_Hugo_Observed_Conversions.txt"))
relevant_gene_name_conversion <- gene_name_conversion %>%
  filter(HugoID %in% outlier_genes_detected)

rsem_kitchen_sink_data <- tibble(file_name = list.files(
  path = rsem_path,
  pattern = "_rsem_genes.results")) %>%
  rowwise() %>%
  mutate(rsem_raw = list(read_tsv(file.path(rsem_path, file_name),
                                     show_col_types = FALSE
                                     ))) %>%
  unnest(rsem_raw) %>%
  filter(gene_id %in% relevant_gene_name_conversion$EnsGeneID) %>%
  mutate(Sample_ID = str_extract(file_name, "TH[R]?[0-9]{2}_[0-9]{4}_S[0-9]{2}")) %>%
  left_join(relevant_gene_name_conversion, 
            by=c("gene_id"="EnsGeneID")) %>%
    group_by(Sample_ID, HugoID) %>%
    summarize(sum_TPM = sum(TPM),
              n=n()) %>%
    mutate(log2TPM1 = log2(sum_TPM +1))

table(rsem_kitchen_sink_data$n)

patient_expression_from_rsem_files <- rsem_kitchen_sink_data %>%
  select(gene = HugoID,
         log2TPM1,
         Sample_ID)

patient_expression_from_compendia <- outliers %>%
  select(Sample_ID, gene) %>%
  distinct() %>%
  left_join(expr, 
            by=c("Sample_ID", "gene"="Gene")) %>%
  na.omit() # excludes samples not in compendium

patient_expression <- bind_rows(
  patient_expression_from_rsem_files,
  patient_expression_from_compendia)
  
length(outlier_genes_detected)

```


```{r}
outliers$Sample_ID[ ! outliers$Sample_ID %in% expr$Sample_ID] %>% unique()
```




# How many colors do i need
```{r}


outliers %>%
  group_by(gene) %>%
  summarize(n_samples = length(unique(Sample_ID))) %>%
  arrange(desc(n_samples))


```
# Calculate statistics for each cohort
```{r}

cohort_thresholds_raw <- left_join(samples_in_cohorts,
                                   expr,
                                   by=c("Sample_ID")) %>%
  group_by(Gene, cohort) %>%
  summarize(q25 = quantile(log2TPM1, 0.25),
            median = median(log2TPM1),
            q75 = quantile(log2TPM1, 0.75),
            IQR = q75-q25,
            up_outlier_threshold = q75 + (1.5*IQR))
  
```

# pediatric vs TCGA for one gene
```{r}

this_gene <- "ETV1"

cohort_thresholds_raw %>%
  filter(Gene == this_gene) %>%
  group_by(Gene) %>%
  pivot_longer(c(-Gene, -cohort)) %>%
  pivot_wider(names_from = cohort, values_from = value) %>%
  mutate(change_in_ped_relative_to_TCGA = 
              (PEDAYA - TCGA) / TCGA,
         change_in_treehouse_relative_to_TCGA = 
            (Treehouse_pc - TCGA) / Treehouse_pc) %>%
  select(-TH03_TH34)


# # is the biggest change to the median or to the IQR?
#   summarize(q_25_change_in_ped_relative_to_TCGA = 
#               (q25[cohort == "PEDAYA"] - q25[cohort == "TCGA"]) / q25[cohort == "TCGA"])

```
# asess changes for all genes
```{r}
cohort_thresholds <- cohort_thresholds_raw %>%
  pivot_longer(c(-Gene, -cohort)) %>%
  pivot_wider(names_from = cohort, values_from = value) %>%
  mutate(change_in_ped_relative_to_TCGA = 
              (PEDAYA - TCGA) / TCGA,
         change_in_treehouse_relative_to_TCGA = 
            (Treehouse_pc - TCGA) / Treehouse_pc)
```


# changes plotted
```{r}
ggplot(cohort_thresholds) +
  geom_point(aes(x=change_in_ped_relative_to_TCGA,
                 y=change_in_treehouse_relative_to_TCGA,
                 color = name)) +
  scale_color_brewer(palette = "Set1")
```


```{r}
cohort_thresholds %>%
  filter(name %in% c("median", "IQR")) %>%
  select(Gene, name, change_in_ped_relative_to_TCGA) %>%
  pivot_wider(names_from = name,
              values_from = change_in_ped_relative_to_TCGA) %>%
  ggplot + 
  geom_point(aes(x=median, y=IQR)) + 
  ggtitle("The IQR usually increased irrespective of the direction of change of the median", "fraction change_in_ped_relative_to_TCGA")


```

```{r fig.height=8, fig.width=8}

cohort_thresholds_for_plot <- cohort_thresholds %>%
         rename(stat = name) %>%
  mutate(stat = factor(stat, levels = c("q25", "median", "q75", "IQR", "up_outlier_threshold"))) %>%
  group_by(Gene) %>%
         mutate(order_by_me = change_in_ped_relative_to_TCGA[stat == "q25"]) 

# %>%
#   ungroup %>%
#   mutate(Gene = factor(Gene) %>% fct_reorder(Gene, order_by_me, .fun = min))
# levels(cohort_thresholds_for_plot$Gene)


ggplot(cohort_thresholds_for_plot) +
  #geom_tile(aes(x=stat, y= Gene, fill = change_in_ped_relative_to_TCGA)) +
  geom_tile(aes(x=stat, y= fct_reorder(Gene, order_by_me), fill = change_in_ped_relative_to_TCGA)) +
  scale_fill_gradient2("pct change in PEDAYA\nrelative to TCGA")
  
#geom_tile(aes(x=stat, y= fct_reorder(Gene, change_in_ped_relative_to_TCGA), fill = change_in_ped_relative_to_TCGA))
```
# plot boxplots for TCGA and PEDAYA 
```{r}

TP_cohort_expr <- left_join(samples_in_cohorts %>%
                                     filter(cohort %in% c("PEDAYA", "TCGA")),
                                   expr,
                                   by=c("Sample_ID")) 
  
```


```{r fig.height=12, fig.width=6}

TP_cohort_expr_subset <- TP_cohort_expr %>%
  slice_sample(n = 10000)

ggplot(TP_cohort_expr_subset) +
  geom_boxplot(aes(y=cohort, x=log2TPM1,
                   fill = cohort),
               outlier.shape = NA) +
  facet_wrap(~Gene, ncol = 1,
             strip.position = "left") +
 theme(strip.text.y.left = element_text(angle = 0),
       axis.text.y = element_blank(),
        panel.spacing = unit(0.2, "lines"))  +
  scale_fill_bright() 
  
```

```{r}

TCGA_not_Treehouse_pc_outliers <- outliers %>%
  group_by(gene, Sample_ID) %>%
  mutate(TCGA_not_Treehouse_pc = "TCGA" %in% comparison_cohort & 
           ! "Treehouse_pc" %in% comparison_cohort) %>%
  filter(TCGA_not_Treehouse_pc) %>%
  arrange(Sample_ID, gene)

```



```{r}

TP_cohort_expr_of_TCGA_not_Treehouse_pc_outliers <- TP_cohort_expr %>%
         filter(Gene %in% TCGA_not_Treehouse_pc_outliers$gene)

ggplot(TP_cohort_expr_of_TCGA_not_Treehouse_pc_outliers) +
  geom_boxplot(aes(y=cohort, x=log2TPM1,
                   fill = cohort),
               outlier.shape = NA) +
  facet_wrap(~Gene, ncol = 1,
             strip.position = "left") +
 theme(strip.text.y.left = element_text(angle = 0),
       axis.text.y = element_blank(),
        panel.spacing = unit(0.2, "lines"))  +
  scale_fill_bright() 

manual_annotation_of_pedaya_relative_to_TCGA <- tribble(
  ~Gene, ~IQR, ~shift,
  "BCL6", "wider", "none",
  "CCND1", "wider", "lower",
  "CCND3", "wider", "higher",
  "CDK9", "wider", "higher",
  "ETV1", "wider", "lower",
  "FGFR1", "wider", "higher",
  "FGFR2", "wider", "lower",
  "FGFR3", "wider", "lower",
  "GATA2", "wider", "none",
  "HMOX1", "wider", "lower",
  "IGF2", "wider", "none",
  "KDR", "wider", "lower",
  "MTOR", "similar", "higher",
  "PARP2", "similar", "higher",
  "PDGFRA", "wider", "lower",
  "PIK3CD", "wider", "higher",
  "PTCH1", "wider", "higher",
  "RPTOR", "similar", "higher",
  "TSC2", "wider", "higher",
  "VEGFA", "wider", "lower")

# cat(paste0("\"", unique(sort(TCGA_not_Treehouse_pc_outliers$gene)), "\""), sep = ",\n")
  
```


```{r}

anno_cohort_thresholds_of_TCGA_not_Treehouse_pc_outliers <-
  left_join(cohort_thresholds_for_plot %>%
              filter(Gene %in% TCGA_not_Treehouse_pc_outliers$gene),
            manual_annotation_of_pedaya_relative_to_TCGA)


  ggplot(anno_cohort_thresholds_of_TCGA_not_Treehouse_pc_outliers  %>%
              filter(stat == "median"),
         aes(x=PEDAYA, y=TCGA, color = shift)) +
  #geom_histogram(aes(x=change_in_ped_relative_to_TCGA))
  geom_point() +
  geom_label_repel(aes(label = Gene)) +
  geom_abline() +
  coord_equal() +
  #scale_y_continuous(breaks = c(0:3)) +
  expand_limits(y=0,x=0) +
  ggtitle("medians for genes that are outliers in TCGA and not Treehouse") +
  scale_color_highcontrast() 

cohort_thresholds_for_plot %>%
  filter(Gene %in% TCGA_not_Treehouse_pc_outliers$gene) %>%
  filter(stat == "median") %>%
  arrange(change_in_ped_relative_to_TCGA)

```


# Compare IQRs
```{r}

ggplot(anno_cohort_thresholds_of_TCGA_not_Treehouse_pc_outliers  %>%
         filter(stat == "IQR"),
       aes(x=PEDAYA, y=TCGA, color = IQR)) +
  #geom_histogram(aes(x=change_in_ped_relative_to_TCGA))
  geom_point() +
  geom_text_repel(aes(label = Gene)) +
  geom_abline() +
  coord_equal() +
  scale_y_continuous(breaks = c(0:3)) +
  expand_limits(y=0,x=0) +
  ggtitle("IQRs for genes that are outliers in TCGA and not Treehouse") +
  scale_color_highcontrast() 

```

# Review data to  identify a good quantitative IQR cutoff
```{r}

anno_cohort_thresholds_of_TCGA_not_Treehouse_pc_outliers %>%
  mutate(abs_change = PEDAYA-TCGA) %>%
  select(Gene, stat, PEDAYA, TCGA,
         pct_change = change_in_ped_relative_to_TCGA,
         abs_change,
         IQR, shift) %>%
  arrange(abs_change) %>% # change to pct_change to view alternative consideration
  adorn_pct_formatting(,,,pct_change) %>%
  filter(stat == "IQR") %>%
  kbl(digits = c(NA, NA, 1, 1, NA, 2, NA, NA)) %>%
  kable_styling(full_width = F)

# IQR change greater than 0.5 log2TPM1

# percent cutoff doesn't really work, because some relatively small pcts are large in log2tpm1 space,
# e.g. FGFR3, BCL6, wider
  
```




# Review data to see identify a good quantitative shift cutoff
```{r}

anno_cohort_thresholds_of_TCGA_not_Treehouse_pc_outliers %>%
  mutate(abs_change = PEDAYA-TCGA) %>%
  select(Gene, stat, PEDAYA, TCGA,
         pct_change = change_in_ped_relative_to_TCGA,
         abs_change,
         IQR, shift) %>%
  arrange(abs_change) %>%
  adorn_pct_formatting(,,,pct_change) %>%
  filter(stat == "median") %>%
  kbl(digits = c(NA, NA, 1, 1, NA, 2, NA, NA)) %>%
  kable_styling(full_width = F)

# if PEDAYA median is 0.25 higher or lower than TCGA median, the shift is higher or lower, respectively

# percent cutoff doesn't really work, because some relatively small pcts are large in log2tpm1 space,

  
```


# Summary of differences

```{r}
manual_annotation_of_pedaya_relative_to_TCGA

tabyl(manual_annotation_of_pedaya_relative_to_TCGA, IQR)

tabyl(manual_annotation_of_pedaya_relative_to_TCGA, shift)
```

# Text summary

We analyzed the distribution of expression for the outlier genes that were 
identified in any samples by comparison with TCGA but not with comparison 
to Treehouse pan cancer cohort.

Of those 20, 17 had wider distributions, while three were similar. 
For 9, the median was higher (by more than 0.25 log2TPM1) in PEDAYA than TCGA.
For 8, the median was lower (by more than 0.25 log2TPM1) in PEDAYA than TCGA. 
For 3, the medians were similar in PEDAYA and TCGA. 


The Treehouse compendium is 77% TCGA (which is 96% adult). The remaining 23% of the Treehouse compendium is 97% PEDAYA (<=30). In the genes we looked at, the changes in distribution between TCGA and Treehouse compendium were consistent with the effects of adding the distribution of PEDAYA samples. 
The differences in the treehouse compendium from TCGA is mostly due to the addition of PEDAYA samples. 
