---
title: "`r gsub('.Rmd', ' ', gsub('_', ' ', knitr::current_input()))`"
author: "`r Sys.getenv('USER')`"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
 html_document:
    toc: yes
---
  
Version 2023.11.21_11.24.15 - focuses on genes found only relative to TCGA, not relative to any other cohort. Previous versions focused on outliers detected relative to TCGA and not treehouse, irrespective of whatever other cohorts they were outliers in
 
Version 2023.11.29_09.19.33 - focuses on all druggable genes, not only genes that were outliers vs one cohort or another


```{r setup, include = FALSE}

library(tidyverse)
library(janitor) 
library(gridExtra)
library(cowplot)
library(ggforce)
library(khroma)
library(ggrepel)
library(kableExtra)

```


# LOAD AND UPDATE v11 METADATA

```{r}

kbl_me <- function(x) {
    adorn_pct_formatting(x) %>% 
    kbl(format.args = list(big.mark = ","), digits = c(NA, 0, NA)) %>%
    kable_styling(full_width = F) 
}

```

# load public compendium data
```{r}

v11_metadata <- read_tsv("https://xena.treehouse.gi.ucsc.edu/download/clinical_TumorCompendium_v11_PolyA_2020-04-09.tsv.gz")

tabyl(v11_metadata$pedaya) %>% kbl_me

v11_metadata %>%
  filter(is.na(pedaya),
         ! is.na(age_at_dx)) %>%
  nrow

v11_metadata_min <- v11_metadata %>%
  select(th_sampleid, age_at_dx, pedaya) %>%
  mutate(pedaya = ifelse(is.na(pedaya) & age_at_dx <30,
                          "Yes, age < 30 years",
                         pedaya))

tabyl(v11_metadata_min$pedaya)  %>% kbl_me
```
# load redcap data
```{r}

redcap_metadata_v11 <- read_tsv("/Users/hbeale/Documents/Dropbox/ucsc/operations/redcap/redcap_combined_nonprivate_sample_donor_data_downloaded_via_API.v10.2023-11-09_09.33.56.txt.gz") %>%
  filter(THid %in% v11_metadata$th_sampleid)

# redcap_metadata_v11 %>%
#   pull(age_at_dx) %>%
#   summary()


redcap_metadata_v11 %>%
  tabyl(pedaya)   %>% 
  kbl_me

redcap_metadata_v11 %>%
  filter(is.na(pedaya),
         ! is.na(age_at_dx)) %>%
  nrow

redcap_metadata_v11_min <- redcap_metadata_v11 %>%
  select(THid, pedaya, age_at_dx) %>%
  mutate(pedaya = ifelse(is.na(pedaya) & age_at_dx <30,
                         "Yes, age < 30 years",
                                pedaya))
redcap_metadata_v11_min %>%
  tabyl(pedaya)  %>% kbl_me


```

# update v11 compendium metadata from redcap
```{r}

combined_metadata <- left_join(v11_metadata_min, 
                               redcap_metadata_v11_min,
                               by=c("th_sampleid"="THid"),
                               suffix = c("_compendium", "_redcap")) 

combined_metadata %>%
  filter(pedaya_compendium != pedaya_redcap) %>%
  kbl() %>%
  kable_styling()

combined_metadata %>%
  filter(is.na(pedaya_compendium),
         !is.na(pedaya_redcap))  %>%
  nrow

combined_metadata %>%
  filter(is.na(pedaya_compendium),
         !is.na(pedaya_redcap))  %>%
  head %>%
  kbl() %>%
  kable_styling()

combined_metadata %>%
  filter(is.na(pedaya_redcap),
         !is.na(pedaya_compendium)) %>%
  kbl() %>%
  kable_styling()


use_redcap <- c("TARGET-40-0A4I4O-01A-01R",
                combined_metadata %>%
                  filter(is.na(pedaya_compendium),
                         !is.na(pedaya_redcap))  %>%
                  pull(th_sampleid))

use_compendium_data <- c("TH27_2347_S02") # and everything else not from redcap

age_updated_v11 <- bind_rows(
  redcap_metadata_v11_min %>%
    filter(THid %in% use_redcap) %>%
    rename(th_sampleid = THid),
  v11_metadata_min %>%
    filter(!th_sampleid %in% use_redcap)) %>%
  rename(Sample_ID = th_sampleid)


```


# Update disease
```{r}

updated_v11 <- left_join(
  age_updated_v11,
redcap_metadata_v11 %>%
  select(THid, disease),
by=c("Sample_ID" = "THid" ))

```


# COMPARE DISTRIBUTIONS FOR  OUTLIERS ACROSS COHORTS

```{r}

stanford_samples  <- read_tsv("../gather_input_data/comparison_to_non_CARE_cohorts/data/TH03_TH34_rollup.sample_list.txt",
                              col_names = "Sample_ID") %>%
  mutate(cohort = "TH03_TH34")

TCGA_samples  <- read_tsv("../gather_input_data/comparison_to_non_CARE_cohorts/data/TCGA_rollup.sample_list.txt",
                              col_names = "Sample_ID") %>%
  mutate(cohort = "TCGA")

PEDAYA_samples  <- read_tsv("../gather_input_data/comparison_to_non_CARE_cohorts/data/PEDAYA_rollup.sample_list.txt",
                              col_names = "Sample_ID") %>%
  mutate(cohort = "PEDAYA")

pan_cancer_samples <- updated_v11 %>%
  select(Sample_ID) %>%
  distinct() %>%
  mutate(cohort = "Treehouse_pc")

cohorts_in_order <- c("Treehouse_pc", "TCGA", "PEDAYA", "TH03_TH34")

dput(unique(samples_in_cohorts$cohort))
samples_in_cohorts <- bind_rows(
  stanford_samples,
  TCGA_samples,
  PEDAYA_samples,
  pan_cancer_samples) %>%
  left_join(updated_v11) %>%
  mutate(cohort = factor(cohort, levels = cohorts_in_order))


tabyl(samples_in_cohorts,
      cohort)
```


```{r}

samples_in_cohorts %>%
  group_by(cohort) %>%
  mutate(cohort_total = n()) %>%
  ungroup %>%
  group_by(disease, cohort) %>%
  summarize(n = n(),
            pct_of_cohort = round(100*n/cohort_total, 2),
            stat_char = paste0(n, " (", pct_of_cohort, "%)")) %>%
  ungroup %>%
  arrange(desc(n)) %>%
  mutate(disease = factor(disease, levels = unique(disease))) %>%
  select(disease, cohort, stat_char)  %>%
  distinct() %>%
    # dplyr::group_by(disease, cohort) %>%
    # dplyr::summarise(n = dplyr::n(), .groups = "drop") %>%
    # dplyr::filter(n > 1L)
  pivot_wider(names_from = cohort, values_from = stat_char, values_fill = "0")


```


```{r}

samples_in_cohorts %>%
  tabyl(disease, cohort) %>%
  arrange(desc(PEDAYA)) %>%
  adorn_title() %>%
  kbl() %>%
  kable_styling(full_width = F)

```

```{r}

ggplot(samples_in_cohorts) +
  geom_bar(aes(y=cohort, fill = pedaya)) +
  scale_fill_bright()

```
```{r}

most_common_ped_diseases <- samples_in_cohorts %>%
  filter(cohort == "PEDAYA") %>%
  tabyl(disease) %>%
  arrange(desc(n)) %>%
  slice_head(n=15)

disease_plot <- samples_in_cohorts %>%
  mutate(common_ped_cancers = ifelse(disease %in% most_common_ped_diseases$disease,
                                     disease,
                                     "other") %>%
           factor() %>%
           fct_relevel("other", after=Inf))
         

ggplot(disease_plot) +
  geom_bar(aes(y=cohort, fill = common_ped_cancers)) +
  scale_fill_bright()

```

