---
title: "`r gsub('.Rmd', ' ', gsub('_', ' ', knitr::current_input()))`"
author: "`r Sys.getenv('USER')`"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
 html_document:
    toc: yes
---
  
Version 2023.11.21_11.24.15 - focuses on genes found only relative to TCGA, not relative to any other cohort. Previous versions focused on outliers detected relative to TCGA and not treehouse, irrespective of whatever other cohorts they were outliers in
 
Version 2023.11.29_09.19.33 - focuses on all druggable genes, not only genes that were outliers vs one cohort or another


```{r setup, include = FALSE}

library(tidyverse)
library(janitor) 
library(gridExtra)
library(cowplot)
library(ggforce)
library(khroma)
library(ggrepel)
library(kableExtra)

```


# LOAD AND UPDATE v11 METADATA

```{r}

kbl_me <- function(x) {
    adorn_pct_formatting(x) %>% 
    kbl(format.args = list(big.mark = ","), digits = c(NA, 0, NA)) %>%
    kable_styling(full_width = F) 
}

```

# load public compendium data
```{r}

v11_metadata <- read_tsv("https://xena.treehouse.gi.ucsc.edu/download/clinical_TumorCompendium_v11_PolyA_2020-04-09.tsv.gz")

tabyl(v11_metadata$pedaya) %>% kbl_me

v11_metadata %>%
  filter(is.na(pedaya),
         ! is.na(age_at_dx)) %>%
  nrow

v11_metadata_min <- v11_metadata %>%
  select(th_sampleid, age_at_dx, pedaya) %>%
  mutate(pedaya = ifelse(is.na(pedaya) & age_at_dx <30,
                          "Yes, age < 30 years",
                         pedaya))

tabyl(v11_metadata_min$pedaya)  %>% kbl_me
```
# load redcap data
```{r}

redcap_metadata_v11 <- read_tsv("/Users/hbeale/Documents/Dropbox/ucsc/operations/redcap/redcap_combined_nonprivate_sample_donor_data_downloaded_via_API.v10.2023-11-09_09.33.56.txt.gz") %>%
  filter(THid %in% v11$th_sampleid)

# redcap_metadata_v11 %>%
#   pull(age_at_dx) %>%
#   summary()


redcap_metadata_v11 %>%
  tabyl(pedaya)   %>% 
  kbl_me

redcap_metadata_v11 %>%
  filter(is.na(pedaya),
         ! is.na(age_at_dx)) %>%
  nrow

redcap_metadata_v11_min <- redcap_metadata_v11 %>%
  select(THid, pedaya, age_at_dx) %>%
  mutate(pedaya = ifelse(is.na(pedaya) & age_at_dx <30,
                         "Yes, age < 30 years",
                                pedaya))
redcap_metadata_v11_min %>%
  tabyl(pedaya)  %>% kbl_me


```

# update v11 compendium metadata from redcap
```{r}

combined_metadata <- left_join(v11_metadata_min, 
                               redcap_metadata_v11_min,
                               by=c("th_sampleid"="THid"),
                               suffix = c("_compendium", "_redcap")) 

combined_metadata %>%
  filter(pedaya_compendium != pedaya_redcap) %>%
  kbl() %>%
  kable_styling()

combined_metadata %>%
  filter(is.na(pedaya_compendium),
         !is.na(pedaya_redcap))  %>%
  nrow

combined_metadata %>%
  filter(is.na(pedaya_compendium),
         !is.na(pedaya_redcap))  %>%
  head %>%
  kbl() %>%
  kable_styling()

combined_metadata %>%
  filter(is.na(pedaya_redcap),
         !is.na(pedaya_compendium)) %>%
  kbl() %>%
  kable_styling()


use_redcap <- c("TARGET-40-0A4I4O-01A-01R",
                combined_metadata %>%
                  filter(is.na(pedaya_compendium),
                         !is.na(pedaya_redcap))  %>%
                  pull(th_sampleid))

use_compendium_data <- c("TH27_2347_S02") # and everything else not from redcap

updated_v11 <- bind_rows(
  redcap_metadata_v11_min %>%
    filter(THid %in% use_redcap) %>%
    rename(th_sampleid = THid),
  v11_metadata_min %>%
    filter(!th_sampleid %in% use_redcap)) %>%
  rename(Sample_ID = th_sampleid)


```




# COMPARE DISTRIBUTIONS FOR  OUTLIERS ACROSS COHORTS

```{r}

stanford_samples  <- read_tsv("../gather_input_data/comparison_to_non_CARE_cohorts/data/TH03_TH34_rollup.sample_list.txt",
                              col_names = "Sample_ID") %>%
  mutate(cohort = "TH03_TH34")

TCGA_samples  <- read_tsv("../gather_input_data/comparison_to_non_CARE_cohorts/data/TCGA_rollup.sample_list.txt",
                              col_names = "Sample_ID") %>%
  mutate(cohort = "TCGA")

PEDAYA_samples  <- read_tsv("../gather_input_data/comparison_to_non_CARE_cohorts/data/PEDAYA_rollup.sample_list.txt",
                              col_names = "Sample_ID") %>%
  mutate(cohort = "PEDAYA")

pan_cancer_samples <- updated_v11 %>%
  select(Sample_ID) %>%
  distinct() %>%
  mutate(cohort = "Treehouse_pc")


samples_in_cohorts <- bind_rows(
  stanford_samples,
  TCGA_samples,
  PEDAYA_samples,
  pan_cancer_samples)


tabyl(samples_in_cohorts,
      cohort)
```
```{r}

```




# Calculate statistics for each cohort
```{r}

cohort_thresholds_raw <- left_join(samples_in_cohorts,
                                   expr,
                                   by=c("Sample_ID")) %>%
  group_by(Gene, cohort) %>%
  summarize(q25 = quantile(log2TPM1, 0.25),
            median = median(log2TPM1),
            q75 = quantile(log2TPM1, 0.75),
            IQR = q75-q25,
            up_outlier_threshold = q75 + (1.5*IQR))
  
```

# assess changes for all genes
```{r}
cohort_thresholds <- cohort_thresholds_raw %>%
  pivot_longer(c(-Gene, -cohort)) %>%
  pivot_wider(names_from = cohort, values_from = value) %>%
  mutate(frac_change_in_ped_relative_to_TCGA = 
              (PEDAYA - TCGA) / TCGA,
         change_in_treehouse_relative_to_TCGA = 
            (Treehouse_pc - TCGA) / Treehouse_pc)
```


# changes plotted
```{r}
ggplot(cohort_thresholds) +
  geom_point(aes(x=frac_change_in_ped_relative_to_TCGA,
                 y=change_in_treehouse_relative_to_TCGA,
                 color = name)) +
  scale_color_brewer(palette = "Set1")
```


```{r}
cohort_thresholds %>%
  filter(name %in% c("median", "IQR")) %>%
  select(Gene, name, frac_change_in_ped_relative_to_TCGA) %>%
  pivot_wider(names_from = name,
              values_from = frac_change_in_ped_relative_to_TCGA) %>%
  ggplot + 
  geom_point(aes(x=median, y=IQR)) + 
  ggtitle("The IQR usually increased irrespective of the direction of change of the median", "fraction change_in_ped_relative_to_TCGA")


```

```{r fig.height=8, fig.width=8}

cohort_thresholds_for_plot <- cohort_thresholds %>%
         rename(stat = name) %>%
  mutate(stat = factor(stat, levels = c("q25", "median", "q75", "IQR", "up_outlier_threshold"))) %>%
  group_by(Gene) %>%
         mutate(order_by_me = frac_change_in_ped_relative_to_TCGA[stat == "q25"]) 

# %>%
#   ungroup %>%
#   mutate(Gene = factor(Gene) %>% fct_reorder(Gene, order_by_me, .fun = min))
# levels(cohort_thresholds_for_plot$Gene)


ggplot(cohort_thresholds_for_plot) +
  #geom_tile(aes(x=stat, y= Gene, fill = frac_change_in_ped_relative_to_TCGA)) +
  geom_tile(aes(x=stat, y= fct_reorder(Gene, order_by_me), fill = frac_change_in_ped_relative_to_TCGA)) +
  scale_fill_gradient2("pct change in PEDAYA\nrelative to TCGA")
  
#geom_tile(aes(x=stat, y= fct_reorder(Gene, frac_change_in_ped_relative_to_TCGA), fill = frac_change_in_ped_relative_to_TCGA))
```
# plot boxplots for TCGA and PEDAYA 
```{r}

TP_cohort_expr <- left_join(samples_in_cohorts %>%
                                     filter(cohort %in% c("PEDAYA", "TCGA")),
                                   expr,
                                   by=c("Sample_ID")) 
  
```

# Boxplots of all genes
```{r fig.height=20, fig.width=6}

# TP_cohort_expr_subset <- TP_cohort_expr %>%
#   slice_sample(n = 10000)

ggplot(TP_cohort_expr) +
  geom_boxplot(aes(y=cohort, x=log2TPM1,
                   fill = cohort),
               outlier.shape = NA) +
  facet_wrap(~Gene, ncol = 1,
             strip.position = "left") +
 theme(strip.text.y.left = element_text(angle = 0),
       axis.text.y = element_blank(),
        panel.spacing = unit(0.2, "lines"))  +
  scale_fill_bright() 
  
```

```{r}

table(outliers$comparison_cohort)

TCGA_only_outliers <- outliers %>%
  group_by(gene, Sample_ID) %>%
  mutate(TCGA_only = "TCGA" %in% comparison_cohort & 
           n() == 1) %>%
  filter(TCGA_only) %>%
  arrange(Sample_ID, gene)

n_distinct(TCGA_only_outliers$gene)

```



```{r}

TP_cohort_expr_of_TCGA_only_outliers <- TP_cohort_expr %>%
         filter(Gene %in% TCGA_only_outliers$gene)

ggplot(TP_cohort_expr_of_TCGA_only_outliers) +
  geom_boxplot(aes(y=cohort, x=log2TPM1,
                   fill = cohort),
               outlier.shape = NA) +
  facet_wrap(~Gene, ncol = 1,
             strip.position = "left") +
 theme(strip.text.y.left = element_text(angle = 0),
       axis.text.y = element_blank(),
        panel.spacing = unit(0.2, "lines"))  +
  scale_fill_bright() 

manual_annotation_of_pedaya_relative_to_TCGA <- tribble(
  ~Gene, ~IQR, ~shift,
  "BCL6", "wider", "none",
  "CCND1", "wider", "lower",
  "CCND3", "wider", "higher",
  "CDK9", "wider", "higher",
  "ETV1", "wider", "lower",
  "FGFR1", "wider", "higher",
  "FGFR2", "wider", "lower",
  "FGFR3", "wider", "lower",
  "GATA2", "wider", "none",
  "HMOX1", "wider", "lower",
  "IGF2", "wider", "none",
  "KDR", "wider", "lower",
  "MTOR", "similar", "higher",
  "PARP2", "similar", "higher",
  "PDGFRA", "wider", "lower",
  "PIK3CD", "wider", "higher",
  "PTCH1", "wider", "higher",
  "RPTOR", "similar", "higher",
  "TSC2", "wider", "higher",
  "VEGFA", "wider", "lower") %>%
  filter(Gene %in% TCGA_only_outliers$gene)

# cat(paste0("\"", unique(sort(TCGA_only_outliers$gene)), "\""), sep = ",\n")
  
```

# medians for all druggable genes
```{r}

n_distinct(cohort_thresholds_for_plot$Gene)
  
ggplot(cohort_thresholds_for_plot  %>%
              filter(stat == "median"),
         aes(x=PEDAYA, y=TCGA)) +
  #geom_histogram(aes(x=frac_change_in_ped_relative_to_TCGA))
  geom_point() +
  #geom_label_repel(aes(label = Gene)) +
  geom_abline() +
  coord_equal() +
  #scale_y_continuous(breaks = c(0:3)) +
  expand_limits(y=0,x=0) +
  ggtitle("medians for all druggable genes") +
  scale_color_highcontrast() 


```


# medians for genes that are outliers in TCGA and not Treehouse
```{r}

anno_cohort_thresholds_of_TCGA_only_outliers <-
  left_join(cohort_thresholds_for_plot %>%
              filter(Gene %in% TCGA_only_outliers$gene),
            manual_annotation_of_pedaya_relative_to_TCGA)


  ggplot(anno_cohort_thresholds_of_TCGA_only_outliers  %>%
              filter(stat == "median"),
         aes(x=PEDAYA, y=TCGA, color = shift)) +
  #geom_histogram(aes(x=frac_change_in_ped_relative_to_TCGA))
  geom_point() +
  geom_label_repel(aes(label = Gene)) +
  geom_abline() +
  coord_equal() +
  #scale_y_continuous(breaks = c(0:3)) +
  expand_limits(y=0,x=0) +
  ggtitle("medians for genes that are outliers in TCGA and not Treehouse") +
  scale_color_highcontrast() 

# cohort_thresholds_for_plot %>%
#   filter(Gene %in% TCGA_only_outliers$gene) %>%
#   filter(stat == "median") %>%
#   arrange(frac_change_in_ped_relative_to_TCGA)

```


#  IQRs for all druggable genes
```{r}

ggplot(cohort_thresholds_for_plot  %>%
         filter(stat == "IQR"),
       aes(x=PEDAYA, y=TCGA)) +
  #geom_histogram(aes(x=frac_change_in_ped_relative_to_TCGA))
  geom_point() +
  # geom_text_repel(aes(label = Gene)) +
  geom_abline() +
  coord_equal() +
  scale_y_continuous(breaks = c(0:3)) +
  expand_limits(y=0,x=0) +
  ggtitle("IQRs for genes that are outliers in TCGA and not Treehouse") +
  scale_color_highcontrast() 

```

# IQRs for genes that are outliers in TCGA and not Treehouse
```{r}

ggplot(anno_cohort_thresholds_of_TCGA_only_outliers  %>%
         filter(stat == "IQR"),
       aes(x=PEDAYA, y=TCGA, color = IQR)) +
  #geom_histogram(aes(x=frac_change_in_ped_relative_to_TCGA))
  geom_point() +
  geom_text_repel(aes(label = Gene)) +
  geom_abline() +
  coord_equal() +
  scale_y_continuous(breaks = c(0:3)) +
  expand_limits(y=0,x=0) +
  ggtitle("IQRs for genes that are outliers in TCGA and not Treehouse") +
  scale_color_highcontrast() 

```

# Assign gene groups
```{r}

gene_groups <- tibble(Gene = unique(expr$Gene),
       gene_group = case_when(
         Gene %in% TCGA_only_outliers$gene ~ "outlier_in_TCGA_only",
         Gene %in% outlier_genes_detected ~ "other outlier",
         TRUE ~ "not a CKCC2 outlier")) 

cohort_thresholds_by_gene_group <- left_join(cohort_thresholds_for_plot,
                                        gene_groups, 
                                        by = "Gene")     


cohort_thresholds_by_gene_group_longer <- cohort_thresholds_by_gene_group %>%
  select(-change_in_treehouse_relative_to_TCGA, -frac_change_in_ped_relative_to_TCGA,
         -order_by_me) %>%
  pivot_longer(cols = c("PEDAYA", "TCGA", "TH03_TH34", "Treehouse_pc"),
               names_to = "cohort")

ggplot(cohort_thresholds_by_gene_group_longer %>%
         filter(stat == "median")) +
  geom_histogram(aes(x=value, fill = cohort)) +
  facet_grid(gene_group ~ cohort) +
  scale_fill_bright()


ggplot(cohort_thresholds_by_gene_group_longer %>%
         filter(stat == "median")) +
  geom_histogram(aes(x=value, fill = cohort)) +
  facet_grid(gene_group ~ cohort) +
  scale_fill_bright()
  

```


```{r fig.height=20, fig.width=6}

expr_by_gene_group <- left_join(TP_cohort_expr,
                                gene_groups, 
                                by = "Gene")
ggplot(expr_by_gene_group) +
  geom_boxplot(aes(y=cohort, x=log2TPM1,
                   fill = cohort),
               outlier.shape = NA) +
  facet_wrap(~gene_group + Gene , ncol = 1,
             strip.position = "left") +
 theme(strip.text.y.left = element_text(angle = 0),
       axis.text.y = element_blank(),
        panel.spacing = unit(0.2, "lines"))  +
  scale_fill_bright() 

```

# Figure out how best threshold for excluding genes with very low expressoin
```{r}

cohort_thresholds_by_gene_group %>%
  filter(stat == "q75",
         gene_group == "not a CKCC2 outlier") %>%
  ungroup %>%
  arrange(PEDAYA) %>%
  head()


cohort_thresholds_by_gene_group %>%
  filter(stat == "up_outlier_threshold",
         gene_group == "not a CKCC2 outlier") %>%
  ungroup %>%
  arrange(PEDAYA)  %>%
  head()

min_outlier_threshold <- 2
# up outlier threshold less than two in all cohorts
included_genes_cohort_thresholds_by_gene_group_longer <- cohort_thresholds_by_gene_group_longer %>%
  group_by(Gene) %>%
  mutate(low_expression = all(value < min_outlier_threshold)) %>%
  filter(! low_expression)
  
  
```
```{r}

included_genes_expr_by_gene_group <- left_join(TP_cohort_expr,
                                gene_groups, 
                                by = "Gene") %>%
  filter(Gene %in% included_genes_cohort_thresholds_by_gene_group_longer$Gene)

```

# characterize relative distribution of all genes by gene group
```{r}

median_diff <- 0.25
IQR_diff <- 0.5

dist_shifts_by_gene_group <- cohort_thresholds_by_gene_group %>% 
  filter(Gene %in% included_genes_cohort_thresholds_by_gene_group_longer$Gene) %>%
  mutate(abs_change_in_ped_relative_to_TCGA = PEDAYA-TCGA) %>%
#  filter(stat == "median") %>%
  group_by(gene_group, Gene) %>%
  summarize(median_shift_in_ped_relative_to_TCGA = case_when(
    abs_change_in_ped_relative_to_TCGA[stat == "median"] > median_diff ~ "increase",
    abs_change_in_ped_relative_to_TCGA[stat == "median"] < (-1*median_diff) ~ "decrease",
    TRUE ~ "no shift"),
    IQR_shift_in_ped_relative_to_TCGA = case_when(
    abs_change_in_ped_relative_to_TCGA[stat == "IQR"] > IQR_diff ~ "wider",
    abs_change_in_ped_relative_to_TCGA[stat == "IQR"] < (-1*IQR_diff) ~ "narrower",
    TRUE ~ "similar")
    )

dist_shifts_by_gene_group %>%
  kbl %>%
  kable_styling(full_width = FALSE)
    
tabyl(dist_shifts_by_gene_group, 
      gene_group, median_shift_in_ped_relative_to_TCGA) %>%
  adorn_title() %>%
  kbl %>%
  kable_styling(full_width = FALSE)

tabyl(dist_shifts_by_gene_group, 
      gene_group, IQR_shift_in_ped_relative_to_TCGA) %>%
  adorn_title() %>%
  kbl %>%
  kable_styling(full_width = FALSE)

```


# Review data to  identify a good quantitative IQR cutoff
```{r}

anno_cohort_thresholds_of_TCGA_only_outliers %>%
  mutate(abs_change = PEDAYA-TCGA) %>%
  select(Gene, stat, PEDAYA, TCGA,
         pct_change = frac_change_in_ped_relative_to_TCGA,
         abs_change,
         IQR, shift) %>%
  arrange(abs_change) %>% # change to pct_change to view alternative consideration
  adorn_pct_formatting(,,,pct_change) %>%
  filter(stat == "IQR") %>%
  kbl(digits = c(NA, NA, 1, 1, NA, 2, NA, NA)) %>%
  kable_styling(full_width = F)

# IQR change greater than 0.5 log2TPM1

# percent cutoff doesn't really work, because some relatively small pcts are large in log2tpm1 space,
# e.g. FGFR3, BCL6, wider
  
```




# Review data to see identify a good quantitative shift cutoff
```{r}

anno_cohort_thresholds_of_TCGA_only_outliers %>%
  mutate(abs_change = PEDAYA-TCGA) %>%
  select(Gene, stat, PEDAYA, TCGA,
         pct_change = frac_change_in_ped_relative_to_TCGA,
         abs_change,
         IQR, shift) %>%
  arrange(abs_change) %>%
  adorn_pct_formatting(,,,pct_change) %>%
  filter(stat == "median") %>%
  kbl(digits = c(NA, NA, 1, 1, NA, 2, NA, NA)) %>%
  kable_styling(full_width = F)

# if PEDAYA median is 0.25 higher or lower than TCGA median, the shift is higher or lower, respectively

# percent cutoff doesn't really work, because some relatively small pcts are large in log2tpm1 space,

  
```


# Summary of differences

```{r}
#manual_annotation_of_pedaya_relative_to_TCGA

tabyl(manual_annotation_of_pedaya_relative_to_TCGA, IQR)

tabyl(manual_annotation_of_pedaya_relative_to_TCGA, shift)
```

# Text summary

Each sample in the study was compared to TCGA, PEDAYA, Stanford and the full Treehouse compendium.
Often genes would be identified as outliers with respect to multiple cohorts. 
We investigated the exceptions, genes found as outliers with respect to only one cohort. 
The cohort with the largest number of uniquely detected outliers was TCGA, 
We analyzed the distribution of expression for the outlier genes that were 
identified in any samples by comparison with TCGA and no other cohorts. Because the addition of pediatric samples is the primary difference between TCGA and the other cohorts, we focussed on comparing the PEDAYA cohort and TCGA

Of those 16, 15 had wider distributions (>0.5log2(TPM+1) bigger) in PEDAYA than TCGA, while one was similar. 
For 7, the median was higher (by more than 0.25 log2TPM1) in PEDAYA than TCGA.
For 6, the median was lower (by more than 0.25 log2TPM1) in PEDAYA than TCGA. 
For 3, the medians were similar in PEDAYA and TCGA. 


The Treehouse compendium is 77% TCGA (which is 96% adult). The remaining 23% of the Treehouse compendium is 97% PEDAYA (<=30). In the genes we looked at, the changes in distribution between TCGA and Treehouse compendium were consistent with the effects of adding the distribution of PEDAYA samples. 
The differences in the treehouse compendium from TCGA is mostly due to the addition of PEDAYA samples. 
