---
title: "`r gsub('.Rmd', ' ', gsub('_', ' ', knitr::current_input()))`"
author: "`r Sys.getenv('USER')`"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
 html_document:
    toc: yes
---
  
Version 2023.11.21_11.24.15 - focuses on genes found only relative to TCGA, not relative to any other cohort. Previous versions focused on outliers detected relative to TCGA and not treehouse, irrespective of whatever other cohorts they were outliers in
 
Version 2023.11.29_09.19.33 - focuses on all druggable genes, not only genes that were outliers vs one cohort or another

Version 2024.01.12_09.24.05 - removes redcap; only public data was used to generate cohorts (evidence: there are 2814 samples listed as ped aya in the public data and 2814 in the pedaya cohort used for the tertiary cohorts. there are 3211 v11 samples listed as pedaya in redcap). 

Version 2024.01.16_14.33.16 - calculates % of Stanford cohort in the pediatric cohort


```{r setup, include = FALSE}

library(tidyverse)
library(janitor) 
library(gridExtra)
library(cowplot)
library(ggforce)
library(khroma)
library(ggrepel)
library(kableExtra)

```


# LOAD v11 METADATA

```{r}

kbl_me <- function(x) {
    adorn_pct_formatting(x) %>% 
    kbl(format.args = list(big.mark = ","), digits = c(NA, 0, NA)) %>%
    kable_styling(full_width = F) 
}

```

# load public compendium data
```{r}
v11_metadata <- read_tsv("../input_data/clinical_TumorCompendium_v11_PolyA_2020-04-09.tsv.gz")

v11_metadata_min <- v11_metadata %>%
  select(Sample_ID = th_sampleid, age_at_dx, pedaya, disease)

```

## There is one sample with age < 30 that is not indicated as pedaya, but ignore it since we didn't use it in the cohorts
```{r}

tabyl(v11_metadata$pedaya) %>% kbl_me

v11_metadata %>%
  filter(is.na(pedaya),
         ! is.na(age_at_dx)) %>%
  nrow


v11_metadata_min_updated <- v11_metadata %>%
  select(th_sampleid, age_at_dx, pedaya)%>%
  mutate(pedaya = ifelse(is.na(pedaya) & age_at_dx <30,
                          "Yes, age < 30 years",
                         pedaya))

tabyl(v11_metadata_min_updated$pedaya)  %>% kbl_me
```

# Load cohorts

```{r}
stanford_samples  <- read_tsv("../gather_input_data/comparison_to_non_CARE_cohorts/data/TH03_TH34_rollup.sample_list.txt",
                              col_names = "Sample_ID") %>%
  mutate(cohort = "TH03_TH34")

TCGA_samples  <- read_tsv("../gather_input_data/comparison_to_non_CARE_cohorts/data/TCGA_rollup.sample_list.txt",
                              col_names = "Sample_ID") %>%
  mutate(cohort = "TCGA")

PEDAYA_samples  <- read_tsv("../gather_input_data/comparison_to_non_CARE_cohorts/data/PEDAYA_rollup.sample_list.txt",
                              col_names = "Sample_ID") %>%
  mutate(cohort = "PEDAYA")
```

# Gather cohorts
```{r}

pan_cancer_samples <- v11_metadata_min %>%
  select(Sample_ID) %>%
  mutate(cohort = "Treehouse_pc")

cohorts_in_order <- c("Treehouse_pc", "TCGA", "PEDAYA", "TH03_TH34")

samples_in_cohorts <- bind_rows(
  stanford_samples,
  TCGA_samples,
  PEDAYA_samples,
  pan_cancer_samples) %>%
  left_join(v11_metadata_min) %>%
  mutate(cohort = factor(cohort, levels = cohorts_in_order))


tabyl(samples_in_cohorts,
      cohort)
```
# Look at overlaps in cohorts
```{r}

cohort_membership <- samples_in_cohorts %>%
  mutate(value = TRUE) %>%
  pivot_wider(names_from = "cohort",
              values_fill = FALSE)

nrow(cohort_membership)

#  the Treehouse compendium is 77% TCGA 

tabyl(cohort_membership, TCGA)

# the Treehouse compendium is and 22% pediatric 

tabyl(cohort_membership, PEDAYA)

cohort_membership %>%
  mutate(PEDAYA_and_TCGA = TCGA & PEDAYA) %>%
  tabyl(PEDAYA_and_TCGA)

```

# fix "the pediatric cohort, which includes the Stanford cohort"

i say "However, the pediatric cohort, which includes the Stanford cohort, "; check that the stanford cohort are all pediatric

```{r}
cohort_membership %>%
  filter(TH03_TH34,
         ! PEDAYA)

# change it to, which includes x% of the stanford cohort 
cohort_membership %>%
  filter(TH03_TH34) %>%
  tabyl(PEDAYA)


```

So I can say "However, the pediatric cohort, which includes 97.3% of the Stanford cohort"

But now I realize the argument I am making is about the datasets in the study


```{r}
study_datasets <- read_tsv("../input_data/ckcc2_samples_and_metadata_from_redcap_2023.10.24_13.26.34.tsv")


cohort_and_study_membership <- left_join(study_datasets %>% 
            select(THid) %>%
            mutate(in_study = TRUE),
          cohort_membership,
          by=c("THid" = "Sample_ID")) %>%
  replace_na(list(TH03_TH34 = FALSE,
                  TCGA = FALSE, 
                  PEDAYA = FALSE,
                  Treehouse_pc = FALSE))


cohort_and_study_membership %>%
  filter(in_study,
         ! PEDAYA)


cohort_and_study_membership %>%
  filter(in_study) %>%
  tabyl(PEDAYA)

```

So I can say "However, the pediatric cohort, which includes 82.9% of the study cohort


###  (non-tcga) cohorts, which contain a minimum of X% pediatric datasets. 
```{r}

cohort_membership %>%
  filter(TH03_TH34) %>%
  tabyl(PEDAYA)


cohort_membership %>%
  filter(Treehouse_pc) %>%
  tabyl(PEDAYA)

```


# DISEASES
# diseases in cohorts
```{r}

samples_in_cohorts %>%
  group_by(cohort) %>%
  mutate(cohort_total = n()) %>%
  ungroup %>%
  group_by(disease, cohort) %>%
  summarize(n = n(),
            pct_of_cohort = round(100*n/cohort_total, 2),
            stat_char = paste0(n, " (", pct_of_cohort, "%)")) %>%
  ungroup %>%
  arrange(desc(n)) %>%
  mutate(disease = factor(disease, levels = unique(disease))) %>%
  select(disease, cohort, stat_char)  %>%
  distinct() %>%
    # dplyr::group_by(disease, cohort) %>%
    # dplyr::summarise(n = dplyr::n(), .groups = "drop") %>%
    # dplyr::filter(n > 1L)
  pivot_wider(names_from = cohort, values_from = stat_char, values_fill = "0")


```


```{r}

samples_in_cohorts %>%
  tabyl(disease, cohort) %>%
  arrange(desc(PEDAYA)) %>%
  adorn_title() %>%
  kbl() %>%
  kable_styling(full_width = F)

```

```{r}

ggplot(samples_in_cohorts) +
  geom_bar(aes(y=cohort, fill = pedaya)) +
  scale_fill_bright()

```
```{r}

most_common_ped_diseases <- samples_in_cohorts %>%
  filter(cohort == "PEDAYA") %>%
  tabyl(disease) %>%
  arrange(desc(n)) %>%
  slice_head(n=15)

disease_plot <- samples_in_cohorts %>%
  mutate(common_ped_cancers = ifelse(disease %in% most_common_ped_diseases$disease,
                                     disease,
                                     "other") %>%
           factor() %>%
           fct_relevel("other", after=Inf))
         

ggplot(disease_plot) +
  geom_bar(aes(y=cohort, fill = common_ped_cancers)) 

```

