---
title: "`r gsub('.Rmd', ' ', gsub('_', ' ', knitr::current_input()))`"
author: "`r Sys.getenv('USER')`"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
---
  
# version notes

version 2023.11.06_11.32.08 is for troubleshooting how many values are in Molecular Category
version 2023.11.07_12.48.01 is for reviewing data to continue troubleshooting which values to use in Molecular Category
version 2023.11.08_10.00.39 is for adding "IGF1R", "RAS/RAF/MEK" to Molecular Category

```{r setup, include = FALSE}

library(tidyverse)
library(knitr)
library(janitor)
library(kableExtra)
```


# Load and combine data
```{r}


findings_file <- "../input_data/Supplementary_Tables - eTable8_dl_2023.11.06_11.03.03.tsv"

findings <- read_tsv(findings_file,
                     skip = 1) %>%
  rename(donorID = `Patient ID`)

```


```{r}

latest_sample_metadata_file_name <- "../input_data/ckcc2_samples_and_metadata_from_redcap_2023.10.24_13.26.34.tsv"

sample_metadata <- read_tsv(latest_sample_metadata_file_name,
                            col_types = cols(.default = "c")) %>%
  mutate(dx_group_factor = fct_lump_min(dx_group, 2),
         dx_group = as.character(dx_group_factor))

donor_disease <- sample_metadata %>%
  select(donorID, disease, dx_group, age_at_dx, gender) %>%
  distinct() 

```




```{r}

findings_with_metadata <- left_join(
  findings,
  donor_disease,
  by = "donorID")

```

# define molecular categories to use
```{r}

molecular_categories <- 
  bind_rows(
    tibble(molecular_category_name = c("PI3K/AKT/mTOR", "Cell Cycle", "RTK", "JAK/STAT"),
       molecular_category_type = "predefined_molecular_category"),
tibble(molecular_category_name = c("IGF1R", "RAS/RAF/MEK"),
       molecular_category_type = "other_molecular_category"))
```


# Review 
```{r}


library(forcats)
finding_categories <- findings_with_metadata %>%
  rename(molecular_category_name = `Molecular Category`,
         molecular_abnormality_name = `Molecular Abnormality`) %>%
  select(molecular_category_name, donorID, molecular_abnormality_name) %>%
  left_join(molecular_categories, by = "molecular_category_name") %>%
  replace_na(list(molecular_category_type = "none")) %>%
  mutate(simplified_molecular_category = ifelse(molecular_category_type != "none",
                                      molecular_category_name,
                                      "Other"))

```

# Findings info for each sample
```{r}

finding_categories %>%
  arrange(donorID) %>%
  kbl() %>%
  kable_styling(full_width = F)

```

# findings per molecular_abnormality
```{r}
finding_categories %>%
  group_by(molecular_abnormality_name, simplified_molecular_category,
         molecular_category_type) %>%
  summarize(n_donors=n()) %>%
  arrange(molecular_category_type, simplified_molecular_category) %>%
  kbl() %>%
  kable_styling(full_width = F)

table(finding_categories$simplified_molecular_category)


```

# findings per molecular_category
```{r}
finding_categories %>%
  group_by(simplified_molecular_category,
         molecular_category_type) %>%
  summarize(n_donors=n()) %>%
  arrange(molecular_category_type, desc(n_donors)) %>%
  kbl() %>%
  kable_styling(full_width = F)



```

# How many more do we need to condense to have similar sized groups?
E.g. not to have "Other" be 2x all other categories?
```{r}


# aside from predefined_molecular_category
# IGF1R
# RAS/RAF/MEK (should "MAPK" fall into this category?)
# have 5 or more 
# but JAK/STAT only has 3
   
finding_categories %>%
  tabyl(molecular_category_name) %>%
  arrange(desc(n)) %>%
  mutate(running_total = cumsum(n),
         remaining_findings = sum(n)-running_total) %>%
  # adorn_pct_formatting() %>%
  kbl() %>%
  kable_styling(full_width = F)

```


```{r}
ggplot(finding_categories) +
  geom_bar(aes(x = simplified_molecular_category))
```


## Load gene and disease groups 
```{r}

gene_group_info <- read_tsv("../input_data/gene_group_info.tsv", col_types = cols(.default = "c"))

disease_group_names <- read_tsv(file.path("../input_data/", "disease_group_info.tsv"), col_types = cols(.default = "c"))

```
# expected gene groups
```{r}

gene_group_info %>%
  select(outlier_gene_group) %>% 
  kbl(format.args = list(big.mark = ","), digits = c(NA, NA, 4, 4, 3, 2)) %>%
  kable_styling(full_width = F)

```

```{r}
pre_defined_arms <- c("JAK/STAT","RTK", "Cell Cycle", "PI3K/AKT/mTOR")
pre_defined_arms
```



# View gene set colors
```{r}

gene_group_info %>%
  mutate(gene_group_color = cell_spec(
    gene_group_color, color = "white", bold = T,
    background = gene_group_color
  )) %>%
  kable(escape = F, align = "c") %>%
  kable_styling(c("striped", "condensed"), full_width = F)
  

```

# Session Info
```{r}
sessionInfo()
```

# Appendix

## Full findings table
```{r}

findings %>%
  select(donorID, Drug, `Molecular Abnormality`, `Molecular Category`) %>%
  kbl() %>%
  kable_styling(full_width = F)
```

