---
title: "`r gsub('.Rmd', ' ', gsub('_', ' ', knitr::current_input()))`"
author: "`r Sys.getenv('USER')`"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
---
  
# version notes

version 2023.11.06_11.32.08 is for troubleshooting how many values are in Molecular Category

version 2023.11.07_12.48.01 is for reviewing data to continue troubleshooting which values to use in Molecular Category

version 2023.11.08_10.00.39 is for adding "IGF1R", "RAS/RAF/MEK" to Molecular Category

version 2023.11.08_12.53.45 is for adding "NTRK", "VEGF", "ERBB" to Molecular Category

version 2023.11.08_14.13.38 improves tables

version 2023.11.15_09.54.01 uses updated data from Yvonne and adds disease

```{r setup, include = FALSE}

library(tidyverse)
library(knitr)
library(janitor)
library(kableExtra)
library(ggforce) # for facet_row
```


# latest files
```{r}

findings_file <- "../input_data/Treehouse finding prioritization by patient-eTable8_2023.11.15_09.44.25.txt"

sample_metadata_file <- "../input_data/ckcc2_metadata_from_redcap_2023-11-09_10.34.27.tsv"

```


# Load and combine data
```{r}


findings <- read_tsv(findings_file,
                     skip = 1) %>%
  rename(donorID = `Patient ID`)

```


```{r}



sample_metadata <- read_tsv(sample_metadata_file,
                            col_types = cols(.default = "c")) %>%
  mutate(dx_group_factor = fct_lump_min(dx_group, 2),
         dx_group = as.character(dx_group_factor))

donor_disease <- sample_metadata %>%
  select(donorID, disease, dx_group, age_at_dx, gender) %>%
  distinct() 

```




```{r}

findings_with_metadata <- left_join(
  findings,
  donor_disease,
  by = "donorID")

```

there are `r nrow(findings_with_metadata) ` findings

# define molecular categories to use
```{r}

molecular_categories <- 
  bind_rows(
    tibble(molecular_category_name = c("PI3K/AKT/mTOR", "Cell Cycle", "RTK", "JAK/STAT"),
       molecular_category_type = "predefined_molecular_category"), # pre_defined_arms?
tibble(molecular_category_name = c("IGF1R", "RAS/RAF/MEK", "NTRK", "VEGF", "ERBB"),
       molecular_category_type = "other_molecular_category"))

```


# annotate findings with more categories
```{r}

finding_categories <- findings_with_metadata %>%
  rename(molecular_category_name = `Molecular Category`,
         molecular_abnormality_name = `Molecular Abnormality`) %>%
  select(molecular_category_name, donorID, molecular_abnormality_name) %>%
  left_join(molecular_categories, by = "molecular_category_name") %>%
  replace_na(list(molecular_category_type = "none")) %>%
  mutate(molecular_category_type = factor(molecular_category_type, 
                                             levels = c(
                                               unique(molecular_categories$molecular_category_type),
                                               "none")),
         simplified_molecular_category = ifelse(molecular_category_type != "none",
                                      molecular_category_name,
                                      "Other") %>%
           factor(levels = c(molecular_categories$molecular_category_name,
                                "Other"))) %>%
  group_by(simplified_molecular_category) %>%
  mutate(n_donors_with_vulnerability = n(),
         vulnerability_label = paste0(simplified_molecular_category, " (n=", 
                                      n_donors_with_vulnerability, ")")) %>%
  arrange(molecular_category_type, simplified_molecular_category, donorID) %>%
  ungroup

```

# split findings

```{r}
finding_categories_not_other <- finding_categories %>%
  filter(!simplified_molecular_category == "Other")
  
finding_categories_other <- finding_categories %>%
  filter(simplified_molecular_category == "Other")
```


# Findings in pre_defined and newly grouped arms

```{r}

finding_categories_not_other %>%
  select(donorID, molecular_abnormality_name, molecular_category_name) %>%
  kbl() %>%
  kable_styling(full_width = F) %>% 
  group_rows(index = setNames(rle(as.vector(finding_categories_not_other$molecular_category_type))[[1]],
                              rle(as.vector(finding_categories_not_other$molecular_category_type))[[2]])) %>% 
  group_rows(index = setNames(rle(as.vector(finding_categories_not_other$vulnerability_label))[[1]],
                              rle(as.vector(finding_categories_not_other$vulnerability_label))[[2]]))
```

# Findings in ungrouped arms
```{r}

finding_categories_other %>%
  select(donorID, molecular_abnormality_name, molecular_category_name, simplified_molecular_category) %>%
  kbl() %>%
  kable_styling(full_width = F) %>% 
  group_rows(index = setNames(rle(as.vector(finding_categories_other$molecular_category_type))[[1]],
                              rle(as.vector(finding_categories_other$molecular_category_type))[[2]]))%>% 
  group_rows(index = setNames(rle(as.vector(finding_categories_other$vulnerability_label))[[1]],
                              rle(as.vector(finding_categories_other$vulnerability_label))[[2]]))


```

# How many findings per category?
```{r}

table(finding_categories$simplified_molecular_category)

```


# Prepare dx group data
```{r}

disease_group_names <- read_tsv("../input_data/disease_group_info.tsv", col_types = cols(.default = "c")) %>%
  mutate(order = 1:n())


dx_group_counts <- sample_metadata %>%
  select(donorID, dx_group) %>% 
  distinct() %>%
  left_join(disease_group_names, by=c("dx_group"="original_disease_group_name")) %>%
  group_by(dx_group) %>%
  summarize(dx_group_count = n()) %>%
  ungroup

disease_group_info <- left_join(disease_group_names,
                                dx_group_counts,
                                by=c("original_disease_group_name" = "dx_group")) %>%
  mutate(long_disease_group_name_for_plot_n = 
           paste0(long_disease_group_name_for_plot, " (n=", dx_group_count, ")"),
         long_disease_group_name_for_plot_n = factor(long_disease_group_name_for_plot_n,
                                                     levels = unique(long_disease_group_name_for_plot_n)))
         
  #        ) %>%
  # mutate(across(.fns = as_factor))
                                

disease_group_vector <- set_names(disease_group_info$disease_group_color, disease_group_info$long_disease_group_name_for_plot_n)


```


# Barplot
```{r}

revised_molecular_category_type <- tibble(old_molecular_category_type = 
c("predefined_molecular_category", "other_molecular_category", "none"),
revised_molecular_category_type = factor(c("predefined_molecular_category", "ad_hoc_molecular_category", 
"ad_hoc_molecular_category"), levels = c("predefined_molecular_category", "ad_hoc_molecular_category")))

finding_categories_for_plot <- finding_categories %>%
  arrange(n_donors_with_vulnerability, simplified_molecular_category) %>%
  mutate(simplified_molecular_category = factor(simplified_molecular_category,
                                                   levels = rev(unique(simplified_molecular_category))) %>%
           fct_relevel("Other", after = Inf)) %>%
  left_join(sample_metadata %>%
              select(donorID, disease, dx_group) %>%
              distinct(), 
            by = "donorID") %>%
  left_join(disease_group_info, by=c("dx_group"="original_disease_group_name")) %>%
  left_join(revised_molecular_category_type, 
            by=c("molecular_category_type" = "old_molecular_category_type"))
                                           
  
wrap_labels <- function(x) str_replace_all(x, "/", "\n") %>%
  str_replace(" ", "\n")

intial_cap_no_underscore <- function(x) {
  str_replace_all(x, "_", " ") %>%
    str_to_sentence()
}

ggplot(finding_categories_for_plot) +
  geom_bar(aes(x = simplified_molecular_category,
               fill = long_disease_group_name_for_plot_n)) +
  facet_row(~revised_molecular_category_type,
            scales = "free_x",
            space = "free",
            labeller = labeller(revised_molecular_category_type=intial_cap_no_underscore) )  + 
  theme_minimal() +
    scale_x_discrete(labels = wrap_labels) +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_text(angle = 0,
                                    vjust = 0.5),
        panel.border = element_rect(color = "black", fill = NA, size = 0.5)) +
  scale_fill_manual("Diagnosis group",
                    values = disease_group_vector)
```



# Session Info
```{r}
sessionInfo()
```

# Appendix

## Full findings table
```{r}

findings %>%
  select(donorID, Drug, `Molecular Abnormality`, `Molecular Category`) %>%
  kbl() %>%
  kable_styling(full_width = F)
```

