---
title: "`r gsub('.Rmd', ' ', gsub('_', ' ', knitr::current_input()))`"
author: "`r Sys.getenv('USER')`"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
---
  


```{r setup, include = FALSE}

library(tidyverse)
library(knitr)
library(janitor)
library(kableExtra)
```


# Load and combine data
```{r}


findings_file <- "../input_data/Supplementary_Tables - eTable8_dl_2023.11.06_11.03.03.tsv"

findings <- read_tsv(findings_file,
                     skip = 1) %>%
  rename(donorID = `Patient ID`)
```


```{r}

latest_sample_metadata_file_name <- "../input_data/ckcc2_samples_and_metadata_from_redcap_2023.10.24_13.26.34.tsv"

sample_metadata <- read_tsv(latest_sample_metadata_file_name,
                            col_types = cols(.default = "c")) %>%
  mutate(dx_group_factor = fct_lump_min(dx_group, 2),
         dx_group = as.character(dx_group_factor))

donor_disease <- sample_metadata %>%
  select(donorID, disease, dx_group, age_at_dx, gender) %>%
  distinct() 

```

```{r}

donor_disease %>%
  kbl() %>%
  kable_styling(full_width = F)

donor_disease %>%
  tabyl(disease) %>%
  kbl() %>%
  kable_styling(full_width = F)

donor_disease %>%
  tabyl(dx_group) %>%
  kbl() %>%
  kable_styling(full_width = F)

donor_disease %>%
  tabyl(gender) %>%
  kbl() %>%
  kable_styling(full_width = F)
```



```{r}

findings_with_metadata <- left_join(
  findings,
  donor_disease,
  by = "donorID")
```

```{r}

raw_care_output <- read_tsv(file.path("../data/", latest_care_output_file),
                            col_types = cols(.default = "c")) %>%
  rename(result_type = assay, result=results, overlap_genes = details2) 

sample_metadata <- read_tsv(file.path("../data/", latest_sample_metadata_file_name),
                            col_types = cols(.default = "c")) %>%
  mutate(dx_group_factor = fct_lump_min(dx_group, 2),
         dx_group = as.character(dx_group_factor))

care_with_metadata <- inner_join(raw_care_output, sample_metadata,
                                 by=c("sample_id"="THid"))

# 
# sample_metadata %>% filter(dx_group == "Other")
# candidate_dx_abbrev_other <- c("Other", "ENDO", "Liver")
# sample_metadata$dx_group_factor
```

## QC
```{r}


# confirm that there are no missing or extra samples betweeen
#  sample metadata and care output 
# expected result: character(0)
setdiff(unique(raw_care_output$sample_id), unique(sample_metadata$THid))
setdiff(unique(sample_metadata$THid), unique(raw_care_output$sample_id))

```



## Load gene and disease groups 
```{r}

genesByCategory=read_tsv("../data/treehouseDruggableGenes_2020-03_25.txt", col_types = cols(.default = "c"))

gene_group_info <- read_tsv("../data/gene_group_info.tsv", col_types = cols(.default = "c"))

disease_group_names <- read_tsv(file.path("../data/", "disease_group_info.tsv"), col_types = cols(.default = "c"))


```

# Select only outlier results and add gene and disease groups
```{r}
care_outliers <- care_with_metadata %>%
  filter(result_type == "druggableUpOutlier",
         result != "No druggableUpOutliers")  %>%
  left_join(genesByCategory, by=c("result"="gene")) %>%
  rename(outlier_gene_group = group) %>%
  mutate(outlier_gene_group_factor = fct_lump_min(outlier_gene_group, 2)) %>%
  left_join(gene_group_info, by=c("outlier_gene_group_factor"="outlier_gene_group")) %>%
  mutate(outlier_gene_group_factor = fct_infreq(outlier_gene_group_factor)) %>%
           #fct_relevel("Other", after = Inf)) %>%
  left_join(disease_group_names, by=c("dx_group_factor"="original_disease_group_name")) %>%
  mutate(dx_group_factor = fct_infreq(dx_group_factor) %>%
           fct_relevel("Other", after = Inf))

# 
# care_outliers %>% filter(disease_group_abbrev_for_plot %in% candidate_dx_abbrev_other)
# ids_to_track <- c("TH34_1179_S01", "TH34_1379_S01", "TH34_1444_S01")
# care_outliers %>% filter(sample_id %in% ids_to_track)
```


# Prepare plot data
```{r}
pre_defined_arms <- c("JAK/STAT","RTK", "Cell Cycle", "PI3K/AKT/mTOR")

care_results_for_barplots <- care_outliers %>% 
  arrange(outlier_gene_group_factor) %>%
  mutate(gene_group_x_axis_label = factor(gene_group_x_axis_label, levels =  unique(gene_group_x_axis_label))) %>%
  arrange(dx_group_factor) %>%
#  mutate(disease_group_abbrev_for_plot = factor(disease_group_abbrev_for_plot, levels = unique(disease_group_abbrev_for_plot))) %>%
  mutate(disease_group_abbrev_for_plot = factor(disease_group_abbrev_for_plot, levels = disease_group_names$disease_group_abbrev_for_plot)) %>%
  mutate(study_arm = ifelse(gene_group_abbrev %in% pre_defined_arms, "Predefined arms", "Other") %>%
           factor(levels = c("Predefined arms", "Other")))

# care_results_for_barplots %>% filter(sample_id == "TH34_1379_S01")
```

# Outliers visualized by gene set
```{r}

min_font_size <- 12


barplot_disease_group_color_vec <- disease_group_names$disease_group_color
names(barplot_disease_group_color_vec) <- disease_group_names$disease_group_abbrev_for_plot

p<- ggplot(care_results_for_barplots) + 
    geom_bar(aes(gene_group_x_axis_label, fill= disease_group_abbrev_for_plot)) + 
		theme_minimal()   + 
  # scale_fill_identity("", labels = disease_group_names$disease_group_abbrev_for_plot , breaks = disease_group_names$disease_group_color,  guide = "legend") +
  scale_fill_manual("", values = barplot_disease_group_color_vec,  guide = "legend") +
		theme(
			axis.title.x=element_text(size=min_font_size), 
			axis.title.y=element_text(angle=90, size=min_font_size),
			axis.text.y=element_text(size=min_font_size),
			legend.text=element_text(size=min_font_size),
			legend.title=element_blank()
		) + 
		ylab("Gene expression\noutliers (count)")  + 
  	xlab("Outlier pathway")

p_arm <- p + facet_grid(~study_arm, scales = "free_x", space="free")	+ 
  theme(panel.border = element_rect(size = 0.2, fill = NA))

p_arm # consider rearrangeing so that "other" is consistently at the top or bottom of the bar

ggsave("../results/outliers_by_pathway.pdf", p_arm)

paste("plot contains data about", length(unique(care_results_for_barplots$sample_id)), "samples")

```


```{r}
# tabyl(sample_metadata, dx_group)
# sample_metadata %>% 
#   filter(dx_group %in% c("Liver", "Other"))
```

# Outliers visualized by disease group set
```{r}


p2<- ggplot(care_results_for_barplots ) + 
  	geom_bar(aes(disease_group_abbrev_for_plot, fill= gene_group_color )) + 
		theme_minimal()   + 
    scale_fill_identity("", labels = gene_group_info$gene_group_abbrev, breaks = gene_group_info$gene_group_color) +
		theme(
			axis.title.x=element_text(size=min_font_size), 
			axis.title.y=element_text(angle=90, size=min_font_size),
			axis.text.y=element_text(size=min_font_size),
			legend.text=element_text(size=min_font_size),
			legend.title=element_blank()
		) + 
		ylab("Gene expression\noutliers (count)")  + 
		xlab("Diagnosis group") +
  theme(legend.position="none")

for_first_legend <-
  care_results_for_barplots %>%
  filter(gene_group_abbrev %in% pre_defined_arms) %>%
  ggplot() + 
	geom_bar(aes(disease_group_abbrev_for_plot, fill= gene_group_color )) + 
  scale_fill_identity(name = "Predefined arms", labels = gene_group_info$gene_group_abbrev, breaks = gene_group_info$gene_group_color, 
                      guide = "legend")


for_second_legend <-
  care_results_for_barplots %>%
  filter(!gene_group_abbrev %in% pre_defined_arms) %>%
  ggplot() + 
	geom_bar(aes(disease_group_abbrev_for_plot, fill= gene_group_color )) + 
  scale_fill_identity(name = "Other", labels = gene_group_info$gene_group_abbrev, breaks = gene_group_info$gene_group_color, 
                      guide = "legend")


p3 <- cowplot::plot_grid(
  p2, 
  cowplot::get_legend(for_first_legend),
  cowplot::get_legend(for_second_legend),
  ncol = 3,
  rel_widths = c(6,2,2)
)

p3 # Do we want HSP to be grey? lavender would probably be better.

ggsave("../results/outliers_by_diagnosis_group.pdf", p3)
```



# Samples visualized by treatment decision

## Prepare plot data
```{r warning=FALSE}
library(RColorBrewer)
library(networkD3)
library(colorspace)



# tx_treatments %>% filter(donorID == "TH34_1380")

tx_treatments_raw <- read_tsv( "../data/Treehouse.Drug.Prioritization.No.Identifiers.final.03.07.21 - case_prioritization.tsv", col_types = cols(.default = "c")) 

```

## QC
```{r}


# confirm that there are no missing or extra samples betweeen
#  sample metadata and care output 
# expected result: character(0)
setdiff(unique(tx_treatments_raw$TH_donorID), unique(sample_metadata$donorID))
setdiff(unique(sample_metadata$donorID), unique(tx_treatments_raw$TH_donorID))


```
```{r}

tx_treatments <- tx_treatments_raw %>%
  #attach the disease data 
  left_join(select(sample_metadata, "donorID", "dx_group"),
            by = c("TH_donorID" = "donorID")) %>%
  #left_join can potentially create duplicates so need to make distinct: 
  distinct() %>%
  #renaming for ease of calling: 
  rename(
    donorID = "TH_donorID",
    prior = "Case prioritization",
    prior_reason = "Reason",
    implement = "Implementation reason for prioritized findings"
  ) %>%
  #add the colors based on the dx_group
  mutate(dx_group = recode(dx_group, !!!deframe(
    select(
      disease_group_names,
      original_disease_group_name,
      disease_group_abbrev_for_plot
    )
  )))

# select only relevant info 
sankey_raw_input <- tx_treatments %>%
  select("donorID",
         "prior",
         "prior_reason",
         "implement")

# make the tibble for the links 
sankey_tibble <- bind_rows(
  sankey_raw_input %>% select(donorID, target = prior) %>% mutate(source = donorID),
  sankey_raw_input %>% select(donorID, source = prior, target = prior_reason),
  sankey_raw_input %>% select(donorID, source = prior_reason, target = implement)
) %>%
  na.omit %>%
  mutate(value = 1) %>%
  mutate(source = str_replace(source, "TH34_", "S"))

# make the nodes for the graph 
tx_nodes <-
  tibble(name = c(
    as.character(sankey_tibble$source),
    as.character(sankey_tibble$target)
  ) %>% unique()) %>% 
  #need to mutate enforce logical ordering 
  mutate(
    name = factor(name) %>% fct_relevel(
      #things we want to explicitly sort
      "Prioritized",
      "Deferred",
      "Declined",
      "Implemented",
      "Not implemented",
      #Implemented category: 
      "Response",
      "Progression",
      #sorting by category required to enforce above sort 
      unique(filter(sankey_tibble, source == "Not implemented")$target),
      unique(filter(sankey_tibble, source == "Deferred")$target),
      unique(filter(sankey_tibble, source == "Declined")$target)
    )
  ) %>%
  arrange(name)


tx_disease <- tx_treatments %>%
  mutate(donorID = str_replace(donorID, "TH34_", "S")) %>%
  select(donorID, dx_group)

# prepping the nodes for sankey 
tx_nodes <- tx_nodes %>% 
  mutate(group = recode(name, !!!deframe(tx_disease))) %>% #tag with the dx_group 
  mutate(group = str_replace_all(group, " ", "_")) %>% #sankey doesn't take spaces
  mutate(node_number = 0:(n() - 1)) #number the nodes 


tx_disease <- #need to append link so that link coloring can be seperate  
  tx_disease %>% mutate(donorID = str_replace(donorID, "S", "TH34_")) %>% mutate(dx_group = paste0(dx_group, "_link"))

#highlighting these by disease
sankey_tibble <-
  sankey_tibble %>% # mutate(group = if_else(target %in% highlights, target,"")) # highlight
  mutate(group = recode(donorID, !!!deframe(tx_disease))) # by disease


# With networkD3, connection must be provided using id, not using real name like in the links dataframe. So we need to reformat it.
sankey_tibble$IDsource = tx_nodes$node_number[match(sankey_tibble$source, tx_nodes$name)]
sankey_tibble$IDtarget = tx_nodes$node_number[match(sankey_tibble$target, tx_nodes$name)]
```
## Coloring 
```{r}
# grabbing these to put into color
not_implemented <- c(unlist(unique(filter(sankey_tibble, source == "Not implemented") %>% mutate(target = str_replace_all(target, " ", "_"), .keep = "none"))),"Not_implemented", "Progression")
declined <- c(unlist(unique(filter(sankey_tibble, source == "Declined") %>% mutate(target = str_replace_all(target, " ", "_"), .keep = "none"))),"Declined")
deferred <- c(unlist(unique(filter(sankey_tibble, source == "Deferred") %>% mutate(target = str_replace_all(target, " ", "_"), .keep = "none"))),"Deferred")

# and then doing the color string. note that domain needs line up to range
# non-specific or excess domain will loop around (to the disease colors) which is undesirable 
my_color <- paste0('d3.scaleOrdinal()
    .domain([', #setting the groups to be colored: 
    paste0('"', disease_group_names$disease_group_abbrev_for_plot, '"', collapse=","),",",
    paste0('"', disease_group_names$disease_group_abbrev_for_plot, '_link"', collapse=","),",",
    '"Presented", "Prioritized", "Implemented","Response",',
    #the 3 negative groups
    paste0('"', not_implemented, '"', collapse=","),",",
    paste0('"', deferred, '"', collapse=","),",",
    paste0('"', declined, '"', collapse=","),
    '])\n\t.range([',#now setting the colors: 
   paste0('"', disease_group_names$disease_group_color, '"', collapse=","),",", #disease coloring
   paste0('"', lighten(disease_group_names$disease_group_color,amount=0.65,space = "HLS"), '"', collapse=","),",", # link coloring
   paste0('"', brewer.pal(name="Blues",n=6)[2:5],'"', collapse=","),",", # positive coloring
    paste0('"', heat_hcl(length(not_implemented)+length(deferred)+length(declined),rev=TRUE, c=c(60,20),l=c(60,100),power=1),'"', collapse=","), #negative coloring 
   '])')
```
## Visualization
```{r fig.height=6, fig.width=10}

set.seed(1213444)

# Make the Network
sp <- sankeyNetwork(
  Links = sankey_tibble,
  Nodes = tx_nodes,
  Source = "IDsource",
  Target = "IDtarget",
  Value = "value",
  NodeID = "name",
  sinksRight = FALSE,
  colourScale = my_color,
  nodeWidth = 40,
  fontSize = 13,
  nodePadding = 5,
  iterations = 10, 
  NodeGroup = "group",
  LinkGroup = "group"
)
sp
# 
# # saving 
# if(!webshot::is_phantomjs_installed()){
#   webshot::install_phantomjs(force=TRUE)
# }
# 
# saveNetwork(sp, "../results/treatment_decisions_river_code.html")
# library(webshot)
# webshot("../results/treatment_decisions_river_code.html", "../results/treatment_decisions_river.pdf") 
```





# Support for statements in text

## Patient demographics
The median age at diagnosis was 11 years with a range of 0-24 years. Fifty-five percent of patients were male and forty-five percent were female. The most frequent disease was soft-tissue sarcoma (n=, %).

```{r}

donor_level_info <- sample_metadata %>%
  select(donorID, disease, age_at_dx, gender, dx_group) %>%
  distinct

median(as.numeric(donor_level_info$age_at_dx))
range(as.numeric(donor_level_info$age_at_dx))

tabyl(donor_level_info, gender)

tabyl(donor_level_info, dx_group) %>%
  arrange(desc(n))
```

# Review


## Disease groups

### confirm that all disease groups in data have info in table
```{r}

sample_metadata$dx_group[! sample_metadata$dx_group %in% disease_group_names$original_disease_group_name]
# expected result: character(0)

```

sample_metadata
```{r}
sample_metadata %>%
  select(donorID, race_and_ethnicity) %>%
  distinct()
```



### Review disease group assignments
```{r}
sample_metadata %>% 
  group_by(disease, dx_group) %>%
  summarise(n_samples_with_disease = n()) %>%
  arrange(dx_group, desc(n_samples_with_disease)) %>%
  kable

```



### What disease groups are present?
```{r}

tabyl(sample_metadata, dx_group) %>%
  arrange(n)
# dput(unique(sample_metadata$dx_group))

```


### Review disease group abbreviations
```{r}

disease_group_names  %>%
mutate(disease_group_color = cell_spec(
    disease_group_color, color = "white", bold = T,
    background = disease_group_color
  )) %>%
  kable(escape = F, align = "c") %>%
  kable_styling(c("striped", "condensed"), full_width = F)

```


## Treehouse druggable gene groups


### What gene groups are present multiple times

```{r}

outlier_gene_groups <- care_outliers %>%
  group_by(outlier_gene_group_factor, outlier_gene_group) %>%
  summarize(n_outlier_events_for_group = n()) %>%
  arrange(desc(n_outlier_events_for_group))
  
outlier_gene_groups %>%
  kable


```



### Review outlier category assignments in results

```{r}

druggable_up_outliers_in_results <- care_outliers %>%
  group_by(result, outlier_gene_group) %>%
  summarize(n_outlier_events = n()) %>%
  arrange(outlier_gene_group)

kable(druggable_up_outliers_in_results)
    
```



# View gene set colors
```{r}
care_outliers %>%
  group_by(outlier_gene_group, outlier_gene_group_factor, gene_group_abbrev, gene_group_color) %>%
  summarize(n_occurances = n()) %>%
  arrange(desc(n_occurances)) %>%
mutate(gene_group_color = cell_spec(
    gene_group_color, color = "white", bold = T,
    background = gene_group_color
  )) %>%
  kable(escape = F, align = "c") %>%
  kable_styling(c("striped", "condensed"), full_width = F)
  
  

```

# Session Info
```{r}
sessionInfo()
```

