---
title: "`r gsub('.Rmd', ' ', gsub('_', ' ', knitr::current_input()))`"
author: "`r Sys.getenv('USER')`"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: yes
---
  

verion 2024.01.12_12.58.16 make table of outliers



```{r setup, include = FALSE}

library(tidyverse)
library(janitor)
library(khroma)
library(kableExtra)
library(gridExtra)
library(cowplot)
```


```{r}

underscore_to_space <- function(x) str_replace_all(x, "_", " ")
underscore_to_space_initial_cap <- function(x) str_replace_all(x, "_", " ") %>% str_to_sentence()

```

```{r}

outliers <- read_tsv("../input_data/druggable_outliers_from_treehouse_and_other_cohorts_2023_11_09-13_46_32_2023.tsv") %>%
  mutate(high_level_cohort = ifelse(str_detect(comparison_cohort, "Treehouse"),
                                    "Treehouse",
                                    comparison_cohort))

         
n_distinct(outliers$Sample_ID)       
n_distinct(outliers$donor_ID)
```

# Define cohort codes
```{r}

cohort_codes <- tibble(
  cohort_name = 
    c("PEDAYA", "TCGA", "TH03_TH34", "Treehouse_pc", "Treehouse_pd"),
  cohort_code = 
    c("P", "T", "S", "C", "D"))

```




# outliers detected by treehouse and other cohorts (excludes pan-disease)
```{r}

treehouse_pc_outliers <- outliers %>%
  filter(comparison_cohort == "Treehouse_pc") %>%
  select(Sample_ID, gene) %>%
  distinct()

outliers_detected_by_treehouse_pc <- left_join(treehouse_pc_outliers, 
                                            outliers %>%
                                              filter(comparison_cohort != "Treehouse_pd"), 
                                            by = c("Sample_ID", "gene"))


treehouse_pc_outlier_summary <- outliers_detected_by_treehouse_pc %>% 
  group_by(comparison_cohort) %>%
  summarize(n_outliers_detected = n())  %>%
  mutate(total = nrow(treehouse_pc_outliers),
         set = "detected by treehouse pc and other cohorts (excludes pan-disease)")



```




# outliers detected by treehouse and other cohorts (includes pan-disease)


```{r}

treehouse_outliers <- outliers %>%
  filter(str_detect(high_level_cohort, "Treehouse")) %>%
  select(Sample_ID, gene) %>%
  distinct()

outliers_detected_by_treehouse <- left_join(treehouse_outliers, 
                                            outliers, 
                                            by = c("Sample_ID", "gene"))
n_outliers_detected_by_treehouse <- treehouse_outliers %>%
  nrow()

treehouse_outlier_summary <- outliers_detected_by_treehouse %>% 
  group_by(comparison_cohort) %>%
  summarize(n_outliers_detected = n()) %>%
  bind_rows(tibble(comparison_cohort = "Treehouse total",
                 n_outliers_detected = n_outliers_detected_by_treehouse)) %>%
  mutate(total = n_outliers_detected_by_treehouse,
         set = "detected by treehouse and other cohorts (includes pan-disease)")

```

# outliers detected by any cohort
```{r}

n_outliers_detected_by_any_method <- outliers %>%
  select(Sample_ID, gene) %>%
  distinct %>%
  nrow()

outlier_summary <- outliers %>% 
  group_by(comparison_cohort) %>%
  summarize(n_outliers_detected = n())  %>%
  bind_rows(tibble(comparison_cohort = "Treehouse total",
                 n_outliers_detected = n_outliers_detected_by_treehouse)) %>%
  mutate(total = n_outliers_detected_by_any_method,
         set = "detected by treehouse OR other cohorts (includes pan-disease)")

```


```{r}
all_outlier_long_table <- bind_rows(outlier_summary,
          treehouse_outlier_summary,
          treehouse_pc_outlier_summary)  %>%
  mutate(stat = paste0(n_outliers_detected, "/", total, " (",
             round(100*n_outliers_detected/total, 1), "%)"))
```


```{r}

names_of_sets <- tibble(
  long_set_name = c("detected by treehouse OR other cohorts (includes pan-disease)", 
"detected by treehouse and other cohorts (includes pan-disease)", 
"detected by treehouse pc and other cohorts (excludes pan-disease)"),
short_set_name = c("all_methods", "treehouse_with_pd", "treehouse_without_pd"))

all_outlier_long_table %>%
  left_join(names_of_sets, 
            by=c("set" = "long_set_name")) %>%
  select(comparison_cohort, stat, short_set_name) %>%
  pivot_wider(names_from = short_set_name,
              values_from = stat)

```

# Sample-specific comparator cohorts provide x% additional findings undetectable by predefined cohorts



## keeping TCGA outliers
```{r}

treehouse_pd_anno_outliers <- outliers %>%
  group_by(Sample_ID, gene) %>%
  summarize(is_Treehouse_pd = any(comparison_cohort == "Treehouse_pd"),
            n_outlier_comparison_cohorts = n())

treehouse_pd_only_outliers <- treehouse_pd_anno_outliers %>%
  filter(is_Treehouse_pd, n_outlier_comparison_cohorts ==1) %>%
  nrow

#13 additional findings  from Sample-specific comparator cohorts 

```

```{r}

predefined_pediatric_inclusive_cohorts <- c("PEDAYA", "TH03_TH34", "Treehouse_pd", "Treehouse_pc")
predefined_cohorts <- c("PEDAYA", "TH03_TH34", "TCGA", "Treehouse_pc")

 outliers %>%
  group_by(Sample_ID, gene) %>%
  summarize(in_predefined_cohort = any(comparison_cohort %in% predefined_cohorts)) %>%
  filter(in_predefined_cohort) %>%
   nrow()
 

# findings of predefined pediatric-inclusive cohorts
# treehouse_pd_only_outliers <- outliers %>%
#   group_by(Sample_ID, gene) %>%
#   summarize(is_Treehouse_pd = any(comparison_cohort == "Treehouse_pd"),
#             n_outlier_comparison_cohorts = n())


```

117 findings detected with predefined cohorts

Sample-specific comparator cohorts provide 13 additional findings undetectable by predefined cohorts
Using sample-specific comparator cohorts increases findings by 11% compared to predefined cohorts 

## removing TCGA outliers
Sample-specific comparator cohorts provide X% additional findings undetectable by predefined pediatric-inclusive cohorts
```{r}

non_TCGA_outliers <- outliers %>%
  filter(comparison_cohort != "TCGA")
  
non_TCGA_treehouse_pd_only_outliers <- non_TCGA_outliers %>%
  group_by(Sample_ID, gene) %>%
  summarize(is_Treehouse_pd = any(comparison_cohort == "Treehouse_pd"),
            n_outlier_comparison_cohorts = n())

non_TCGA_treehouse_pd_only_outliers %>%
  filter(is_Treehouse_pd, n_outlier_comparison_cohorts ==1) %>%
  nrow

# 16 additional findings  from Sample-specific comparator cohorts 

```

```{r}

predefined_pediatric_inclusive_cohorts <- c("PEDAYA", "TH03_TH34", "Treehouse_pd", "Treehouse_pc")
predefined_cohorts <- c("PEDAYA", "TH03_TH34", "TCGA", "Treehouse_pc")

 non_TCGA_outliers %>%
  group_by(Sample_ID, gene) %>%
  summarize(in_predefined_cohort = any(comparison_cohort %in% predefined_cohorts)) %>%
  filter(in_predefined_cohort) %>%
   nrow()
 

# findings of predefined pediatric-inclusive cohorts
# treehouse_pd_only_outliers <- outliers %>%
#   group_by(Sample_ID, gene) %>%
#   summarize(is_Treehouse_pd = any(comparison_cohort == "Treehouse_pd"),
#             n_outlier_comparison_cohorts = n())


```

93 findings detected with predefined cohorts

Sample-specific comparator cohorts provide 16 additional findings undetectable by predefined predefined pediatric-inclusive cohorts cohorts
16/93

Using sample-specific comparator cohorts increases findings by 17% compared to predefined pediatric-inclusive cohorts 



## removing TCGA-ONLY outliers
Sample-specific comparator cohorts provide X% additional findings undetectable by predefined pediatric-inclusive cohorts
```{r}

outliers_excluding_TCGA_only <- outliers %>%
  group_by(Sample_ID, gene) %>%
  summarize(is_TCGA = any(comparison_cohort == "TCGA"),
            n_outlier_comparison_cohorts = n()) %>%
  ungroup %>%
  mutate(TCGA_only = is_TCGA & n_outlier_comparison_cohorts ==1) %>%
  filter(! TCGA_only)

nrow(outliers_excluding_TCGA_only)


# 109 findings from predefined comparator cohorts 
# still 13 findings from pd only
109-13
13/109
13/96

```

We show that comparing a dataset to dynamically generated, patient-specific cohorts generates orthogonal results. 
Comparing to multiple cohorts increased the number of outliers by 11%. 
(
```{r}
# pan-disease outliers


treehouse_pd_anno_outliers %>%
  filter(is_Treehouse_pd) %>%
  mutate(pd_only = n_outlier_comparison_cohorts == 1) %>%
  tabyl(pd_only) %>%
  adorn_totals()
```
Of the 38 pan-disease findings, 34.2% were not detected by any predefined-cohort analysis



```{r}
treehouse_pd_anno_outliers %>%
  filter(is_Treehouse_pd) %>%
  mutate(pd_only = n_outlier_comparison_cohorts == 1) %>%
  tabyl(pd_only) %>%
  adorn_totals()


treehouse_pd_pc_anno_outliers <- outliers %>%
  group_by(Sample_ID, gene) %>%
  summarize(is_Treehouse_pd = any(comparison_cohort == "Treehouse_pd"),
            is_Treehouse_pc = any(comparison_cohort == "Treehouse_pc"),
            is_all_Treehouse = all(high_level_cohort == "Treehouse"),
            n_outlier_comparison_cohorts = n())

treehouse_pd_pc_anno_outliers  %>%
  filter(is_Treehouse_pd) %>%
  tabyl(is_Treehouse_pc) %>%
  adorn_totals()

treehouse_pd_only_outliers <- treehouse_pd_anno_outliers %>%
  filter(is_Treehouse_pd, n_outlier_comparison_cohorts ==1) %>%
  nrow

```


```{r}

```

