---
title: "`r gsub('.Rmd', ' ', gsub('_', ' ', knitr::current_input()))`"
author: "`r Sys.getenv('USER')`"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
 html_document:
    toc: yes
---
  


```{r setup, include = FALSE}

library(tidyverse)
library(janitor) 
library(gridExtra)
library(cowplot)
library(ggforce)
library(khroma)
library(ggrepel)
library(kableExtra)

```


# LOAD AND UPDATE v11 METADATA

```{r}

kbl_me <- function(x) {
    adorn_pct_formatting(x) %>% 
    kbl(format.args = list(big.mark = ","), digits = c(NA, 0, NA)) %>%
    kable_styling(full_width = F) 
}

```

# load public compendium data
```{r}

v11_metadata <- read_tsv("https://xena.treehouse.gi.ucsc.edu/download/clinical_TumorCompendium_v11_PolyA_2020-04-09.tsv.gz")

tabyl(v11_metadata$pedaya) %>% kbl_me

v11_metadata %>%
  filter(is.na(pedaya),
         ! is.na(age_at_dx)) %>%
  nrow

v11_metadata_min <- v11_metadata %>%
  select(th_sampleid, age_at_dx, pedaya) %>%
  mutate(pedaya = ifelse(is.na(pedaya) & age_at_dx <30,
                          "Yes, age < 30 years",
                         pedaya))

tabyl(v11_metadata_min$pedaya)  %>% kbl_me
```
# load redcap data
```{r}

redcap_metadata_v11 <- read_tsv("/Users/hbeale/Documents/Dropbox/ucsc/operations/redcap/redcap_combined_nonprivate_sample_donor_data_downloaded_via_API.v10.2023-11-09_09.33.56.txt.gz") %>%
  filter(THid %in% v11_metadata$th_sampleid)

# redcap_metadata_v11 %>%
#   pull(age_at_dx) %>%
#   summary()


redcap_metadata_v11 %>%
  tabyl(pedaya)   %>% 
  kbl_me

redcap_metadata_v11 %>%
  filter(is.na(pedaya),
         ! is.na(age_at_dx)) %>%
  nrow

redcap_metadata_v11_min <- redcap_metadata_v11 %>%
  select(THid, pedaya, age_at_dx) %>%
  mutate(pedaya = ifelse(is.na(pedaya) & age_at_dx <30,
                         "Yes, age < 30 years",
                                pedaya))
redcap_metadata_v11_min %>%
  tabyl(pedaya)  %>% kbl_me


```

# update v11 compendium metadata from redcap
```{r}

combined_metadata <- left_join(v11_metadata_min, 
                               redcap_metadata_v11_min,
                               by=c("th_sampleid"="THid"),
                               suffix = c("_compendium", "_redcap")) 

combined_metadata %>%
  filter(pedaya_compendium != pedaya_redcap) %>%
  kbl() %>%
  kable_styling()

combined_metadata %>%
  filter(is.na(pedaya_compendium),
         !is.na(pedaya_redcap))  %>%
  nrow

combined_metadata %>%
  filter(is.na(pedaya_compendium),
         !is.na(pedaya_redcap))  %>%
  head %>%
  kbl() %>%
  kable_styling()

combined_metadata %>%
  filter(is.na(pedaya_redcap),
         !is.na(pedaya_compendium)) %>%
  kbl() %>%
  kable_styling()


use_redcap <- c("TARGET-40-0A4I4O-01A-01R",
                combined_metadata %>%
                  filter(is.na(pedaya_compendium),
                         !is.na(pedaya_redcap))  %>%
                  pull(th_sampleid))

use_compendium_data <- c("TH27_2347_S02") # and everything else not from redcap

age_updated_v11 <- bind_rows(
  redcap_metadata_v11_min %>%
    filter(THid %in% use_redcap) %>%
    rename(th_sampleid = THid),
  v11_metadata_min %>%
    filter(!th_sampleid %in% use_redcap)) %>%
  rename(Sample_ID = th_sampleid)


```


# Update disease
```{r}

updated_v11 <- left_join(
  age_updated_v11,
redcap_metadata_v11 %>%
  select(THid, disease),
by=c("Sample_ID" = "THid" ))

```


# COMPARE DISTRIBUTIONS FOR  OUTLIERS ACROSS COHORTS

```{r}

stanford_samples  <- read_tsv("../gather_input_data/comparison_to_non_CARE_cohorts/data/TH03_TH34_rollup.sample_list.txt",
                              col_names = "Sample_ID") %>%
  mutate(cohort = "TH03_TH34")

TCGA_samples  <- read_tsv("../gather_input_data/comparison_to_non_CARE_cohorts/data/TCGA_rollup.sample_list.txt",
                              col_names = "Sample_ID") %>%
  mutate(cohort = "TCGA")

PEDAYA_samples  <- read_tsv("../gather_input_data/comparison_to_non_CARE_cohorts/data/PEDAYA_rollup.sample_list.txt",
                              col_names = "Sample_ID") %>%
  mutate(cohort = "PEDAYA")

pan_cancer_samples <- updated_v11 %>%
  select(Sample_ID) %>%
  distinct() %>%
  mutate(cohort = "Treehouse_pc")

cohorts_in_order <- c("Treehouse_pc", "TCGA", "PEDAYA", "TH03_TH34")


samples_in_cohorts <- bind_rows(
  stanford_samples,
  TCGA_samples,
  PEDAYA_samples,
  pan_cancer_samples) %>%
  left_join(updated_v11) %>%
  mutate(cohort = factor(cohort, levels = cohorts_in_order))


tabyl(samples_in_cohorts,
      cohort)


```


# PATIENT SPECIFIC ANALYSIS

```{r}
this_sample <- "TH34_1455_S01"
this_gene <- "FGFR3"
```

```{r}

updated_v11 %>%
  filter(Sample_ID == this_sample)

```

# confirm outlier status

```{r}
outliers <- read_tsv("../input_data/druggable_outliers_from_treehouse_and_other_cohorts_2023_11_09-13_46_32_2023.tsv") %>%
  mutate(high_level_cohort = ifelse(str_detect(comparison_cohort, "Treehouse"),
                                    "Treehouse",
                                    comparison_cohort))


outliers %>%
  filter(Sample_ID == this_sample, 
         gene == this_gene)

```

# review disease abundance in each cohort
```{r}
underscore_to_space_initial_cap <- function(x) str_replace_all(x, "_", " ") %>% str_to_sentence()

pd_cohorts <- read_tsv("../input_data/pan_disease_cohort_members_2023_11_30-16_37_31_2023.tsv") %>%
  rename(original_cohort_name = cohort,
         focus_sample_ID = TH_id,
         Sample_ID = cohort_member) %>%
  mutate(
         cohort_pd_subset = str_replace(original_cohort_name,
                              "first_degree_mcs_cohort", "pan_disease_1st_degree") %>%
           str_replace("first_and_second_degree_mcs_cohort", "pan_disease_1st_and_2nd_degree") %>%
           str_replace("diagnosed_disease_cohort", "pan_disease_same_diagnosis") %>%
           str_replace("pandisease_samples", "pan_disease_same_inferred_diagnosis"),
         cohort = paste(focus_sample_ID, cohort_pd_subset))

this_focus_sample_pd_cohorts <- pd_cohorts %>%
  filter(focus_sample_ID == this_sample) %>%
  rename(cohort_sample = Sample_ID) %>%
  left_join(updated_v11,
            by=c("cohort_sample" = "Sample_ID"))
```


##  diseases in all pd cohorts
```{r}
this_focus_sample_pd_cohorts %>%
  tabyl(disease, cohort_pd_subset) %>%
  rename_all(underscore_to_space_initial_cap) %>%
  kbl() %>%
  kable_styling(full_width = F)
```

## pan_disease_same_diagnosis
```{r}
this_focus_sample_pd_cohorts %>%
  tabyl(disease, cohort_pd_subset) %>%
  filter(pan_disease_same_diagnosis > 0) %>%
  rename_all(underscore_to_space_initial_cap) %>%
  kbl() %>%
  kable_styling(full_width = F)
```


## pan_disease_same_inferred_diagnosis

```{r}
this_focus_sample_pd_cohorts %>%
  tabyl(disease, cohort_pd_subset) %>%
  filter(pan_disease_same_inferred_diagnosis > 0) %>%
  rename_all(underscore_to_space_initial_cap) %>%
  kbl() %>%
  kable_styling(full_width = F)

```


# review ages in each cohort
```{r}
ggplot(this_focus_sample_pd_cohorts) +
  geom_boxplot(aes(x=age_at_dx, y=cohort_pd_subset))
  
ggplot(this_focus_sample_pd_cohorts) +
  geom_histogram(aes(x=age_at_dx)) +
  facet_wrap(~cohort_pd_subset,
             ncol= 1, scales = "free_y")
```

