---
title: "`r gsub('.Rmd', ' ', gsub('_', ' ', knitr::current_input()))`"
author: "`r Sys.getenv('USER')`"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: yes
---
  
file name: plot distributions for FGFR3 in TH34_1455_S01 in TCGA ped and Treehouse.Rmd
5/14 I'm using the code from "plot distributions for TCGA only genes.Rmd" and not removing exploratory code


```{r setup, include = FALSE}

library(tidyverse)
library(ggforce)

```


# Load outlier results
```{r}

outlier_file_with_expression <- 
  "../input_data/druggable_outliers_from_treehouse_and_other_cohorts_with_expression_2024_05_14-15_23_49_2024.tsv"


outliers <- read_tsv(outlier_file_with_expression)

```

```{r}
samples_to_exclude <- c("TH34_2351_S01", "TH34_2292_S01") # TH34_2351_S01 is the other example; TH34_2292_S01 failed QC
genes_to_exclude <- "HMOX1"  # HMOX1 is used in another example
```


# Candidate example samples
```{r}

candidate_example_samples <- outliers %>%
  filter(! Sample_ID %in% samples_to_exclude) %>% 
  select(Sample_ID) %>%
  distinct()


```



# Identify outliers present in TCGA but not other cohorts

```{r}


TCGA_only_outliers <- outliers %>%
  group_by(gene, Sample_ID) %>%
  mutate(TCGA_only = "TCGA" %in% comparison_cohort & 
           n() == 1) %>%
  filter(TCGA_only,
         Sample_ID %in% candidate_example_samples$Sample_ID,
         ! gene == genes_to_exclude) %>%
  arrange(Sample_ID, gene)


```

## Genes among TCGA_only_outliers
```{r}

table(TCGA_only_outliers$gene)



```


#  load expression for samples in v11 compendium
```{r}


v11_expr <- read_tsv("../input_data/druggable_plus_TumorCompendium_v11_PolyA_hugo_log2tpm_58581genes_2020-04-09.tsv.gz") %>%
  rename(Sample_ID = TH_id) %>%
  filter(Gene %in% TCGA_only_outliers$gene)

```

# Load lists of samples in TCGA and pedaya cohorts
```{r}

TCGA_samples  <- read_tsv("../gather_input_data/comparison_to_non_CARE_cohorts/data/TCGA_rollup.sample_list.txt",
                              col_names = "Sample_ID",
                          show_col_types = FALSE) %>%
  mutate(cohort = "TCGA")

PEDAYA_samples  <- read_tsv("../gather_input_data/comparison_to_non_CARE_cohorts/data/PEDAYA_rollup.sample_list.txt",
                              col_names = "Sample_ID",
                            show_col_types = FALSE) %>%
  mutate(cohort = "PEDAYA")


pan_cancer_samples <- v11_expr %>%
  select(Sample_ID) %>%
  distinct() %>%
  mutate(cohort = "Treehouse_pc")

samples_in_cohorts <- bind_rows(
  TCGA_samples,
  PEDAYA_samples,
  pan_cancer_samples
  )

```

## confirm that all samples in cohorts have expression data
```{r}

samples_in_cohorts %>%
  filter(!Sample_ID %in% v11_expr$Sample_ID)

```


# Calculate statistics for each cohort
```{r}

cohort_thresholds_raw <- left_join(samples_in_cohorts,
                                   v11_expr,
                                   by=c("Sample_ID")) %>%
  group_by(Gene, cohort) %>%
  summarize(q25 = quantile(log2TPM1, 0.25),
            median = median(log2TPM1),
            q75 = quantile(log2TPM1, 0.75),
            IQR = q75-q25,
            up_outlier_threshold = q75 + (1.5*IQR))
  
```

# assess changes for all genes
```{r}
cohort_thresholds <- cohort_thresholds_raw %>%
  pivot_longer(c(-Gene, -cohort)) %>%
  pivot_wider(names_from = cohort, values_from = value) %>%
  mutate(increase_in_ped_relative_to_TCGA = 
              (PEDAYA - TCGA) / TCGA)



```



# changes plotted
```{r}
ggplot(cohort_thresholds) +
  geom_point(aes(x=TCGA,
                 y=increase_in_ped_relative_to_TCGA,
                 color = name)) +
  geom_hline(yintercept = 0) +
  scale_color_brewer(palette = "Set1") +
  ggtitle("increase in various statistics in the pediatric cohort relative to the TCGA cohort")
```

```{r}

cohort_thresholds %>%
  filter(name %in% c("median", "IQR")) %>%
  select(Gene, name, increase_in_ped_relative_to_TCGA) %>%
  pivot_wider(names_from = name,
              values_from = increase_in_ped_relative_to_TCGA) %>%
  ggplot + 
  geom_point(aes(x=median, y=IQR)) + 
  geom_vline(xintercept = 0) +
  expand_limits(y=0) +
  ggtitle("The IQR usually increased irrespective of the direction of change of the median", "fraction increase_in_ped_relative_to_TCGA")


```

```{r}

cohort_thresholds %>%
  filter(name %in% c("median", "IQR"))

iqrs <- cohort_thresholds %>%
  filter(name=="IQR") %>%
  rename(PEDAYA_IQR = "PEDAYA",
         TCGA_IQR = "TCGA",
         increase_in_ped_IQR_relative_to_TCGA = "increase_in_ped_relative_to_TCGA") %>%
  select(-name)


medians <- cohort_thresholds %>%
  filter(name=="median") %>%
  rename(PEDAYA_median = "PEDAYA",
         TCGA_median = "TCGA",
         increase_in_ped_median_relative_to_TCGA = "increase_in_ped_relative_to_TCGA") %>%
  select(-name)

table_of_compared_stats <- left_join(iqrs, medians) %>%
  ungroup() %>%
  mutate(across(where(is.double), round, 2))

```


```{r}

file_output_name <- paste0("../results/comparison_of_IQR_and_median_in_TCGA_vs_PedAYA_distributions", format(Sys.time(), "%Y_%m_%d-%H_%M_%S_%Y"), ".tsv")

#write_tsv(table_of_compared_stats, file_output_name)

```

this output was saved as `r file_output_name`

# Find example
```{r}

potential_example_genes <- cohort_thresholds %>%
  filter(Gene %in% TCGA_only_outliers$gene) %>% 
  group_by(Gene) %>%
  mutate(lower_med_in_pedaya_than_TCGA = PEDAYA[name == "median"] < TCGA[name == "median"]) %>%
  rename(increase = increase_in_ped_relative_to_TCGA) 

potential_example_genes

potential_example_genes %>%
  filter(lower_med_in_pedaya_than_TCGA,
         name == "up_outlier_threshold")

potential_example_genes %>%
  filter(name == "up_outlier_threshold")


```


# Other genes considered
```{r}

table_of_compared_stats %>%
  filter(Gene == "CCND3")

cohort_thresholds  %>%
  filter(Gene == "CCND3") #this isn't a good example because CCND3 is dramatic and has a higher median in ped than tcga


table_of_compared_stats %>%
  filter(Gene == "PDGFRA")

cohort_thresholds  %>%
  filter(Gene == "PDGFRA")  #this isn't a good esxample beause it's is dramatic and has a higher median in ped than tcga


  
```

```{r}
calculate_outlier_threshold <- function(x) {
  # x <- 1:100
  q75 <- quantile(x, 0.75)
  iqr <- q75 - quantile(x, 0.25)
  threshold <- q75 + 1.5*iqr
  return(threshold)
}

```

# Rename cohorts here
```{r}

update_cohort_strings <- function(x) {
  str_replace(x,
              "PEDAYA", 
              "Pediatric") %>%
    str_replace("TH03_TH34",
                "Stanford")%>%
    str_replace("Treehouse_pc",
                "Treehouse pan-cancer")%>%
    str_replace("pan_disease",
                "Treehouse pan-disease,") %>%
    str_replace("1st_degree", "1st degree neighbors") %>%
    str_replace("1st_and_2nd_degree", "1st and 2nd degree neighbors") %>%
    str_replace("same_diagnosis", "same disease as focus sample") %>%
    str_replace("same_inferred_diagnosis", "diseases of top 6 most correlated datasets") %>%
    str_replace_all("_([a-z12][a-z])", " \\1")
}

        
        
```

# plot one gene and all samples for which it is an outlier in TCGA
```{r}

# plot_distributions <- function(this_gene, this_sample_id, outlier_relative_to) {
plot_distributions <- function(this_gene) {
  # this_sample_id <- "TH34_2351_S01"
  #this_gene <- "GATA2"
  
  # print(this_gene)
  
  #print(this_sample_id)
  
  possible_levels <- c("Treehouse_pc", "TCGA", "PEDAYA", "TH03_TH34") %>%
    update_cohort_strings
  
  expr_relevant_to_this_outlier_raw <-  left_join(
    samples_in_cohorts,
    v11_expr %>%
      filter(Gene == this_gene),
    by=c("Sample_ID")) %>%
    mutate(cohort = update_cohort_strings(cohort))
  
  
  n_distinct(samples_in_cohorts$cohort)
  
  n_distinct(expr_relevant_to_this_outlier_raw$cohort)
  
  expected_levels <- possible_levels[possible_levels %in% expr_relevant_to_this_outlier_raw$cohort]
  
  expr_relevant_to_this_outlier <- expr_relevant_to_this_outlier_raw %>%
    # na.omit() %>%  # while I'm missing expression data for some samples
    mutate(cohort = 
             fct_relevel(cohort, expected_levels)) %>%
    arrange(cohort) %>%
    group_by(cohort) %>%
    mutate(cohort_with_n = paste0(cohort, " (n=", scales::label_comma()(n()), ")")) %>%
    ungroup %>%
    mutate(cohort_with_n = factor(cohort_with_n, 
                                  levels = unique(cohort_with_n))) 
  
  
  outlier_thresholds <- 
    expr_relevant_to_this_outlier %>%
    group_by(cohort_with_n) %>%
    summarize(
      outlier_threshold = calculate_outlier_threshold(log2TPM1)) %>%
    ungroup()
  
  
  p <- ggplot(expr_relevant_to_this_outlier) +
    geom_boxplot(aes(y=cohort_with_n, x=log2TPM1),
                 #fill = cohort_with_n),
                 outlier.shape = NA) +
    geom_rect(data = outlier_thresholds,
              aes(xmin = outlier_threshold,
                  xmax = Inf,
                  ymin = -Inf,
                  ymax = Inf),
              fill = "yellow",
              alpha = 0.7) +
    geom_vline(data = TCGA_only_outliers %>%
                 filter(gene == this_gene),
               aes(xintercept = log2TPM1,
                   #lty = patient_is_outlier
               ), 
               color = "red") +
    facet_col(~cohort_with_n, scales = "free_y") +
    theme_minimal() +
    xlab(paste0(this_gene," expression (log2(TPM+1))")) +
    theme(axis.text.y = element_blank(),
          axis.title.y = element_blank(),
          legend.position="none",
          panel.border = element_rect(colour = NA, fill=NA, size=1),
          strip.background = element_rect(color = "lightgrey", size = 1, fill = "lightgrey"),
          panel.grid.major.y = element_blank()) +         #remove horizontal grid line
    ggtitle(paste0(this_gene, " cohorts and expression of ", TCGA_only_outliers %>%
                 filter(gene == this_gene) %>%
                   pull(Sample_ID) %>% paste(collapse = " and ")),
            "red line is patient expression (dashed if outlier); yellow area is above the outlier threshold")  
  
  
  # print(p)
  
return(p)
}
```

# show all plots
```{r}


# lapply(TCGA_only_outliers$gene, plot_distributions)

```


```{r}

this_plot <- lapply("FGFR3", plot_distributions)

this_plot[[1]]

```


```{r}

plot_output_name <- paste0("../results/fig6b_FGFR3_TH34_1455_S01_", format(Sys.time(), "%Y_%m_%d-%H_%M_%S_%Y"), ".png")

ggsave(plot = this_plot[[1]], 
       filename = plot_output_name, 
       bg = "white",
       width = 6,
       height = 3)

```

this output was saved as `r plot_output_name`
