---
title: "`r gsub('.Rmd', ' ', gsub('_', ' ', knitr::current_input()))`"
author: "`r Sys.getenv('USER')`"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: yes
---
  
Updates:  
File name: review expression distributions in different comparison_cohorts_2024.05.09_14.47.48 
Updates/contents: Figure 6, change to BCL in TH34_2292_S01, change formatting. B contains only TCGA, Pediatric, and 
Treehouse pan-caner. C contains those, plus standford and 4 pan-disease cohorts for HMOX1 and TH34_2351_S01

oops, this code doesn't include pan disease


```{r setup, include = FALSE}

library(tidyverse)
library(janitor) 
library(gridExtra)
library(cowplot)
library(ggforce)

```



```{r}

outliers <- read_tsv("../input_data/druggable_outliers_from_treehouse_and_other_cohorts_2023_11_09-13_46_32_2023.tsv") %>%
  mutate(high_level_cohort = ifelse(str_detect(comparison_cohort, "Treehouse"),
                                    "Treehouse",
                                    comparison_cohort))


focus_sample_gene_info_to_keep <- tribble(
  ~Sample_ID, ~gene,
  "TH34_2292_S01", "BCL6",
  "TH34_2351_S01", "HMOX1")

outliers_to_keep <- left_join(focus_sample_gene_info_to_keep,
          outliers,
          by = c("Sample_ID", "gene"))
         
         
```



# COMPARE DISTRIBUTIONS FOR FOR OUTLIERS ACROSS COHORTS

```{r}


f <- "../input_data/druggable_plus_TumorCompendium_v11_PolyA_hugo_log2tpm_58581genes_2020-04-09.tsv.gz"
# f="../input_data/druggable_TumorCompendium_v11_PolyA_hugo_log2tpm_58581genes_2020-04-09.tsv.gz"
expr <- read_tsv(f) %>%
  rename(Sample_ID = TH_id) %>%
  filter(Gene %in% outliers_to_keep$gene)

stanford_samples  <- read_tsv("../gather_input_data/comparison_to_non_CARE_cohorts/data/TH03_TH34_rollup.sample_list.txt",
                              col_names = "Sample_ID") %>%
  mutate(cohort = "TH03_TH34")

TCGA_samples  <- read_tsv("../gather_input_data/comparison_to_non_CARE_cohorts/data/TCGA_rollup.sample_list.txt",
                              col_names = "Sample_ID") %>%
  mutate(cohort = "TCGA")

PEDAYA_samples  <- read_tsv("../gather_input_data/comparison_to_non_CARE_cohorts/data/PEDAYA_rollup.sample_list.txt",
                              col_names = "Sample_ID") %>%
  mutate(cohort = "PEDAYA")

pan_cancer_samples <- expr %>%
  select(Sample_ID) %>%
  distinct() %>%
  mutate(cohort = "Treehouse_pc")


samples_in_cohorts <- bind_rows(
  stanford_samples,
  TCGA_samples,
  PEDAYA_samples,
  pan_cancer_samples)


tabyl(samples_in_cohorts,
      cohort)
```


# expression in samples not in the compendium
```{r}

rsem_path <- "../input_data/non_compendium_expression"

gene_name_conversion <- read_tsv(file.path(rsem_path,
                                           "EnsGeneID_Hugo_Observed_Conversions.txt"))

relevant_gene_name_conversion <- gene_name_conversion %>%
  filter(HugoID %in% outliers_to_keep$gene)

rsem_kitchen_sink_data <- tibble(file_name = list.files(
  path = rsem_path,
  pattern = "_rsem_genes.results")) %>%
  rowwise() %>%
  mutate(rsem_raw = list(read_tsv(file.path(rsem_path, file_name),
                                     show_col_types = FALSE
                                     ))) %>%
  unnest(rsem_raw) %>%
  filter(gene_id %in% relevant_gene_name_conversion$EnsGeneID) %>%
  mutate(Sample_ID = str_extract(file_name, "TH[R]?[0-9]{2}_[0-9]{4}_S[0-9]{2}")) %>%
  left_join(relevant_gene_name_conversion, 
            by=c("gene_id"="EnsGeneID")) %>%
    group_by(Sample_ID, HugoID) %>%
    summarize(sum_TPM = sum(TPM),
              n=n()) %>%
    mutate(log2TPM1 = log2(sum_TPM +1))

table(rsem_kitchen_sink_data$n)

patient_expression_from_rsem_files <- 
  left_join(focus_sample_gene_info_to_keep,
            rsem_kitchen_sink_data %>%
              select(gene = HugoID,
                     log2TPM1,
                     Sample_ID))  %>%
  na.omit() # excludes samples not in rsem files

patient_expression_from_compendia <- focus_sample_gene_info_to_keep %>%
  distinct() %>%
  left_join(expr, 
            by=c("Sample_ID", "gene"="Gene")) %>%
  na.omit() # excludes samples not in compendium

patient_expression <- bind_rows(
  patient_expression_from_rsem_files,
  patient_expression_from_compendia)


```

# how many samples got data from rsem files?
```{r}
outliers_to_keep$Sample_ID[ ! outliers_to_keep$Sample_ID %in% expr$Sample_ID] %>% unique()
```

outliers and expression
```{r}

patient_expression

```
# Function to rename cohorts
```{r}
update_cohort_strings <- function(x) {
  str_replace(x,
              "PEDAYA", 
              "Pediatric") %>%
    str_replace("TH03_TH34",
                "Stanford")%>%
    str_replace("Treehouse_pc",
                "Treehouse pan-cancer")%>%
    str_replace("pan_disease",
                "Treehouse pan-disease,") %>%
    str_replace("1st_degree", "1st degree neighbors") %>%
    str_replace("1st_and_2nd_degree", "1st and 2nd degree neighbors") %>%
    str_replace("same_diagnosis", "same disease as focus sample") %>%
    str_replace("same_inferred_diagnosis", "diseases of top 6 most correlated datasets") %>%
    str_replace_all("_([a-z12][a-z])", " \\1")
}
```

# Function to calculate_outlier_thresholds
```{r}

calculate_outlier_threshold <- function(x) {
  # x <- 1:100
  q75 <- quantile(x, 0.75)
  iqr <- q75 - quantile(x, 0.25)
  threshold <- q75 + 1.5*iqr
  return(threshold)
}

```


```{r}

plot_distributions <- function(this_gene, this_sample_id, outlier_relative_to) {
  # this_sample_id <- "TH34_2351_S01"
  # this_gene <- "HMOX1"
  
  print(this_gene)
  
  print(this_sample_id)
  
  possible_levels <- c("Treehouse_pc", "TCGA", "PEDAYA", "TH03_TH34",
                       paste(this_sample_id, c("pan_disease_same_diagnosis",
                                               "pan_disease_same_inferred_diagnosis",
                                               "pan_disease_1st_degree",
                                               "pan_disease_1st_and_2nd_degree"))) %>%
    update_cohort_strings
  
  expr_relevant_to_this_outlier_raw <-  left_join(
    samples_in_cohorts %>%
      filter(str_detect(cohort, this_sample_id) | 
               cohort %in% c("PEDAYA", "TCGA", "TH03_TH34", "Treehouse_pc")),
    expr %>%
      filter(Gene == this_gene),
    by=c("Sample_ID")) %>%
    mutate(cohort = update_cohort_strings(cohort))
  
  expected_levels <- possible_levels[possible_levels %in% expr_relevant_to_this_outlier_raw$cohort]
  
  expr_relevant_to_this_outlier <- expr_relevant_to_this_outlier_raw%>%
    # na.omit() %>%  # while I'm missing expression data for some samples
    mutate(cohort = 
             fct_relevel(cohort, expected_levels)) %>%
    arrange(cohort) %>%
    group_by(cohort) %>%
    mutate(cohort_with_n = paste0(cohort, " (n=", scales::label_comma()(n()), ")")) %>%
    ungroup %>%
    mutate(cohort_with_n = factor(cohort_with_n, 
                                  levels = unique(cohort_with_n))) 
  
  
  outlier_thresholds <- expr_relevant_to_this_outlier %>%
    group_by(cohort_with_n) %>%
    summarize(
      outlier_threshold = calculate_outlier_threshold(log2TPM1)) %>%
    ungroup() %>%
    mutate(patient_expression = patient_expression %>%
             filter(gene == this_gene,
                    Sample_ID == this_sample_id) %>%
             pull(log2TPM1),
           patient_is_outlier = patient_expression > outlier_threshold)
  
  
  p <- ggplot(expr_relevant_to_this_outlier) +
    geom_boxplot(aes(y=cohort_with_n, x=log2TPM1),
                 #fill = cohort_with_n),
                 outlier.shape = NA) +
    geom_rect(data = outlier_thresholds,
              aes(xmin = outlier_threshold,
                  xmax = Inf,
                  ymin = -Inf,
                  ymax = Inf),
              fill = "yellow",
              alpha = 0.7) +
    geom_vline(data = outlier_thresholds,
               aes(xintercept = patient_expression,
                   #lty = patient_is_outlier
                   ), 
               color = "red") +
    facet_col(~cohort_with_n, scales = "free_y") +
    theme_minimal() +
    theme(axis.text.y = element_blank(),
          axis.title.y = element_blank(),
          legend.position="none",
          panel.border = element_rect(colour = NA, fill=NA, size=1),
          strip.background = element_rect(color = "lightgrey", size = 1, fill = "lightgrey"),
          panel.grid.major.y = element_blank()) +         #remove horizontal grid line
    ggtitle(paste0(this_gene, " cohorts for ", this_sample_id),
            "red line is patient expression (dashed if outlier); yellow area is above the outlier threshold")  
  
  
  print(p)
  
  return(p)
}
```



```{r fig.height=12, fig.width=8}

lapply(head(sort(outliers_to_keep$gene), 2), function(this_gene){
  # this_gene <- "BCL6"
  relevant_patient_expression <- patient_expression %>%
    filter(gene == this_gene) %>%
    filter(Sample_ID %in% (outliers_to_keep %>%
                             filter(gene == this_gene) %>% 
                             pull(Sample_ID)))
                           
  
  one_gene_expr_per_cohort <- left_join(samples_in_cohorts,
                                        expr %>%
                                          filter(Gene == this_gene))
   
  outlier_table <- outliers_to_keep %>%
    select(Sample_ID, gene, comparison_cohort) %>%
    mutate(found = TRUE) %>%
    pivot_wider(names_from = comparison_cohort,
                values_from = found,
                values_fill = FALSE) %>%
    filter(gene == this_gene) %>%
    select(-Treehouse_pd) %>%
    left_join(relevant_patient_expression,
              by = c("Sample_ID", "gene")) %>%
    mutate(log2TPM1 = round(log2TPM1, 3))
  
  outlier_table_long <- outlier_table %>%
    pivot_longer(cols = c(-Sample_ID, -log2TPM1, -gene),
                 names_to = "cohort",
                 values_to = "outlier") %>%
    mutate(outlier_status = c("non outlier", "outlier")[1+outlier])
  
 cohort_thresholds <-  one_gene_expr_per_cohort %>%
  group_by(cohort) %>%
  summarize(q25 = quantile(log2TPM1, 0.25),
            median = median(log2TPM1),
            q75 = quantile(log2TPM1, 0.75),
            IQR = q75-q25,
            up_outlier_threshold = q75 + (1.5*IQR)) %>%
  pivot_longer(-cohort) %>%
  mutate(value = round(value, 2)) %>%
  pivot_wider()
  
  
   p1 <- ggplot(one_gene_expr_per_cohort) +
    geom_histogram(aes(x=log2TPM1)) +
    geom_vline(data = relevant_patient_expression,
               aes(xintercept = log2TPM1,
                   #color = Sample_ID,
                   #lty = outlier_status
                   ), color = "red") +
    scale_color_brewer(palette = "Set1") +
    facet_col(~cohort, scales = "free_y") +
  ggtitle(this_gene)
  
  p2 <- ggplot(one_gene_expr_per_cohort) +
    geom_boxplot(aes(x=log2TPM1)) +
    geom_vline(data = relevant_patient_expression,
               aes(xintercept = log2TPM1,
                   #color = Sample_ID,
                   #lty = outlier_status
                   ), color = "red") +
    scale_color_brewer(palette = "Set1") +
    facet_col(~cohort) 
  
  t1 <- tableGrob(outlier_table, theme=ttheme_minimal(), 
                  rows=NULL)  # transform into a tableGrob
  
  t2 <- tableGrob(cohort_thresholds, theme=ttheme_minimal(), 
                  rows=NULL)  # transform into a tableGrob
  
  plot_grid(p1, p2, t1, t2,
                     ncol = 1)
})



```


# table for annotating TCGA vs Treehouse pc
```{r}

outlier_table_for_annotation <- outliers_to_keep %>%
    select(Sample_ID, gene, comparison_cohort) %>%
    mutate(found = TRUE) %>%
    pivot_wider(names_from = comparison_cohort,
                values_from = found,
                values_fill = FALSE) %>%
  arrange(gene) #%>%
  #mutate(TCGA_not_treehouse_pc = TCGA & ! Treehouse_pc) %>%
  #select(gene, Sample_ID, TCGA, Treehouse_pc, TCGA_not_treehouse_pc, everything())
  
# write_tsv(outlier_table_for_annotation, "../gather_input_data/comparison_to_non_CARE_cohorts/TCGA_vs_TH_outlier_table_for_annotation.tsv")
# 
# length(unique(outlier_table_for_annotation$gene))
#   filter(
# table(outlier_table_for_annotation$TCGA_not_treehouse_pc)
#   table(outlier_table_for_annotation$TCGA_not_treehouse_pc)
#   
#   outlier_table_for_annotation %>%
#     filter(TCGA_not_treehouse_pc,
#            !PEDAYA,
#            !TH03_TH34)

```

