---
title: "`r gsub('.Rmd', ' ', gsub('_', ' ', knitr::current_input()))`"
author: "`r Sys.getenv('USER')`"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
---
  
  
# version notes  

2023.12.20_13.42.16 - i got it to recreate the existing plot

make minimal incremental improvements

2023.12.21_11.12.06 - improvement 1: update three labels, e.g. Standard of care to Received therapy considered standard of care

2023.12.21_11.21.42 - improvement 2: add "accepted", e.g. Standard of care to Received therapy considered standard of care

conclusion for 2023.12.21_11.21.42: color assignments work only on the first word; see how both "Received therapy" sections are the same color. I'm going to consider this a provisional success, since changing color is much easier than changing text

2023.12.21_12.07.04 - improvement 3: add n = 
conclusion for 2023.12.21_12.07.04: colors are totally off; i can't wrap text, so i had to make it wider. still this includes the numbers, which i think are important. i also can't move columns closer or farther separately, so I have to make the whole thing wider. 

river_plot_for max_copy of 023.12.21_12.07.04
remove all output to simplify html for max to review


```{r setup, include = FALSE}

library(tidyverse)
library(networkD3)
library(RColorBrewer)
library(colorspace)

```




# import data
```{r, include = FALSE}
findings_file <- "../input_data/Treehouse.Drug.Prioritization.No.Identifiers.final.03.07.21 - case_prioritization.tsv"



tx_treatments_raw <- read_tsv(findings_file, 
                              col_types = cols(.default = "c")) %>% 
   select(-TH_sampleID, -`Case #`, -Note)


length(unique(tx_treatments_raw$TH_donorID))
```

# nodes in order
```{r, include = FALSE}

nodes_in_order_for_plotting <- c(  #things we want to explicitly sort
  "Prioritized",
  "Deferred",
  "Declined",
  "Implemented",
  "Not implemented",
  #Implemented category: 
  "Response",
  "Progression",
  # Not implemented
  "Rapid disease progression", 
  "Waiting for treatment approval", 
  "Therapy unavailable",
  # Deferred
  "Standard of care", 
  "No further treatment needed",
  "Treatment with more disease data", 
  # Declined
  "Evidence against Treehouse therapy", 
  "Not FDA approved")


nodes_in_order_for_color <- c(  #positive events
  "Prioritized",
  "Implemented",
  "Response",
   #second rank events
  "Progression",
  "Not implemented",
  "Rapid disease progression", 
  # Not implemented
  "Waiting for treatment approval", 
  "Therapy unavailable",
  # deferred
  "Deferred",
  "Standard of care", 
  "No further treatment needed",
  "Treatment with more disease data", 
  "Declined",
  "Evidence against Treehouse therapy", 
  "Not FDA approved")

```

# assemble data in format required for sankey plot
```{r, include = FALSE}

all_nodes <- tibble(name = c(tx_treatments_raw$TH_donorID, 
                                tx_treatments_raw$`Case prioritization`, 
                                tx_treatments_raw$Reason, 
                                tx_treatments_raw$`Implementation reason for prioritized findings`)) %>% 
  na.omit()

nodes <- all_nodes %>%
  distinct() %>%
  mutate(name_factor = factor(name) %>% 
           fct_relevel(c(tx_treatments_raw$TH_donorID,
                         nodes_in_order_for_plotting))) %>%
  arrange(name_factor) %>%
  mutate(node = 0:(n()-1)) 


```


# Update language
improvements 1, 2 and 3
```{r, include = FALSE}

node_labels <- all_nodes  %>%
  mutate(is_donor = str_detect(name, "TH[R]?[0-9]{2}_[0-9]{4}")) %>%
  rename(unnumbered_name = name) %>%
  group_by(is_donor, unnumbered_name) %>%
  summarize(n = n()) 

update_names <- function(x) {
  str_replace_all(x, "Standard of care", 
                  "Received therapy considered standard of care") %>%
    str_replace_all("Treatment with more disease data", 
                    "Received therapy with more published evidence of efficacy") %>%
    str_replace_all("Evidence against Treehouse therapy", 
                    "Evidence against therapy identified by CARE analysis") %>%
    str_replace_all("Prioritized",
                    "Accepted; prioritized") %>%
    str_replace_all("Deferred",
                    "Accepted; deferred")
}


add_count_to_name <- function(x) {
  # x <- all_nodes$name %>% unique
  tibble(name = x) %>%
    left_join(node_labels, by=c("name"="unnumbered_name")) %>%
  mutate(name_with_number = ifelse(is_donor,
                                   name,
                                   paste0(update_names(name), " (n=", n, ")"))) %>%
    pull(name_with_number)
}

#add_count_to_name(all_nodes$name)
```

# Update node data
```{r, include = FALSE}
sample_id_placeholder <- ""

# nodes_wo_sample_names <- nodes %>%
#    mutate(name2  = name)
nodes_wo_sample_names <- nodes %>%
#  mutate(name2 = name)
  mutate(name2 = case_when(str_detect(name_factor, "^TH[0-9]") ~ sample_id_placeholder,
            TRUE ~ name))

updated_node_names_wo_sample_names <- nodes_wo_sample_names %>%
  mutate(name2 = add_count_to_name(name2))

```

# create does
```{r, include = FALSE}
# links <- source target value
donor_to_prioritization <- tx_treatments_raw %>%
  select(TH_donorID, `Case prioritization`) %>%
  left_join(nodes, by=c("TH_donorID"="name")) %>%
              rename(source = node) %>%
  left_join(nodes, by=c(`Case prioritization`="name")) %>%
              rename(target = node) %>%
  mutate(value = 1)
  
prioritization_to_reason <- tx_treatments_raw %>%
  select(`Case prioritization`, Reason) %>%
  left_join(nodes, by=c("Case prioritization"="name")) %>%
              rename(source = node) %>%
  left_join(nodes, by=c("Reason"="name")) %>%
              rename(target = node) %>%
  mutate(value = 1)

reason_to_implementation <-  tx_treatments_raw %>%
  select(Reason, `Implementation reason for prioritized findings`) %>%
  left_join(nodes, by=c("Reason"="name")) %>%
              rename(source = node) %>%
  left_join(nodes, by=c("Implementation reason for prioritized findings"="name")) %>%
              rename(target = node) %>%
  mutate(value = 1) %>%
  na.omit

links_noisy <- bind_rows(donor_to_prioritization,
                   prioritization_to_reason,
                   reason_to_implementation) 

links <- links_noisy %>%
  select(source, target, value) %>%
  arrange(source, target, value)

  

```

# define colors
```{r, include = FALSE}

#names_to_color_blue <- tibble(name = c("Accepted; prioritized", "Implemented", "Response"), 
names_to_color_blue <- tibble(name = c("Prioritized", "Implemented", "Response"), 
                              color_code =  c("#9ECAE1", "#6BAED6", "#3182BD"))
                              #brewer.pal(name="Blues",n=6)[3:5])

sample_box_colors <- tibble(name = sample_id_placeholder,
                            color_code = "#D3D3D3")

node_names <- tibble(name = unique(nodes_wo_sample_names$name2)) %>%
  filter(! name %in% c(names_to_color_blue$name, sample_box_colors$name)) %>%
  mutate(name_f = factor(name, levels = nodes_in_order_for_color)) %>%
  arrange(name_f) %>%
  mutate(color_code = heat_hcl(n(),rev=TRUE, c=c(60,20),l=c(60,100),power=1)) %>%
  bind_rows(names_to_color_blue) %>%
  bind_rows(sample_box_colors) %>%
  mutate(name = add_count_to_name(name))

color_specification <- JS(
sprintf(
'd3.scaleOrdinal()  
  .domain(%s)
  .range(%s)
',
jsonlite::toJSON(node_names$name),
jsonlite::toJSON(node_names$color_code)
   )
  )



```


# check for overlap in colors and names
```{r, include = FALSE}
setdiff(node_names$name, nodes_wo_sample_names$name2) # expected output: character(0)
setdiff(nodes_wo_sample_names$name2, node_names$name) # expected output: character(0)

```


```{r}

plotly_sankey_data_template <- list(type = "sankey", 
     domain = list(x = c(0, 1), y = c(0, 1)), 
     orientation = "v", 
     valueformat = ".0f", 
     valuesuffix = "TWh",
     node = 
       list(pad = 15, 
            thickness = 15, 
            line = list(color = "black", width = 0.5), 
            label = c("Agricultural 'waste'", "Bio-conversion", 
                      "Liquid", "Losses", "Solid", "Gas", "Biofuel imports", "Biomass imports", 
                      "Coal imports", "Coal", "Coal reserves", "District heating", 
                      "Industry", "Heating and cooling - commercial", "Heating and cooling - homes", 
                      "Electricity grid", "Over generation / exports", "H2 conversion", 
                      "Road transport", "Agriculture", "Rail transport", "Lighting & appliances - commercial", 
                      "Lighting & appliances - homes", "Gas imports", "Ngas", "Gas reserves", 
                      "Thermal generation", "Geothermal", "H2", "Hydro", "International shipping", 
                      "Domestic aviation", "International aviation", "National navigation", 
                      "Marine algae", "Nuclear", "Oil imports", "Oil", "Oil reserves", 
                      "Other waste", "Pumped heat", "Solar PV", "Solar Thermal", 
                      "Solar", "Tidal", "UK land based bioenergy", "Wave", "Wind"
            )),
     link = 
       list(
         source = c(0, 1, 1, 1, 1, 6, 7, 8, 10, 9, 
                    11, 11, 11, 15, 15, 15, 15, 15, 15, 15, 15, 15, 15, 15, 23, 
                    25, 5, 5, 5, 5, 5, 27, 17, 17, 28, 29, 2, 2, 2, 2, 2, 2, 
                    2, 2, 34, 24, 35, 35, 36, 38, 37, 39, 39, 40, 40, 41, 42, 
                    43, 43, 4, 4, 4, 26, 26, 26, 44, 45, 46, 47), 
         target = c(1, 
                    2, 3, 4, 5, 2, 4, 9, 9, 4, 12, 13, 14, 16, 14, 17, 12, 18, 
                    19, 13, 3, 20, 21, 22, 24, 24, 13, 3, 26, 19, 12, 15, 28, 
                    3, 18, 15, 12, 30, 18, 31, 32, 19, 33, 20, 1, 5, 26, 26, 
                    37, 37, 2, 4, 1, 14, 13, 15, 14, 42, 41, 19, 26, 12, 15, 
                    3, 11, 15, 1, 15, 15), 
         value = c(124.729, 0.597, 26.862, 
                   280.322, 81.144, 35, 35, 11.606, 63.965, 75.571, 10.639, 
                   22.505, 46.184, 104.453, 113.726, 27.14, 342.165, 37.797, 
                   4.412, 40.858, 56.691, 7.863, 90.008, 93.494, 40.719, 82.233, 
                   0.129, 1.401, 151.891, 2.096, 48.58, 7.013, 20.897, 6.242, 
                   20.897, 6.995, 121.066, 128.69, 135.835, 14.458, 206.267, 
                   3.64, 33.218, 4.413, 14.375, 122.952, 500, 339.978, 504.287, 
                   107.703, 611.99, 56.587, 77.81, 193.026, 70.672, 59.901, 
                   19.263, 19.263, 59.901, 0.882, 400.12, 46.477, 525.531, 787.129, 
                   79.329, 9.452, 182.01, 19.013, 289.366), 
         label = c("stream 1", 
                   "", "", "", "stream 1", "", "", "", "", "", "", "", "", "", 
                   "", "", "", "", "", "", "", "", "", "", "", "", "", "", "", 
                   "", "stream 1", "", "", "", "", "", "", "", "", "", "", "", 
                   "", "", "", "", "Old generation plant (made-up)", "New generation plant (made-up)", 
                   "", "", "", "", "", "", "", "", "", "", "", "", "", "", "", 
                   "", "", "", "", "", ""))
)

```


```{r}
plotly_links <- links %>%
  mutate(label = "")
```

```{r}
library(plotly)
library(rjson)
fig <- plot_ly(
  type = "sankey",
  domain = list(
    x =  c(0,1),
    y =  c(0,1)
  ),
  orientation = "h",
  valueformat = ".0f",
  valuesuffix = "TWh",
  
  node = list(
    # label = json_data$data[[1]]$node$label,
    label = updated_node_names_wo_sample_names$name2,
   #  label = labels_with_breaks,
    #color = json_data$data[[1]]$node$color,
    pad = 15,
    thickness = 15,
    line = list(
      color = "black",
      width = 0.5
    )
  ),
  
  link = list(
    source = plotly_links$source,
    target = plotly_links$target,
    value =  plotly_links$value,
    label =  plotly_links$label
  )
)
fig <- fig %>% layout(
  title = "Energy forecast for 2050<br>Source: Department of Energy & Climate Change, Tom Counsell via <a href='https://bost.ocks.org/mike/sankey/'>Mike Bostock</a>",
  font = list(
    size = 10,
    color = 'white'
  ),
  xaxis = list(showgrid = F, zeroline = F, showticklabels = F),
  yaxis = list(showgrid = F, zeroline = F, showticklabels = F),
  plot_bgcolor = 'black',
  paper_bgcolor = 'black'
)

fig
```

# plot
```{r fig.height=6, fig.width=10}

networkD3::sankeyNetwork(Links = links, Nodes = updated_node_names_wo_sample_names, Source = 'source', 
                         Target = 'target', Value = 'value', NodeID = 'name2',
                         units = '',
  sinksRight = FALSE,
  nodeWidth = 40,
  fontSize = 13,
  nodePadding = 5,
  iterations = 10,
  colourScale = color_specification)

```



