---
title: "`r gsub('.Rmd', ' ', gsub('_', ' ', knitr::current_input()))`"
author: "`r Sys.getenv('USER')`"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: yes
---
  


```{r setup, include = FALSE}

library(tidyverse)

```



# Load outlier results
```{r}

outliers <- read_tsv("../input_data/druggable_outliers_from_treehouse_and_other_cohorts_2023_11_09-13_46_32_2023.tsv") %>%
  mutate(high_level_cohort = ifelse(str_detect(comparison_cohort, "Treehouse"),
                                    "Treehouse",
                                    comparison_cohort))

```

# Identify outliers present in TCGA but not other cohorts

```{r}


TCGA_only_outliers <- outliers %>%
  group_by(gene, Sample_ID) %>%
  mutate(TCGA_only = "TCGA" %in% comparison_cohort & 
           n() == 1) %>%
  filter(TCGA_only) %>%
  arrange(Sample_ID, gene)


```

## Featuers of TCGA_only_outliers
```{r}

table(TCGA_only_outliers$gene)



```


# Load lists of samples in TCGA and pedaya cohorts
```{r}
TCGA_samples  <- read_tsv("../gather_input_data/comparison_to_non_CARE_cohorts/data/TCGA_rollup.sample_list.txt",
                              col_names = "Sample_ID") %>%
  mutate(cohort = "TCGA")

PEDAYA_samples  <- read_tsv("../gather_input_data/comparison_to_non_CARE_cohorts/data/PEDAYA_rollup.sample_list.txt",
                              col_names = "Sample_ID") %>%
  mutate(cohort = "PEDAYA")


samples_in_cohorts <- bind_rows(
  TCGA_samples,
  PEDAYA_samples,
  )

```

#  load expression for samples in TCGA and pedaya cohorts
```{r}
expr_file <- "../input_data/druggable_plus_TumorCompendium_v11_PolyA_hugo_log2tpm_58581genes_2020-04-09.tsv.gz"
expr <- read_tsv(expr_file) %>%
  rename(Sample_ID = TH_id) %>%
  filter(Gene %in% TCGA_only_outliers$gene)
```

## confirm that all samples in cohorts have expression data
```{r}

samples_in_cohorts %>%
  filter(!Sample_ID %in% expr$Sample_ID)

```

```{r}

```


# Calculate statistics for each cohort
```{r}

cohort_thresholds_raw <- left_join(samples_in_cohorts,
                                   expr,
                                   by=c("Sample_ID")) %>%
  group_by(Gene, cohort) %>%
  summarize(q25 = quantile(log2TPM1, 0.25),
            median = median(log2TPM1),
            q75 = quantile(log2TPM1, 0.75),
            IQR = q75-q25,
            up_outlier_threshold = q75 + (1.5*IQR))
  
```

# assess changes for all genes
```{r}
cohort_thresholds <- cohort_thresholds_raw %>%
  pivot_longer(c(-Gene, -cohort)) %>%
  pivot_wider(names_from = cohort, values_from = value) %>%
  mutate(change_in_ped_relative_to_TCGA = 
              (PEDAYA - TCGA) / TCGA)
```



# changes plotted
```{r}
ggplot(cohort_thresholds) +
  geom_point(aes(x=TCGA,
                 y=change_in_ped_relative_to_TCGA,
                 color = name)) +
  geom_hline(yintercept = 1) +
  scale_color_brewer(palette = "Set1")
```

```{r}

cohort_thresholds %>%
  filter(name %in% c("median", "IQR")) %>%
  select(Gene, name, change_in_ped_relative_to_TCGA) %>%
  pivot_wider(names_from = name,
              values_from = change_in_ped_relative_to_TCGA) %>%
  ggplot + 
  geom_point(aes(x=median, y=IQR)) + 
  ggtitle("The IQR usually increased irrespective of the direction of change of the median", "fraction change_in_ped_relative_to_TCGA")


```

```{r}

cohort_thresholds %>%
  filter(name %in% c("median", "IQR"))

iqrs <- cohort_thresholds %>%
  filter(name=="IQR") %>%
  rename(PEDAYA_IQR = "PEDAYA",
         TCGA_IQR = "TCGA",
         change_in_ped_IQR_relative_to_TCGA = "change_in_ped_relative_to_TCGA") %>%
  select(-name)


medians <- cohort_thresholds %>%
  filter(name=="median") %>%
  rename(PEDAYA_median = "PEDAYA",
         TCGA_median = "TCGA",
         change_in_ped_median_relative_to_TCGA = "change_in_ped_relative_to_TCGA") %>%
  select(-name)

table_of_compared_stats <- left_join(iqrs, medians) %>%
  ungroup() %>%
  mutate(across(where(is.double), round, 2))

```


```{r}

file_output_name <- paste0("../results/comparison_of_IQR_and_median_in_TCGA_vs_PedAYA_distributions", format(Sys.time(), "%Y_%m_%d-%H_%M_%S_%Y"), ".tsv")

write_tsv(table_of_compared_stats, file_output_name)

```

this output was saved as `r file_output_name`

# Find example
```{r}
cohort_thresholds %>%
  group_by(Gene) %>%
  mutate(lower_med_in_pedaya_than_TCGA = PEDAYA[name == "median"] < TCGA[name == "median"]) %>%
  rename(change = change_in_ped_relative_to_TCGA) %>%
  filter(lower_med_in_pedaya_than_TCGA,
         name == "up_outlier_threshold")

cohort_thresholds %>%
  filter(name == "up_outlier_threshold")

BCL6
```
```{r}
table_of_compared_stats %>%
  filter(Gene == "BCL6")

cohort_thresholds  %>%
  filter(Gene == "BCL6")
```

# Other genes considered
```{r}
table_of_compared_stats %>%
  filter(Gene == "CCND3")

cohort_thresholds  %>%
  filter(Gene == "CCND3") #this isn't a good example because CCND3 is dramatic and has a higher median in ped than tcga


table_of_compared_stats %>%
  filter(Gene == "PDGFRA")

cohort_thresholds  %>%
  filter(Gene == "PDGFRA")  #this isn't a good esxample beause CCND3 is dramatic and has a higher median in ped than tcga


  
```


### support for text
21 outliers were uniquely detected in this (TCGA) comparison and not present in comparisons against the Stanford or pediatric cohorts.

```{r}

nrow(TCGA_only_outliers)

```

these 21 outliers were detected in 16 genes
```{r}
n_distinct(TCGA_only_outliers$gene)
```

all 16 have wider distributions of expression in pediatric cohorts compared to TCGA. 

```{r}
sum(table_of_compared_stats$change_in_ped_IQR_relative_to_TCGA<0)
```


```{r}
table(table_of_compared_stats$change_in_ped_median_relative_to_TCGA<0)
```
```{r}

```
# more analysis

## sample for plot
TH34_2292_S01
```{r}
TCGA_only_outliers %>%
  filter(gene == "BCL6")
```

