

```{r}
library(tidyverse)

```


```{r}
data_dir <- "/private/groups/treehouse/working-projects/hbeale/ckcc2_2023_07"
paths_file <- "PathsToOriginalTertiary.2022_01_18.tsv"
samples_file <- "ckcc2_metadata_from_redcap_2021-07-29_11.38.45.tsv"

paths_raw <- read_tsv(file.path(data_dir, paths_file),
                      col_types = cols(.default = "c"))
nrow(paths_raw)

project_samples <- read_tsv(file.path(data_dir, samples_file),
                      col_types = cols(.default = "c"))
nrow(project_samples)
# reminder; there are more lines in the paths_raw tibble than
# in the project_samples tibble because some samples have
# multiple tert outputs


# check whether all samples are present in both data sources
# if both groups are identical,  result will be "character(0)"
setdiff(paths_raw$Sample, project_samples$THid)
setdiff(project_samples$THid, paths_raw$Sample)

tertiary_output_locations <- paths_raw %>%
  select(sample_id = Sample, tertiary_path = `Original tertiary path`)

```

# Compendium and protocol version survey
```{r}
c_p_survey <- tertiary_output_locations %>%
  mutate(compendium = str_remove(tertiary_path, "^.*/"),
         protocol = str_replace(tertiary_path, "^.*treehouse-[a-z]*-([^/]*)/.*$", "\\1")) %>%
  select(-tertiary_path)

c_p_survey
install.packages("janitor")
library(janitor)
tabyl(c_p_survey , compendium)
tabyl(c_p_survey , protocol)
```

# Identify samples with multiple tertiaries:
```{r}

tertiary_output_locations %>%
group_by(sample_id) %>%
  mutate(n_terts=n()) %>%
  filter(n_terts > 1) %>%
  pull(tertiary_path)

```

# Load druggable genes
```{r}

druggable_genes <- read_tsv(file.path(data_dir, "treehouseDruggableGenes_2020-03_25.txt"))

```

# Load pan cancer outlier results
details on file structure: https://github.com/UCSC-Treehouse/reference-file-info/blob/master/Explainer%20for%20outlier_results_files.txt
```{r}

tertiary_output_data <- tertiary_output_locations %>%
 group_by(sample_id) %>% 
  top_n(1, tertiary_path) %>% # select only one of two for the samples with multiple outputs
mutate(leads_raw = map(paste0(tertiary_path, "/outlier_results_", sample_id), read_tsv, col_types = cols(.default = "c"))) %>%
unnest(leads_raw)

top_5 <- tertiary_output_data %>%
  filter(is_top_5 == "top5") %>%
  select(sample_id, Gene)
  
pc_outliers <- tertiary_output_data %>%
  filter(pc_outlier=="pc_up") %>%
  mutate(is_druggable = Gene %in% druggable_genes$gene)
table(pc_outliers$is_druggable)


```

# Identify outliers present in two custom pan-disease groups from different tertiary rounds
In this step, we analyze the outputs across the custom cohorts (ACC) to account for these samples with multiple tertiary outputs

```{r}
# code for samples with two cohorts, currently TH34_1349_S01, TH34_1349_S02
library(jsonlite)

get_pan_disease_outliers <- function(this_tert_path){
  j4 <- fromJSON(file.path(this_tert_path, "4.0.json"))
  tibble(gene = names(j4$personalized_consensus_counts$up_outliers))
}

# confirmed with code: everything in the personalized_consensus_counts up_outliers in top 5

pan_disease_outliers <- tertiary_output_locations %>%
  mutate(pan_disease_outliers_raw = map(tertiary_path, get_pan_disease_outliers)) %>%
  unnest(pan_disease_outliers_raw)

# if there are no pan-disease outliers, the function returns nothing and that sample has no line in the outliers tibble, so e.g. TH34_1149_S02 isn't in the pan-disease outlier list


pan_disease_outliers_anno <- pan_disease_outliers %>%
            mutate(is_druggable = gene %in% druggable_genes$gene)

```

```{r}

write_tsv(pan_disease_outliers_anno, paste0("pan_disease_outliers_from_", paths_file, ".tsv"))

write_tsv(pc_outliers %>%
  select(sample_id,
         tertiary_path,
         Gene,
    is_druggable
  ), paste0("pan_cancer_outliers_from_", paths_file, ".tsv"))

```

