---
title: "`r gsub('.Rmd', ' ', gsub('_', ' ', knitr::current_input()))`"
author: "`r Sys.getenv('USER')`"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
---
  
# version notes

version 2023.11.06_11.32.08 is for troubleshooting how many values are in Molecular Category
version 2023.11.07_12.48.01 is for reviewing data to continue troubleshooting which values to use in Molecular Category
version 2023.11.08_10.00.39 is for adding "IGF1R", "RAS/RAF/MEK" to Molecular Category
version 2023.11.08_12.53.45 is for adding "NTRK", "VEGF", "ERBB" to Molecular Category

```{r setup, include = FALSE}

library(tidyverse)
library(knitr)
library(janitor)
library(kableExtra)
```


# Load and combine data
```{r}


findings_file <- "../input_data/Supplementary_Tables - eTable8_dl_2023.11.06_11.03.03.tsv"

findings <- read_tsv(findings_file,
                     skip = 1) %>%
  rename(donorID = `Patient ID`)

```


```{r}

latest_sample_metadata_file_name <- "../input_data/ckcc2_samples_and_metadata_from_redcap_2023.10.24_13.26.34.tsv"

sample_metadata <- read_tsv(latest_sample_metadata_file_name,
                            col_types = cols(.default = "c")) %>%
  mutate(dx_group_factor = fct_lump_min(dx_group, 2),
         dx_group = as.character(dx_group_factor))

donor_disease <- sample_metadata %>%
  select(donorID, disease, dx_group, age_at_dx, gender) %>%
  distinct() 

```




```{r}

findings_with_metadata <- left_join(
  findings,
  donor_disease,
  by = "donorID")

```

# define molecular categories to use
```{r}

molecular_categories <- 
  bind_rows(
    tibble(molecular_category_name = c("PI3K/AKT/mTOR", "Cell Cycle", "RTK", "JAK/STAT"),
       molecular_category_type = "predefined_molecular_category"),
tibble(molecular_category_name = c("IGF1R", "RAS/RAF/MEK", "NTRK", "VEGF", "ERBB"),
       molecular_category_type = "other_molecular_category"))

```


# Review 
```{r}


library(forcats)
finding_categories <- findings_with_metadata %>%
  rename(molecular_category_name = `Molecular Category`,
         molecular_abnormality_name = `Molecular Abnormality`) %>%
  select(molecular_category_name, donorID, molecular_abnormality_name) %>%
  left_join(molecular_categories, by = "molecular_category_name") %>%
  replace_na(list(molecular_category_type = "none")) %>%
  mutate(molecular_category_type = factor(molecular_category_type, 
                                             levels = c(
                                               unique(molecular_categories$molecular_category_type),
                                               "none")),
         simplified_molecular_category = ifelse(molecular_category_type != "none",
                                      molecular_category_name,
                                      "Other") %>%
           factor(levels = c(molecular_categories$molecular_category_name,
                                "Other"))) %>%
  group_by(simplified_molecular_category) %>%
  mutate(n_donors_with_vulnerability = n(),
         vulnerability_label = paste0(simplified_molecular_category, " (n=", 
                                      n_donors_with_vulnerability, ")")) %>%
  arrange(molecular_category_type, simplified_molecular_category, donorID) %>%
  ungroup

```

# Findings info for each donor
```{r}

finding_categories_not_other <- finding_categories %>%
  filter(!simplified_molecular_category == "Other")
  
finding_categories_other <- finding_categories %>%
  filter(simplified_molecular_category == "Other")
  

finding_categories_not_other %>%
  select(donorID, molecular_abnormality_name, molecular_category_name) %>%
  # select(-n_donors_with_vulnerability,
  #        -molecular_category_type,
  #        -simplified_molecular_category, 
  #        -vulnerability_label) %>%
  kbl() %>%
  kable_styling(full_width = F) %>% 
  group_rows(index = setNames(rle(as.vector(finding_categories_not_other$molecular_category_type))[[1]],
                              rle(as.vector(finding_categories_not_other$molecular_category_type))[[2]])) %>% 
  group_rows(index = setNames(rle(as.vector(finding_categories_not_other$vulnerability_label))[[1]],
                              rle(as.vector(finding_categories_not_other$vulnerability_label))[[2]]))


finding_categories_other %>%
  select(donorID, molecular_abnormality_name, molecular_category_name, simplified_molecular_category) %>%
  # select(-n_donors_with_vulnerability,
  #        -molecular_category_type,
  #        -vulnerability_label) %>%
  kbl() %>%
  kable_styling(full_width = F) %>% 
  group_rows(index = setNames(rle(as.vector(finding_categories_other$molecular_category_type))[[1]],
                              rle(as.vector(finding_categories_other$molecular_category_type))[[2]]))%>% 
  group_rows(index = setNames(rle(as.vector(finding_categories_other$vulnerability_label))[[1]],
                              rle(as.vector(finding_categories_other$vulnerability_label))[[2]]))


```

# findings per molecular_abnormality
```{r}

findings_per_molecular_abnormality <- finding_categories %>%
  group_by(molecular_abnormality_name, simplified_molecular_category,
         molecular_category_type) %>%
  summarize(n_donors=n()) %>%
  group_by(simplified_molecular_category) %>%
  mutate(n_total_donors = sum(n_donors),
         category_label = paste0(simplified_molecular_category, 
                                 ", n=(", n_total_donors, ")")) %>%
  arrange(molecular_category_type, simplified_molecular_category) 

findings_per_molecular_abnormality %>%
  ungroup %>%
  select(-simplified_molecular_category, -n_total_donors, 
         -category_label) %>%
  kbl() %>%
  kable_styling(full_width = F) %>% 
  group_rows(index = setNames(rle(as.vector(findings_per_molecular_abnormality$category_label))[[1]],
                              rle(as.vector(findings_per_molecular_abnormality$category_label))[[2]]))

table(finding_categories$simplified_category)


```

# findings per aggregated molecular_category
```{r}

finding_categories %>%
  group_by(simplified_molecular_category,
         molecular_category_type) %>%
  summarize(n_donors=n()) %>%
  arrange(molecular_category_type, desc(n_donors)) %>%
  kbl() %>%
  kable_styling(full_width = F)



```

# findings per molecular category (not aggregated)
```{r}

   
finding_categories %>%
  group_by(molecular_category_type, molecular_category_name) %>%
  summarize(n=n()) %>%
  arrange(desc(n)) %>%
  mutate(running_total = cumsum(n),
         remaining_findings = sum(n)-running_total) %>%
  # adorn_pct_formatting() %>%
  kbl() %>%
  kable_styling(full_width = F)

```

# Example barplot
```{r}
ggplot(finding_categories) +
  geom_bar(aes(x = simplified_molecular_category))
```


## Load gene and disease groups 
```{r}

gene_group_info <- read_tsv("../input_data/gene_group_info.tsv", col_types = cols(.default = "c"))

disease_group_names <- read_tsv(file.path("../input_data/", "disease_group_info.tsv"), col_types = cols(.default = "c"))

```
# expected gene groups
```{r}

gene_group_info %>%
  select(outlier_gene_group) %>% 
  kbl(format.args = list(big.mark = ","), digits = c(NA, NA, 4, 4, 3, 2)) %>%
  kable_styling(full_width = F)

```

```{r}
pre_defined_arms <- c("JAK/STAT","RTK", "Cell Cycle", "PI3K/AKT/mTOR")
pre_defined_arms
```



# View gene set colors
```{r}

gene_group_info %>%
  mutate(gene_group_color = cell_spec(
    gene_group_color, color = "white", bold = T,
    background = gene_group_color
  )) %>%
  kable(escape = F, align = "c") %>%
  kable_styling(c("striped", "condensed"), full_width = F)
  

```

# Session Info
```{r}
sessionInfo()
```

# Appendix

## Full findings table
```{r}

findings %>%
  select(donorID, Drug, `Molecular Abnormality`, `Molecular Category`) %>%
  kbl() %>%
  kable_styling(full_width = F)
```

