---
title: "`r gsub('.Rmd', ' ', gsub('_', ' ', knitr::current_input()))`"
author: "`r Sys.getenv('USER')`"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
---
  
# version notes

version 2023.11.06_11.32.08 is for troubleshooting how many values are in Molecular Category

version 2023.11.07_12.48.01 is for reviewing data to continue troubleshooting which values to use in Molecular Category

version 2023.11.08_10.00.39 is for adding "IGF1R", "RAS/RAF/MEK" to Molecular Category

version 2023.11.08_12.53.45 is for adding "NTRK", "VEGF", "ERBB" to Molecular Category

version 2023.11.08_14.13.38 improves tables

version 2023.11.15_09.54.01 uses updated data from Yvonne and adds disease

version 2023.12.07_16.19.18 removes the facet for "ad hoc"

version 2023.12.15_13.55.55 uses different color schemes

version 2023.12.19_14.22.18 after picking a color scheme



```{r setup, include = FALSE}

library(tidyverse)
library(knitr)
library(janitor)
library(kableExtra)
library(ggforce) # for facet_row
```


# latest files
```{r}

findings_file <- "../input_data/Treehouse finding prioritization by patient-eTable8_2023.11.15_09.44.25.txt"

sample_metadata_file <- "../input_data/ckcc2_metadata_from_redcap_2023-11-09_10.34.27.tsv"

```


# Load and combine data
```{r}


findings <- read_tsv(findings_file,
                     skip = 1) %>%
  rename(donorID = `Patient ID`)

```


```{r}



sample_metadata <- read_tsv(sample_metadata_file,
                            col_types = cols(.default = "c")) %>%
  mutate(dx_group_factor = fct_lump_min(dx_group, 2),
         dx_group = as.character(dx_group_factor))

donor_disease <- sample_metadata %>%
  select(donorID, disease, dx_group, age_at_dx, gender) %>%
  distinct() 

```




```{r}

findings_with_metadata <- left_join(
  findings,
  donor_disease,
  by = "donorID")

```

there are `r nrow(findings_with_metadata) ` findings


# annotate findings with more categories
```{r}

finding_categories <- findings_with_metadata %>%
  rename(molecular_category_name = `Molecular Category`,
         molecular_abnormality_name = `Molecular Abnormality`) %>%
  select(molecular_category_name, donorID, molecular_abnormality_name, Prioritization) %>%
  group_by(molecular_category_name) %>%
  mutate(n_donors_with_vulnerability = n()) %>%
  ungroup

table(finding_categories$molecular_category_name)

```


# Prepare dx group data
```{r}

disease_group_names <- read_tsv("../input_data/disease_group_info.tsv", col_types = cols(.default = "c")) %>%
  mutate(order = 1:n())


dx_group_counts <- sample_metadata %>%
  select(donorID, dx_group) %>% 
  distinct() %>%
  left_join(disease_group_names, by=c("dx_group"="original_disease_group_name")) %>%
  group_by(dx_group) %>%
  summarize(dx_group_count = n()) %>%
  ungroup

disease_group_info <- left_join(disease_group_names,
                                dx_group_counts,
                                by=c("original_disease_group_name" = "dx_group")) %>%
  mutate(long_disease_group_name_for_plot_n = 
           paste0(long_disease_group_name_for_plot, " (n=", dx_group_count, ")"),
         long_disease_group_name_for_plot_n = factor(long_disease_group_name_for_plot_n,
                                                     levels = unique(long_disease_group_name_for_plot_n)))
         
  #        ) %>%
  # mutate(across(.fns = as_factor))

revised_colors <- c("#A8CFE4", "#468FC3", "#F3E0C4", "#CCDDAA", "#999999")

disease_group_vector <- set_names(revised_colors, disease_group_info$long_disease_group_name_for_plot_n)




```


# Barplot
```{r}

finding_categories_for_plot <- finding_categories %>%
  arrange(n_donors_with_vulnerability, molecular_category_name) %>%
  mutate(simplified_molecular_category = factor(molecular_category_name,
                                                   levels = rev(unique(molecular_category_name))) %>%
            fct_lump_min(3),
           fct_relevel("Other", after = Inf)) %>%
  left_join(sample_metadata %>%
              select(donorID, disease, dx_group) %>%
              distinct(), 
            by = "donorID") %>%
  left_join(disease_group_info, by=c("dx_group"="original_disease_group_name")) 
                                           
  
wrap_labels <- function(x) str_replace_all(x, "/", "\n") %>%
  str_replace(" ", "\n")

intial_cap_no_underscore <- function(x) {
  str_replace_all(x, "_", " ") %>%
    str_to_sentence()
}

```


```{r fig.height=5.5, fig.width=8}
p <- ggplot(finding_categories_for_plot) +
  geom_bar(aes(x = simplified_molecular_category,
               fill = long_disease_group_name_for_plot_n)) +
  theme_minimal() +
    scale_x_discrete(labels = wrap_labels) +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_text(angle = 0,
                                    vjust = 0.5),
        panel.border = element_rect(color = "black", fill = NA, size = 0.5)) +
  ylab("Number\nof\nfindings")
```

```{r}

p1 <- p +
  scale_fill_manual("Diagnostic group",
                    values = disease_group_vector) 

p1 
```


```{r}

p2 <- p1  +
  facet_wrap(~Prioritization, ncol = 1)

```


```{r fig.height=5.5, fig.width=8}

p2 

plot_output_name <- paste0("../results/number_of_findings_", format(Sys.time(), "%Y_%m_%d-%H_%M_%S_%Y"), ".png")



ggsave(plot_output_name, 
       plot = p2,
       bg = "white")

```

this output was saved as `r plot_output_name`





# Session Info
```{r}
sessionInfo()
```

