

```{r}
library(tidyverse)

```


```{r}
data_dir <- "/private/groups/treehouse/working-projects/hbeale/ckcc2_2023_07"
paths_file <- "PathsToOriginalTertiary.2022_01_18.tsv"
samples_file <- "ckcc2_metadata_from_redcap_2021-07-29_11.38.45.tsv"

paths_raw <- read_tsv(file.path(data_dir, paths_file),
                      col_types = cols(.default = "c"))
nrow(paths_raw)

project_samples <- read_tsv(file.path(data_dir, samples_file),
                      col_types = cols(.default = "c"))
nrow(project_samples)
# reminder; there are more lines in the paths_raw tibble than
# in the project_samples tibble because some samples have
# multiple tert outputs
```


```{r}
# check whether all samples are present in both data sources
# if both groups are identical,  result will be "character(0)"
setdiff(paths_raw$Sample, project_samples$THid)
setdiff(project_samples$THid, paths_raw$Sample)
# "TH34_2693_S02" "TH34_2699_S01" are in project_samples but not paths_raw
```

> nrow(project_samples)                                                                                                                                                                       
[1] 41
> setdiff(paths_raw$Sample, project_samples$THid)
character(0)
> setdiff(project_samples$THid, paths_raw$Sample)
[1] "TH34_2693_S02" "TH34_2699_S01"


```{r}

tertiary_output_locations <- paths_raw %>%
  select(sample_id = Sample, tertiary_path = path)

```

# Compendium and protocol version survey
```{r}
c_p_survey <- tertiary_output_locations %>%
  mutate(compendium = str_remove(tertiary_path, "^.*/"),
         protocol = str_replace(tertiary_path, "^.*treehouse-[a-z]*-([^/]*)/.*$", "\\1")) %>%
  select(-tertiary_path)

c_p_survey
install.packages("janitor")
library(janitor)
tabyl(c_p_survey , compendium)
tabyl(c_p_survey , protocol)
```


> tabyl(c_p_survey , compendium)
           compendium  n    percent
 compendium-v10_polya  4 0.10810811
 compendium-v11_polya  2 0.05405405
        compendium-v7  7 0.18918919
        compendium-v8 17 0.45945946
        compendium-v9  7 0.18918919
> tabyl(c_p_survey , protocol)
           protocol  n    percent
   0.17.1.0-8107fe7 28 0.75675676
 0.17.1.0-8107fe7.1  1 0.02702703
 0.17.1.0-8107fe7.2  2 0.05405405
 0.17.1.0-8107fe7.3  2 0.05405405
     14.4.0-e4460c5  4 0.10810811

# Identify samples with multiple tertiaries:
```{r}

tertiary_output_locations %>%
group_by(sample_id) %>%
  mutate(n_terts=n()) %>%
  filter(n_terts > 1) %>%
  pull(tertiary_path)

```

# Load druggable genes
```{r}

druggable_genes <- read_tsv(file.path(data_dir, "treehouseDruggableGenes_2020-03_25.txt"))

```

# Load pan cancer outlier results (irrespective of whether they are druggable)
details on file structure: https://github.com/UCSC-Treehouse/reference-file-info/blob/master/Explainer%20for%20outlier_results_files.txt
```{r}

tertiary_output_data <- tertiary_output_locations %>%
 group_by(sample_id) %>% 
  top_n(1, tertiary_path) %>% # select only one of two for the samples with multiple outputs
mutate(leads_raw = map(paste0(tertiary_path, "/outlier_results_", sample_id), read_tsv, col_types = cols(.default = "c"))) %>%
unnest(leads_raw)

top_5 <- tertiary_output_data %>%
  filter(is_top_5 == "top5") %>%
  select(sample_id, Gene)
  
pc_outliers <- tertiary_output_data %>%
  filter(pc_outlier=="pc_up") %>%
  mutate(is_druggable = Gene %in% druggable_genes$gene)
table(pc_outliers$is_druggable)


```


# Load druggable pan cancer and pan-disease outlier results
Note: at this step, we select only one of the two tertiary outputs for the samples with multiple outputs. At a later step, we analyze the outputs across the custom cohorts (ACC) to account for these samples with multiple tertiary outputs

```{r}


druggable_outliers <- tertiary_output_locations %>%
 group_by(sample_id) %>% 
  top_n(1, tertiary_path) %>% # select only one of two for the samples with multiple outputs
mutate(leads_raw = map(paste0(tertiary_path, "/automatedLeadsIdentified.tsv"), read_tsv, col_types = cols(.default = "c"))) %>%
unnest(leads_raw)

```

```{r}
# code for samples with two cohorts, currently TH34_1349_S01, TH34_1349_S02
library(jsonlite)

get_pan_disease_outliers <- function(this_tert_path){
  j4 <- fromJSON(file.path(this_tert_path, "4.0.json"))
  tibble(gene = names(j4$personalized_consensus_counts$up_outliers))
}

# confirmed with code: everything in the personalized_consensus_counts up_outliers in top 5

acc_outliers <- tertiary_output_locations %>%
  filter(!sample_id %in% c("TH34_1349_S01", "TH34_1349_S02")) %>%
  mutate(pan_disease_outliers_raw = map(tertiary_path, get_pan_disease_outliers)) %>%
  unnest(pan_disease_outliers_raw)
# acc_druggable_outliers has no rows; =0

# teh
acc_druggable_outliers <- acc_outliers %>%
            mutate(is_druggable = gene %in% druggable_genes$gene)
```

# Combine if necessary
```{r}

# if there are no pan-disease outliers, the function returns nothing and that sample has no line in the outliers tibble, so e.g. TH34_1149_S02 isn't in the pan-disease outlier list

#output_data_with_acc <- bind_rows(druggable_outliers) #, # acc_druggable_outliers)
                                           #there are no acc druggable outliers
```


# save output
```{r}

data_output_name <- paste0("druggable_outliers_from_", paths_file, ".",  format(Sys.time(), "%Y_%m_%d-%H_%M_%S_%Y"), ".tsv")

write_tsv(druggable_outliers, file.path(data_dir, data_output_name))

```

this output was saved as `r data_output_name`

# Scratch

# Identify outliers present in two custom pan-disease groups from different tertiary rounds
In this step, we analyze the outputs across the custom cohorts (ACC) to account for these samples with multiple tertiary outputs

```{r}
# code for samples with two cohorts, currently TH34_1349_S01, TH34_1349_S02
library(jsonlite)

#

get_nof1_disease_up_outliers <- function(this_tert_path){
    # this_tert_path = "/private/groups/treehouse/archive/downstream/TH34_1349_S01/tertiary/treehouse-protocol-12.0.0-9a30111.2/compendium-v8"
  j4 <- fromJSON(file.path(this_tert_path, "4.0.json"))
  j4_po <- j4["personalized_outliers"]
  j4_po$personalized_outliers$nof1_disease_outliers$up[[1]] %>% enframe()
}

# acc is for "across custom cohorts", for outliers found across multiple custom cohorts

acc_outliers <- lapply(c("TH34_1349_S01", "TH34_1349_S02"), function(this_sample){
  # this_sample="TH34_1349_S01"
  paths <- subset(tertiary_output_locations, sample_id ==this_sample)$tertiary_path
  out1 <- get_nof1_disease_up_outliers(paths[1])
  out2 <- get_nof1_disease_up_outliers(paths[2])
  return(tibble(THid=this_sample, results = intersect(out1, out2), details = "acc", file_path = paste(paths, collapse = ",")))
}) %>% bind_rows

# filter for druggable genes

druggable_genes <- read_tsv(file.path(data_dir, "treehouseDruggableGenes_2020-03_25.txt"))

acc_druggable_outliers <- acc_outliers %>%
  filter(results %in% druggable_genes$gene) %>%
  mutate(assay="druggableUpOutlier",
         sample_id = THid)

tertiary_output_data_with_acc <- bind_rows(tertiary_output_data, acc_druggable_outliers)

```

# Are any of the acc_druggable_outliers also in original tertiary_output_data?
```{r}
left_join(acc_druggable_outliers %>% select(THid, results, acc_details = details),
          tertiary_output_data %>% select(THid, results, tod_details = details))
```


# Identify outliers present in two custom pan-disease groups from different tertiary rounds
In this step, we analyze the outputs across the custom cohorts (ACC) to account for these samples with multiple tertiary outputs

```{r}
# code for samples with two cohorts, currently TH34_1349_S01, TH34_1349_S02
library(jsonlite)

get_pan_disease_outliers <- function(this_tert_path){
  j4 <- fromJSON(file.path(this_tert_path, "4.0.json"))
  tibble(gene = names(j4$personalized_consensus_counts$up_outliers))
}

# confirmed with code: everything in the personalized_consensus_counts up_outliers in top 5

acc_druggable_outliers <- tertiary_output_locations %>%
  mutate(pan_disease_outliers_raw = map(tertiary_path, get_pan_disease_outliers)) %>%
  unnest(pan_disease_outliers_raw)

# if there are no pan-disease outliers, the function returns nothing and that sample has no line in the outliers tibble, so e.g. TH34_1149_S02 isn't in the pan-disease outlier list


acc_druggable_outliers_anno <- pan_disease_outliers %>%
            mutate(is_druggable = gene %in% druggable_genes$gene)

```

```{r}

write_tsv(pan_disease_outliers_anno, paste0("pan_disease_outliers_from_", paths_file, ".tsv"))

write_tsv(pc_outliers %>%
  select(sample_id,
         tertiary_path,
         Gene,
    is_druggable
  ), paste0("pan_cancer_outliers_from_", paths_file, ".tsv"))

```

